### 2020-09-03 code/R_markdown/P1_analysis_prep.Rmd
added code and figures, did not add data or results - need to check consent re data availability

---
title: "Microbiome analysis - data prep"
author: "Maureen A. Carey"
date: "09/01/2020"
output: html_document
---

Metadata variables: Metadata variables included age-related information (month/year of birth, age in days), socio-economic status (income, perceived food deficit), time of collection (month, year, amount of rainfall that month), the child’s home conditions (presence of an indoor drain, drinking water source, water treatment method, number of children under the age of 5 years), coinfection status (positive qPCR for Giardia spp. or E. histolytica), repeat infection v. primary infection, and growth status (z-scores for weight-for-age, height-for-age, weight-for-height, and body mass index for age). These are known to influence the development of diarrhea broadly or the acquisition of Cryptosporidium (Korpe et al. 2016; Steiner et al. 2018; Mondal et al. 2012)

Code for reference:
/Users/maureencarey/local_documents/work/crypto_microbiome/data_cannot_be_shared/open_and_save_all_data.Rmd to generate /Users/maureencarey/local_documents/work/crypto_microbiome/data_cannot_be_shared/data.RData

The above code processes the private patient metadata and so cannot be shared at this point.

# Load data and packages

```{r setup, include=FALSE, echo=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/")

# load packages
library(phyloseq); library(reshape2); library(tibble)
library(rmarkdown); library(dplyr); library(tidyr)

# set constants
SEED_INT = 10; set.seed(SEED_INT); options(scipen=10000)
colors = c("#1F78B4","#FF7F00","#666666", 'light gray')
names(colors) = c('diarrheal','subclinical','no crypto','both')
# blue = "#1F78B4" orange = "#FF7F00" gray = "#666666"

# load data
load("/Users/maureencarey/local_documents/work/crypto_microbiome/data_cannot_be_shared/data.RData")

# reorganizing DOB files
DOB_df_with_study = DOB_df %>% filter(CHILDID %in% burden_mirpur$SID)
DOB_df = DOB_df %>% filter(CHILDID %in% burden_mirpur$SID) %>% select(-study)
left_study_ids = left_study_ids[left_study_ids>=2000 & left_study_ids <3406] # all files already have these kids removed

# subset for the cohort that is used for microbiome study
all_got_crypto = all_got_crypto %>% filter(CHILDID %in% burden_mirpur$SID)
all_kids = all_kids %>% filter(CHILDID %in% burden_mirpur$SID)
all_pathogens = all_pathogens %>% filter(CHILDID %in% burden_mirpur$SID)
burden_all_path = burden_all_path %>% filter(CHILDID %in% burden_mirpur$SID)
burden_crypto = burden_crypto %>% filter(CHILDID %in% burden_mirpur$SID)
burden_sex = burden_sex %>% filter(SID %in% burden_mirpur$SID)
children_without_diarrhea = children_without_diarrhea %>% filter(CHILDID %in% burden_mirpur$SID)
everything_df = everything_df %>% filter(CHILDID %in% burden_mirpur$SID)
first_infection = first_infection %>% filter(CHILDID %in% burden_mirpur$SID) # remaining kids never had a parasite
inflam_burden = inflam_burden %>% filter(CHILDID %in% burden_mirpur$SID)
total_abx_c = total_abx_c %>% filter(CHILDID %in% burden_mirpur$SID)        # remaining kids never took ABx
total_abx_c_with_diarr = total_abx_c_with_diarr %>% filter(CHILDID %in% burden_mirpur$SID)
total_diarrhea = total_diarrhea  %>% filter(CHILDID %in% burden_mirpur$SID) # remaining kids never were sick
crypto_cts = burden_all_path %>% filter(crypto == 1) %>%
  mutate(date_of_collection = as.Date(gsub('.{8}$', '', as.character(DOC)), "%m/%d/%Y")) %>%
  select(CHILDID, date_of_collection, CRYCTRT, type) %>% filter(CHILDID %in% burden_mirpur$SID)
all_cause_diar = total_diarrhea %>% group_by(CHILDID,study) %>% summarize(episodes = n()) %>% ungroup()

rm(list= ls()[! (ls() %in% c('left_study_ids','SEED_INT','colors','DOB_df',"burden_mirpur",'everything_df','total_diarrhea','burden_sex','crypto_cts'))])

```

# Generate supplemental file 1B to get read counts

```{r}

unadjusted_files_r1 = read.table("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/starting_seqs_run1.csv", 
                                 sep = ",", header = T, stringsAsFactors = F)
unadjusted_files_r2 = read.table("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/starting_seqs_run2.csv", 
                                 sep = ",", header = T, stringsAsFactors = F)
unadjusted_files_r1$X = NULL; unadjusted_files_r2$X = NULL
run1 = read.table("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/track_reads_run1.csv", 
                  sep = ",", header = T, stringsAsFactors = F); colnames(run1)[colnames(run1) == 'X'] = 'sample'
run2 = read.table("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/track_reads_run2.csv", 
                  sep = ",", header = T, stringsAsFactors = F); colnames(run2)[colnames(run2) == 'X'] = 'sample'
df1 = merge(run1, unadjusted_files_r1, by = "sample", all = T)
df2 = merge(run2, unadjusted_files_r2, by = "sample", all = T)
df = rbind(df1, df2); rm(df1, df2, run1, run2, unadjusted_files_r1, unadjusted_files_r2)

df_plot = melt(df %>% filter(!is.na(input))) %>% filter(variable %in% c("forward","input","filtered","nonchim"))
df_plot$variable = plyr::revalue(df_plot$variable, c("forward"="raw reads","input"="after adapter/\nprimer trimming", "filtered"="after quality\nfiltration", "nonchim"="after merging forward &\nreverse reads"))
df_plot$variable = factor(df_plot$variable, levels = c("raw reads","after adapter/\nprimer trimming", "after quality\nfiltration", "after merging forward &\nreverse reads"), ordered = T)
df_mean = df_plot %>% group_by(variable) %>% summarize(mean = mean(value), median = median(value))

df_summary = df
read_count_stats_df = df_mean
read_counts_df = df_plot

```

## Reading the phyloseq object

Read in the phyloseq object generated via dada2. Remove any mitochondrial or chloroplast reads.

```{r}

ps = readRDS("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/ps_2020_08_12.rds")

# Quality control - remove mitochondrial and chloroplast reads
og_taxa = data.frame(tax_table(ps))
# first remove non-bacteria
keep = rownames(og_taxa[og_taxa$Kingdom %in% c("Bacteria","Archaea"),])
ps = prune_taxa(keep,ps)
keep = rownames(og_taxa[og_taxa$Family != "Mitochondria",])
ps = prune_taxa(keep,ps)
keep = rownames(og_taxa[og_taxa$Class != "Chloroplast",])
ps = prune_taxa(keep,ps)

rm(df, keep, og_taxa)

```

Prep plot positive controls

```{r}

pos_controls1 = subset_samples(ps, startsWith(rownames(sample_data(ps)), 'pos') & endsWith(rownames(sample_data(ps)), 'run1'))
pos_controls2 = subset_samples(ps, startsWith(rownames(sample_data(ps)), 'pos') & endsWith(rownames(sample_data(ps)), 'run2'))
rare_min = 10000
pos_controls1= rarefy_even_depth(pos_controls1,sample.size=rare_min,rngseed=SEED_INT)
pos_controls2= rarefy_even_depth(pos_controls2,sample.size=rare_min,rngseed=SEED_INT)

# run 1
pos_controls = pos_controls1; top_quantity = 16
# Get the most abundant ASVs across the entire dataset
most_abundant_names = paste(1:top_quantity,
                            tax_table(pos_controls)[names(sort(taxa_sums(pos_controls),TRUE)[1:top_quantity]),"Genus"],"")
most_abundant_genus = tax_table(pos_controls)[names(sort(taxa_sums(pos_controls), TRUE)[1:top_quantity]),"Genus"]
most_abundant = names(sort(taxa_sums(pos_controls), TRUE)[1:top_quantity])
ps_for_abundplot = prune_taxa(most_abundant,pos_controls)
readsumsdf = data.frame(nreads = sort(taxa_sums(ps_for_abundplot), TRUE), sorted = 1:ntaxa(ps_for_abundplot),  type = "OTUs")
readsumsdf$names = factor(most_abundant_names,levels=most_abundant_names)
readsumsdf$genus = factor(most_abundant_genus)
rownames(readsumsdf) = NULL

plot_run1 = readsumsdf
rm(pos_controls, top_quantity,most_abundant_names, most_abundant, most_abundant_genus,ps_for_abundplot,
   readsumsdf)

# run 2
pos_controls = pos_controls2; top_quantity = 15
# Get the most abundant ASVs across the entire dataset
most_abundant_names = paste(1:top_quantity,
                            tax_table(pos_controls)[names(sort(taxa_sums(pos_controls),TRUE)[1:top_quantity]),"Genus"],"")
most_abundant_genus = tax_table(pos_controls)[names(sort(taxa_sums(pos_controls), TRUE)[1:top_quantity]),"Genus"]
most_abundant = names(sort(taxa_sums(pos_controls), TRUE)[1:top_quantity])
ps_for_abundplot = prune_taxa(most_abundant,pos_controls)
readsumsdf = data.frame(nreads = sort(taxa_sums(ps_for_abundplot), TRUE), sorted = 1:ntaxa(ps_for_abundplot),  type = "OTUs")
readsumsdf$names = factor(most_abundant_names,levels=most_abundant_names)
readsumsdf$genus = factor(most_abundant_genus)
rownames(readsumsdf) = NULL
plot_run2 = readsumsdf

rm(pos_controls, top_quantity, rare_min,most_abundant_names, most_abundant, most_abundant_genus,ps_for_abundplot,
   readsumsdf)

```

## Get only Mirpur samples.

```{r}

ps = subset_samples(ps, Site == 1) 
ps = subset_samples(ps, !startsWith(rownames(sample_data(ps)), 'pos'))

# remove if error in any reporting re Crypto infection
keep1 = ifelse(sample_data(ps)$Crypto..Any. == '1' & as.numeric(sample_data(ps)$Clinical.type) < 3, 0, 1)
ps = prune_samples(as.logical(keep1),ps)
keep2 = ifelse(sample_data(ps)$Crypto..Any. == '999' & as.numeric(sample_data(ps)$Clinical.type) > 3, 0, 1)
ps = prune_samples(as.logical(keep2),ps)
keep3 = ifelse(sample_data(ps)$Crypto..Any. == '888' & as.numeric(sample_data(ps)$Clinical.type) > 3, 0, 1)
ps = prune_samples(as.logical(keep3),ps)

rm(keep1, keep2, keep3)

```

Convert sample data to factors and convert dates to month only for easier handling.

```{r}

# convert some sample data to factors
tempsampledata = sample_data(ps)
tempsampledata$Site = NULL
tempsampledata$Clinical.type = as.factor(tempsampledata$Clinical.type)
tempsampledata$Number.of.crypto.positive.events = as.factor(tempsampledata$Number.of.crypto.positive.events)
# Clinical types
# 12=diarrhea; 10=Clinic Visit; 1 =pre-clinic visit positive; 2=pre-diarrhea case positive
tempsampledata$symptomatic = tempsampledata$Clinical.type %in% c(2,12)
tempsampledata$pre_detection = tempsampledata$Clinical.type %in% c(1,2)

# fix dates
tempsampledata$Date.of.collection = as.Date(as.character(tempsampledata$Date.of.collection),format = "%d-%B-%y")
tempsampledata$year.of.collection = as.factor(as.numeric(format(tempsampledata$Date.of.collection,'%Y')))
tempsampledata$month.of.collection = as.factor(as.numeric(format(tempsampledata$Date.of.collection,'%m')))
tempsampledata$Date.of.birth = as.Date(as.character(tempsampledata$Date.of.birth),format = "%d-%B-%y")
tempsampledata$year.of.birth = as.factor(as.numeric(format(tempsampledata$Date.of.birth,'%Y')))
tempsampledata$month.of.birth = as.factor(as.numeric(format(tempsampledata$Date.of.birth,'%m')))

# ONE PROBLEM => STILL A PROBLEM
sum(tempsampledata$Age.in.days != tempsampledata$Date.of.collection-tempsampledata$Date.of.birth)

# Add rainfall data
rainfall = read.table("~/local_documents/work/crypto_microbiome/data_cannot_be_shared/Rainfall.csv",header=TRUE, quote="",sep='\t') %>% filter(Site == 1) %>% select(-Site)
rainfall$month.of.collection = as.factor(rainfall$month)
rainfall$year.of.collection = as.factor(paste('20',rainfall$year, sep = ''))
rainfall$rainfall = rainfall$Rainfall..mm.
rainfall$Rainfall..mm. = NULL; rainfall$month = NULL; rainfall$year = NULL
tempsampledata = merge(tempsampledata,rainfall,by=c("month.of.collection","year.of.collection"), all.x = T)

# convert all "Any assays to factors"
any_colnames <- names(tempsampledata)[grep("Any",names(tempsampledata))]
tempsampledata[,any_colnames] <- lapply(tempsampledata[,any_colnames], factor)
any_colnames <- names(tempsampledata)[grep("any",names(tempsampledata))]
tempsampledata[,any_colnames] <- lapply(tempsampledata[,any_colnames], factor)

# Correct for sample name-dropping that occurs when merging
tempsampledata = sample_data(tempsampledata)
sample_names(tempsampledata) = tempsampledata$sample.id

# save
sample_data(ps) = tempsampledata
rm(rainfall,any_colnames, tempsampledata)

# remove children who were lost to follow up
ps = subset_samples(ps, !(Child.ID %in% left_study_ids))

```

## Adding genotype, pan-crypto Ct data, and species-specific Ct data

These data were obtained after the initial processing of the sequence data, so we will add them to the metadata now.

```{r}

tempsampledata = as(sample_data(ps),"data.frame")

# Add genotype information for all samples
tempsampledata = merge(tempsampledata %>% tibble::rownames_to_column('use_as_rows'), 
                       everything_df %>% select(-DOB,-study,-birth_order,-WAZ_2yr,-HAZ_2yr,-WHZ_2yr), 
                       by.x = "Child.ID", by.y = "CHILDID", all.x = T)
tempsampledata$drain = as.factor(tempsampledata$drain)
tempsampledata$food_deficit = as.factor(tempsampledata$food_deficit)
tempsampledata$toilet = as.factor(tempsampledata$toilet)
tempsampledata$drinking_water_source = as.factor(tempsampledata$drinking_water_source)

# Add crypto Cts
tempsampledata$Child.ID = as.character(tempsampledata$Child.ID)
tempsampledata = merge(tempsampledata, crypto_cts %>% select(-type), by.x = c("Child.ID","Date.of.collection"), by.y = c("CHILDID","date_of_collection"), all.x = T)
tempsampledata[tempsampledata$pre_detection == T,"CRYCTRT"] = 45 # if predetection, Crypto DNA was not assessed or detected

# get other Ct values (targeted qPCRs)
all_ct = read.table("~/local_documents/work/crypto_microbiome/data_cannot_be_shared/Cq_values_all.txt",header=TRUE, sep='\t',comment.char = "",stringsAsFactors = F, na.strings = c("#N/A"))
all_ct[is.na(all_ct)] = 45 # replace #NA and real NAs with 0
all_ct[all_ct == 0] = 45 # replace #NA and real NAs with 0
tempsampledata = tempsampledata %>% mutate(sample_id_mod = substr(sample.id,1,nchar(sample.id)-5))
tempsampledata = merge(tempsampledata, all_ct, by.x = "sample_id_mod", by.y = "sample.id", all.x = T)
rownames(tempsampledata) = tempsampledata$use_as_rows; tempsampledata$use_as_rows = NULL
sample_data(ps) = tempsampledata

df = all_ct %>% mutate(double_pos = ifelse(LIB13.C..hominis<45 & C..mel<45,1,0)) %>%
  melt(id.vars = c("Pan_crypto",'sample.id','double_pos')) %>%
  mutate(infected = ifelse(double_pos,'double positive',ifelse(Pan_crypto<45,'single',ifelse(value<45,'single','not')))) %>%
  dcast(...~variable, value = 'value') %>%
  mutate(infected_use = ifelse(LIB13.C..hominis<45 & C..mel<45,'double positive',
                               ifelse(LIB13.C..hominis<45, 'C. hominis',
                                      ifelse(C..mel<45, 'C. meleagridis','not')))) %>%
  subset(infected != "not") %>% 
  melt(id.vars = c("Pan_crypto",'sample.id','double_pos','infected','infected_use'))

df = rbind(rbind(df[df$infected == 'single' & df$variable == 'LIB13.C..hominis' & df$infected_use == 'C. hominis' & df$value > 0,],
           df[df$infected == 'single' & df$variable == 'C..mel' & df$infected_use == 'C. meleagridis' & df$value > 0,]),
           df[df$infected == 'double positive' ,])

ct_df = df

rm(df,tempsampledata,all_ct)

```

### Redo figure with Cts
```{r}

crypto_symptoms = merge(crypto_cts, DOB_df %>% select(-DOB), by = "CHILDID", all = T) %>% #all x only
  select(-date_of_collection) %>%
  mutate(type_use = ifelse(type == 'subclinical', 1, ifelse(type == 'diarrheal', 2, 0))) %>% select(-type) %>%
  mutate(type = ifelse(is.na(type_use), 0, type_use),
         site = ifelse(CHILDID %in% burden_mirpur$SID, 'Mirpur','Mirzapur')) %>% select(-type_use) %>%
  group_by(CHILDID, type, site) %>% 
  summarize(min_ct = ifelse(sum(is.na(CRYCTRT)) == length(CRYCTRT),45,
                            min(CRYCTRT,na.rm = T))) %>% ungroup() %>% 
  group_by(CHILDID, site) %>% summarize(val = mean(type), min_ct_overall = min(min_ct,na.rm = T)) %>% 
  mutate(TYPE = ifelse(val == 0, 'never infected',
                       ifelse(val == 2, 'diarrheal', 
                              ifelse(val == 1, 'subclinical',
                                     'both')))) %>% select(-val)

temp_df = merge(total_diarrhea, crypto_symptoms, by = "CHILDID", all = T)

# get kids that had no diarrhea
no_diarr = temp_df[is.na(temp_df$EPINUM),] %>% mutate(n_diar = 0) %>% select(n_diar, TYPE, CHILDID, min_ct_overall)

# summarize for kids who had diarrhea
yes_diarr = temp_df %>% filter(!is.na(EPINUM)) %>% group_by(CHILDID) %>% mutate(n_diar = n()) %>% select(n_diar, TYPE, CHILDID, min_ct_overall) %>% dplyr::slice(1) %>% ungroup()

# join 
df_plot = rbind(no_diarr, yes_diarr) 
df_plot$TYPE <- factor(df_plot$TYPE, levels = c("never infected","subclinical","diarrheal","both" ))
levels(df_plot$TYPE)[levels(df_plot$TYPE) == 'never infected'] = 'no infection\ndetected'
ct_updated_df = df_plot

rm(no_diarr, yes_diarr,temp_df,df_plot)

```

To correct for differences in sequencing depth across samples, we will subsample to a uniform depth of 10000 reads for each sample.

```{r message=FALSE}

# Rarify reads in all samples
rare_min = 10000
# remove samples with fewer counts than the minimum
ps_rare = prune_samples(sample_sums(ps) >=rare_min,ps) #8 removed
# remove samples with NA age
ps_rare = prune_samples(!is.na(sample_data(ps_rare)[,"Age.in.days"])[,1],ps_rare) # none removed
ps_rare = rarefy_even_depth(ps_rare,sample.size=rare_min,rngseed=SEED_INT) # OTUs no longer in the dataset will be removed
num_samples = length(sample_sums(ps_rare))

```

Add diversity to metadata.

```{r}

# adding Inverse Simplson to metadata 
sample_data(ps_rare)$InvSimp = as.numeric(unlist(estimate_richness(ps_rare,measures="InvSimpson")))

# adding Numebr of ASVs to metadata
richness = estimate_richness(ps_rare,measures="Observed")
colnames(richness) = c("NumASVs")

row.names(richness) = gsub("X","",row.names(richness)) # for some reason, estimate richness adds an "X" at the beginneing of row names.
asdf = as(sample_data(ps_rare),"data.frame")
asdf = merge(asdf,richness,by="row.names")
row.names(asdf) = asdf$Row.names
asdf = asdf[,!colnames(asdf) %in% c("Row.names")]
asdf$NumASVs = as.numeric(unlist(asdf$NumASVs))
sample_data(ps_rare) = asdf

rm(asdf, richness,num_samples)

```

## computation of change in each ASV between pre- and post- detection events

```{r}

# select only the pre/post detection samples, which will exclude any clinical type other than 1,2,10, or 12.
# also exclude repeat infections (only get first)
ps_rare_diff = subset_samples(ps_rare,sample_data(ps_rare)$Clinical.type %in% c(1,2,10,12))
tempsampledata = as(sample_data(ps_rare_diff),"data.frame")
new_data = tempsampledata
new_data$Clinical.type = as.numeric(as.character(new_data$Clinical.type))

# Add the sample ID for the previous surveilance sample and the age of the child at that time
new_data$previous_surv = NA; new_data$age_at_surv = NA
new_data$pre_InvSimp = NA; new_data$pre_NumASVs = NA
for (i in 1:nrow(new_data)) {
  # get TOD info
  kid = new_data$Child.ID[i]
  age_during_this_sample = new_data$Age.in.days[i] # in days
  clinical_type_of_this = new_data$Clinical.type[i]

  # make sure we're looking at the right group of pre_detection samples
  if (clinical_type_of_this == 10) { desired_clinical_type = 1
  } else if (clinical_type_of_this == 12) { desired_clinical_type = 2
  } else { desired_clinical_type = NA }
  one_kids_other_samples = new_data[new_data$Child.ID == kid,] %>% filter(Clinical.type == desired_clinical_type)

  # make sure pre_detection sample is within 30 days
  # if one option
  if ( nrow(one_kids_other_samples) == 1 ) {
    if (one_kids_other_samples$Age.in.days < age_during_this_sample & one_kids_other_samples$Age.in.days > age_during_this_sample - 42) {
      new_data$previous_surv[i] = one_kids_other_samples$sample.id
      new_data$age_at_surv[i] = one_kids_other_samples$Age.in.days
      new_data$pre_InvSimp[i] = one_kids_other_samples$InvSimp
      new_data$pre_NumASVs[i] = one_kids_other_samples$NumASVs }
  # if one option
    } else if ( nrow(one_kids_other_samples) > 1) {
      for (x in 1:nrow(one_kids_other_samples) ) {
        if (one_kids_other_samples$Age.in.days[x] < age_during_this_sample & one_kids_other_samples$Age.in.days[x] > age_during_this_sample - 42) {
          new_data$previous_surv[i] = one_kids_other_samples$sample.id[x]
          new_data$age_at_surv[i] = one_kids_other_samples$Age.in.days[x]
          new_data$pre_InvSimp[i] = one_kids_other_samples$InvSimp[x]
          new_data$pre_NumASVs[i] = one_kids_other_samples$NumASVs[x] }
        }
      } else { next }
}
rm(age_during_this_sample,one_kids_other_samples,desired_clinical_type,kid,i,x)

new_data$Clinical.type = as.factor(as.character(new_data$Clinical.type))
use_this = sample_data(new_data)
sample_names(use_this) = use_this$sample.id
sample_data(ps_rare_diff) = use_this

#omit if predetection sample doesn't have a paired detection sample
ps_rare_diff = (subset_samples(ps_rare_diff, (sample_data(ps_rare_diff)$sample.id %in% sample_data(ps_rare_diff)$previous_surv) | sample_data(ps_rare_diff)$pre_detection == F))
#omit if detection sample doesn't have a piared predetection sample
ps_rare_diff = (subset_samples(ps_rare_diff, !(is.na(sample_data(ps_rare_diff)$previous_surv)) | sample_data(ps_rare_diff)$pre_detection == T))

rm(new_data,tempsampledata,use_this, clinical_type_of_this)

for_stats = ps_rare_diff

```

# match samples and get CHANGE in microbiome
``` {r}

# get dataset where samples are matched
ps_rare_for_beta = ps_rare_diff

# calculate difference
temp_otu = otu_table(ps_rare_for_beta)
keep_these = c()
for (i in 1:nrow(sample_data(ps_rare_for_beta))) {
  time_of_detection_sample_id = rownames(sample_data(ps_rare_for_beta))[i]
  predetection_sample_id = sample_data(ps_rare_for_beta)$previous_surv[i]
  # make true if NA to remove from dataset
  if (is.na(time_of_detection_sample_id)) {sample_data(ps_rare_for_beta)$pre_detection[i] = T}
  if (is.na(predetection_sample_id)) {sample_data(ps_rare_for_beta)$pre_detection[i] = T}

  if (!sample_data(ps_rare_for_beta)$pre_detection[i]) {
    keep_these[i] = T
    temp_otu[rownames(temp_otu) == time_of_detection_sample_id,] = 
      temp_otu[rownames(temp_otu) == time_of_detection_sample_id,] - temp_otu[rownames(temp_otu) == predetection_sample_id,]
  }
  else { # we are looking at a predetection sample and aren't interested in it
    keep_these[i] = F }
}
# resave
otu_table(ps_rare_for_beta) = temp_otu

# remove predetection samples
ps_rare_for_beta = prune_samples(!sample_data(ps_rare_for_beta)$pre_detection,ps_rare_for_beta)

# get samples with fewer than 5 non-zero OTUs
nonzero_count = apply(otu_table(ps_rare_for_beta), 1, function(c)sum(c!= 0))
ps_rare_for_beta = prune_samples(nonzero_count > 5, ps_rare_for_beta)
change_in_microbiota_ps = ps_rare_for_beta
dis_df = dist(otu_table(ps_rare_for_beta))
pca_df = prcomp(dis_df)
axis_location = cbind(pca_df$rotation, sample_data(ps_rare_for_beta))

axis_location$symptomatic = as.factor(axis_location$symptomatic)
levels(axis_location$symptomatic)[levels(axis_location$symptomatic)=="FALSE"] = "subclinical"
levels(axis_location$symptomatic)[levels(axis_location$symptomatic)=="TRUE"] = "diarrheal"

```

# Save data for other scripts

```{r}

save(list = c("df_summary","read_count_stats_df", "read_counts_df","pos_controls1","pos_controls2","plot_run1", "plot_run2", "ct_df","ct_updated_df","ps_rare_diff","ps","axis_location","for_stats","ps_rare","change_in_microbiota_ps","burden_sex"), file = paste0(paste0("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P1output_plot_" , format(Sys.time(), "%Y_%m_%d")),".RData"))

save(list = c("for_stats","ps","ps_rare","ps_rare_diff","change_in_microbiota_ps"), file = paste0(paste0("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P1output_" , format(Sys.time(), "%Y_%m_%d")),".RData"))

```

# Make figures

```{r}

#rmarkdown::render("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/code/R_markdown/P1_analysis_prep_figures.Rmd")

```

### 2020-09-03 code/R_markdown/P1_analysis_prep_figures.Rmd
added code and figures, did not add data or results - need to check consent re data availability

---
title: "Microbiome analysis - data prep"
author: "Maureen A. Carey"
date: "09/01/2020"
output: html_document
---

Metadata variables: Metadata variables included age-related information (month/year of birth, age in days), socio-economic status (income, perceived food deficit), time of collection (month, year, amount of rainfall that month), the child’s home conditions (presence of an indoor drain, drinking water source, water treatment method, number of children under the age of 5 years), coinfection status (positive qPCR for Giardia spp. or E. histolytica), repeat infection v. primary infection, and growth status (z-scores for weight-for-age, height-for-age, weight-for-height, and body mass index for age). These are known to influence the development of diarrhea broadly or the acquisition of Cryptosporidium (Korpe et al. 2016; Steiner et al. 2018; Mondal et al. 2012)

Code for reference:
/Users/maureencarey/local_documents/work/crypto_microbiome/data_cannot_be_shared/open_and_save_all_data.Rmd to generate /Users/maureencarey/local_documents/work/crypto_microbiome/data_cannot_be_shared/data.RData

The above code processes the private patient metadata and so cannot be shared at this point.

# Load data and packages

```{r setup, include=FALSE, echo=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/")

# load packages
library(ggplot2); library(tibble); library(dplyr); library(phyloseq)

# set constants
SEED_INT = 10; set.seed(SEED_INT); options(scipen=10000); theme_set(theme_bw())
colors = c("#1F78B4","#FF7F00","#666666", 'light gray')
names(colors) = c('diarrheal','subclinical','no crypto','both')
# blue = "#1F78B4" orange = "#FF7F00" gray = "#666666"

# load data
load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P1output_plot_2020_09_02.RData")

```

# Generate supplemental file 1B to get read counts

```{r}

p = ggplot(read_counts_df) + geom_histogram(aes(x = value)) + facet_wrap(~variable, nrow = 1) + ylab('samples') + xlab('reads') + 
  geom_vline(data = read_count_stats_df, aes(xintercept = mean), color = "red", linetype = "dashed") +
  theme(axis.text.x = element_text(size = 6))
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/SFig1D_read_counts.png", 
       p, width = 8, height = 2, dpi = 600)

```

# Sample numbers

Total (raw) reads: `r sum(df_summary$forward)`

Mean (raw) reads per sample: `r mean(df_summary$forward)`

Median (raw) reads per sample: `r median(df_summary$forward)`

Total (filtered) reads per sample: `r sum(na.omit(df_summary$nonchim))`

Mean (filtered) reads per sample: `r mean(na.omit(df_summary$nonchim))`

Median (filtered) reads per sample: `r median(na.omit(df_summary$nonchim))`

Total samples (excluding controls): `r length(df_summary$sample[!startsWith(df_summary$sample,'neg') & !startsWith(df_summary$sample,'pos') ])`

# Plot positive controls

```{r}

# run 1
rare_min = 10000
readsumsdf = plot_run1
p1 = ggplot(readsumsdf, 
           aes(x = names, y = nreads/(rare_min*nrow(sample_data(pos_controls1))), fill=genus)) + geom_bar(stat="identity") + theme(axis.text.x = element_text(size=14,angle=45,hjust=1),
              axis.text.y = element_text(size=14),
              axis.title.x = element_text(size=20),
              axis.title.y = element_text(size=20),
              plot.margin=unit(c(.5,.5,.5,.5),"cm")) + 
  labs(y="Fraction of reads\nfrom ASV in run1\npositive controls", x="ASV assigned genus") +
  scale_x_discrete(labels = readsumsdf$genus) + 
  scale_fill_grey() + guides(fill = F) + scale_y_continuous(labels = scales::percent)
p1
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/SFig1A_abundant_asvs_run1_pos_controls.png", 
plot=p1, width = 5.6,height = 4.2,units = "in",dpi = 300)

# run 2
readsumsdf = plot_run2
p1 = ggplot(readsumsdf, 
           aes(x = names, y = nreads/(rare_min*nrow(sample_data(pos_controls2))), fill=genus)) + geom_bar(stat="identity") + theme(axis.text.x = element_text(size=14,angle=45,hjust=1),
              axis.text.y = element_text(size=14),
              axis.title.x = element_text(size=20),
              axis.title.y = element_text(size=20),
              plot.margin=unit(c(.5,.5,.5,.5),"cm")) + 
  labs(y="Fraction of reads\nfrom ASV in run2\npositive controls", x="ASV assigned genus") +
  scale_x_discrete(labels = readsumsdf$genus) + 
  scale_fill_grey() + guides(fill = F) + scale_y_continuous(labels = scales::percent)
p1
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/SFig1B_abundant_asvs_run2_pos_controls.png", plot=p1, width = 5.5,height = 4.2,units = "in",dpi = 300)

```

# Sample numbers (prior to filtration due to correct pairing)

Number of children: `r length(unique(sample_data(ps)$Child.ID))`

Number of samples: `r dim(sample_data(ps))[[2]]`

Total reads: `r sum(sample_sums(ps))`

Mean reads per sample: `r mean(sample_sums(ps))`

Median reads per sample: `r median(sample_sums(ps))`

## Plot Ct data

```{r}

df = ct_df

df %>%
  ggplot() + 
  geom_point(aes(x = Pan_crypto, y = value, color = variable),alpha = 0.5) + xlab('Ct for pan-genus qPCR')+ 
  facet_wrap(~infected_use) + ylab('Ct, species-specific qPCR') +guides(color = F)

```

### Redo figure with Cts
```{r}

df_plot = ct_updated_df

colors_temp = colors; colors_temp['no infection\ndetected'] = "#666666"
p = ggplot(df_plot %>% dplyr::rename(`clinical type` = TYPE), aes(x = min_ct_overall, y = n_diar,color = `clinical type`)) +
     geom_point(alpha = .2) + stat_smooth(method = "glm",se = F)+scale_color_manual(values = colors_temp) +
  xlab('max parasite burden\n(Ct)') + ylab('number of diarrheal episodes')
p

```

# Sample numbers (AFTER filtration due to correct pairing)

Number of children: `r length(unique(sample_data(ps_rare_diff)$Child.ID))`

Number of samples: `r length(sample_data(ps_rare_diff)$Child.ID)`

Total number of taxa: `r ntaxa(ps_rare_diff)`

# Figure 2

``` {r}

ggplot(axis_location, aes(x = PC1, y = PC2, color = symptomatic, shape = Number.of.crypto.positive.events)) + 
  geom_point() + 
  scale_color_manual(values = c(colors["diarrheal"],colors["subclinical"]))+ # 12 is diarrhea, 1,2 are pre
  theme(strip.text.x=element_text(size=20,face="bold"), 
        plot.margin=unit(c(.5,.5,.5,.5),"cm")) + 
  labs(color = "symptomatic?", shape = "Number of infections")
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/SFig4_pcoa_change_mirpur.png",  width = 5.4,height = 3,units = "in",dpi = 600)

```

# Table 1
```{r}

ps_raw = readRDS("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/ps_2020_08_12.rds")
table(sample_data(subset_samples(ps_raw, Site == 1))$Clinical.type)
table(sample_data(subset_samples(ps_raw, Site == 2))$Clinical.type)
length(sample_data(subset_samples(ps_raw, Site == 1))$Clinical.type)
length(sample_data(subset_samples(ps_raw, Site == 2))$Clinical.type)

ggplot(as(sample_data(ps_raw),"data.frame"), aes(x = Age.in.days)) +
  geom_histogram(binwidth = 30)

length(unique(sample_data(ps_rare_diff)$Child.ID))
length((sample_data(ps_rare_diff)$Child.ID))
temp  = as(sample_data(subset_samples(ps_rare_diff, pre_detection == F )),"data.frame") %>% select(Child.ID, Number.of.crypto.positive.events) %>% tibble::rownames_to_column('rown') %>% select(-rown) %>% group_by(Child.ID) %>% top_n(n=1)
table(temp$Number.of.crypto.positive.events)

temp  = as(sample_data(subset_samples(ps_rare_diff, pre_detection == F & symptomatic == F )),"data.frame") %>% select(Child.ID, Number.of.crypto.positive.events) 
table(temp$Number.of.crypto.positive.events)
temp  = as(sample_data(subset_samples(ps_rare_diff, pre_detection == F & symptomatic == T )),"data.frame") %>% select(Child.ID, Number.of.crypto.positive.events) 
table(temp$Number.of.crypto.positive.events)
temp  = as(sample_data(subset_samples(ps_rare_diff, pre_detection == F)),"data.frame") %>% select(Child.ID, Number.of.crypto.positive.events) 
table(temp$Number.of.crypto.positive.events)

sample_data(ps_rare_diff)$Date.of.collection %>% min()
sample_data(ps_rare_diff)$Date.of.collection %>% max()

table(merge(as(sample_data(subset_samples(ps_rare_diff, pre_detection == F )),"data.frame")%>% select(Child.ID) %>% distinct(), 
            burden_sex, by.x = 'Child.ID', by.y = "SID", all.x = T)$SEX)
#1 = male, 2 = female

mean(sample_data(subset_samples(ps_rare_diff, symptomatic == T & pre_detection == F))$Age.in.days)
sd(sample_data(subset_samples(ps_rare_diff, symptomatic == T & pre_detection == F))$Age.in.days)
mean(sample_data(subset_samples(ps_rare_diff, symptomatic == F & pre_detection == F))$Age.in.days)
sd(sample_data(subset_samples(ps_rare_diff, symptomatic == F & pre_detection == F))$Age.in.days)
mean(sample_data(subset_samples(ps_rare_diff, pre_detection == F))$Age.in.days)
sd(sample_data(subset_samples(ps_rare_diff, pre_detection == F))$Age.in.days)
t.test(sample_data(subset_samples(ps_rare_diff, symptomatic == T & pre_detection == F))$Age.in.days, sample_data(subset_samples(ps_rare_diff, symptomatic == F & pre_detection == F))$Age.in.days)

mean(sample_data(subset_samples(ps_rare_diff, symptomatic == T & pre_detection == F))$Pan_crypto)
sd(sample_data(subset_samples(ps_rare_diff, symptomatic == T & pre_detection == F))$Pan_crypto)
mean(sample_data(subset_samples(ps_rare_diff, symptomatic == F & pre_detection == F))$Pan_crypto)
sd(sample_data(subset_samples(ps_rare_diff, symptomatic == F & pre_detection == F))$Pan_crypto)
mean(sample_data(subset_samples(ps_rare_diff, pre_detection == F))$Pan_crypto)
sd(sample_data(subset_samples(ps_rare_diff, pre_detection == F))$Pan_crypto)
t.test(sample_data(subset_samples(ps_rare_diff, symptomatic == T & pre_detection == F))$Pan_crypto, sample_data(subset_samples(ps_rare_diff, symptomatic == F & pre_detection == F))$Pan_crypto)

nrow(sample_data(subset_samples(ps_rare_diff, symptomatic == T & pre_detection == F)))
nrow(sample_data(subset_samples(ps_rare_diff, symptomatic == F & pre_detection == F)))
nrow(sample_data(subset_samples(ps_rare_diff, symptomatic == T & pre_detection == T)))
nrow(sample_data(subset_samples(ps_rare_diff, symptomatic == F & pre_detection == T)))
nrow(sample_data((ps_rare_diff)))

table(sample_data(subset_samples(ps_rare_diff, pre_detection == F))$Number.of.crypto.positive.events)
table(sample_data(subset_samples(ps_rare_diff, pre_detection == F & symptomatic == F))$Number.of.crypto.positive.events)
table(sample_data(subset_samples(ps_rare_diff, pre_detection == F & symptomatic == T))$Number.of.crypto.positive.events)

t1 = as(sample_data(subset_samples(ps_rare_diff, pre_detection == F & symptomatic == T)), "data.frame") %>% 
  select(Child.ID,Age.in.days,age_at_surv) %>% 
  mutate(age_diff = Age.in.days- age_at_surv) %>% 
  select(age_diff) 
mean(t1$age_diff)
sd(t1$age_diff)
t2 = as(sample_data(subset_samples(ps_rare_diff, pre_detection == F & symptomatic == F)), "data.frame") %>% 
  select(Child.ID,Age.in.days,age_at_surv) %>% 
  mutate(age_diff = Age.in.days- age_at_surv) %>% 
  select(age_diff)
mean(t2$age_diff)
sd(t2$age_diff)
t.test(t1$age_diff,t2$age_diff)

t3 = as(sample_data(subset_samples(ps_rare_diff, pre_detection == F)), "data.frame") %>% 
  select(Child.ID,Age.in.days,age_at_surv) %>% 
  mutate(age_diff = Age.in.days- age_at_surv) %>% 
  select(age_diff)
mean(t3$age_diff)
sd(t3$age_diff)
min(t3$age_diff)
max(t3$age_diff)

```

### 2020-09-03 code/R_markdown/P2_cohort.Rmd
added code and figures, did not add data or results - need to check consent re data availability

---
title: "Microbiome analysis - cohort figures"
author: "Maureen A. Carey"
date: "09/01/2020"
output: html_document
---

# Load packages and data

```{r setup, include=FALSE, echo=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/")

# load packages
library(phyloseq); library(reshape2); library(rmarkdown)
library(dplyr); library(tidyr); library(vegan)

# set constants
SEED_INT = 10; set.seed(SEED_INT); options(scipen=10000)
colors = c("#1F78B4","#FF7F00","#666666", 'light gray')
names(colors) = c('diarrheal','subclinical','no crypto','both')
# blue = "#1F78B4" orange = "#FF7F00" gray = "#666666"

# load data
load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P1output_2020_09_02.RData")
load("/Users/maureencarey/local_documents/work/crypto_microbiome/data_cannot_be_shared/data.RData")

# reorganizing DOB files
DOB_df_with_study = DOB_df %>% filter(CHILDID %in% burden_mirpur$SID)
DOB_df = DOB_df %>% filter(CHILDID %in% burden_mirpur$SID) %>% select(-study)
left_study_ids = left_study_ids[left_study_ids>=2000 & left_study_ids <3406] # all files already have these kids removed
rm(inflam_nih,inflam_provide,nih_antibiotics_key,nih_crypto,provide_antibiotics_key,provide_crypto)

# subset for the cohort that is used for microbiome study
all_got_crypto = all_got_crypto %>% filter(CHILDID %in% burden_mirpur$SID)
all_kids = all_kids %>% filter(CHILDID %in% burden_mirpur$SID)
all_pathogens = all_pathogens %>% filter(CHILDID %in% burden_mirpur$SID)
burden_all_path = burden_all_path %>% filter(CHILDID %in% burden_mirpur$SID)
burden_crypto = burden_crypto %>% filter(CHILDID %in% burden_mirpur$SID)
burden_sex = burden_sex %>% filter(SID %in% burden_mirpur$SID)
children_without_diarrhea = children_without_diarrhea %>% filter(CHILDID %in% burden_mirpur$SID)
everything_df = everything_df %>% filter(CHILDID %in% burden_mirpur$SID)
first_infection = first_infection %>% filter(CHILDID %in% burden_mirpur$SID) # remaining kids never had a parasite
inflam_burden = inflam_burden %>% filter(CHILDID %in% burden_mirpur$SID)
total_abx_c = total_abx_c %>% filter(CHILDID %in% burden_mirpur$SID)        # remaining kids never took ABx
total_abx_c_with_diarr = total_abx_c_with_diarr %>% filter(CHILDID %in% burden_mirpur$SID)
total_diarrhea = total_diarrhea  %>% filter(CHILDID %in% burden_mirpur$SID) # remaining kids never were sick
crypto_cts = burden_all_path %>% filter(crypto == 1) %>%
  mutate(date_of_collection = as.Date(gsub('.{8}$', '', as.character(DOC)), "%m/%d/%Y")) %>%
  select(CHILDID, date_of_collection, CRYCTRT, type) %>% filter(CHILDID %in% burden_mirpur$SID)
all_cause_diar = total_diarrhea %>% group_by(CHILDID,study) %>% summarize(episodes = n()) %>% ungroup()
rm(burden_all_path)

rm(list= ls()[! (ls() %in% c("ps","ps_rare","ps_rare_diff",'left_study_ids','SEED_INT','colors','DOB_df','burden_crypto',
                             "burden_mirpur",'total_abx_c_with_diarr','total_diarrhea'))])

```

## Figure 2 - Emphasizing amount of diarrhea and antibiotic usage

Figure 2A
``` {r}

df_plot = merge(DOB_df, total_diarrhea %>% group_by(CHILDID) %>% summarize(n_diarr = n()), 
  by = "CHILDID", all = T)
df_plot$n_diarr[is.na(df_plot$n_diarr)] = 0
df_plot = merge(df_plot,as(sample_data(subset_samples(ps, pre_detection == F)),'data.frame'),
                by.x = "CHILDID", by.y = "Child.ID", all = T) %>% 
  select(CHILDID, DOB, n_diarr,symptomatic)

# note: slice must be dplyr::slice with these packages
df_plot = df_plot  %>% filter(CHILDID %in% burden_mirpur$SID)  %>%
  group_by(CHILDID) %>% mutate(occurances = n()) %>% ungroup() %>%
  mutate(sympt_use = ifelse(is.na(symptomatic),0,ifelse(symptomatic,2,1))) %>%
  group_by(CHILDID) %>% mutate(val = mean(sympt_use)) %>%
  mutate(TYPE = ifelse(val == 0, 'not in microbiome cohort',
                       ifelse(val == 2, 'diarrheal',
                              ifelse(val == 1, 'subclinical',
                                     'both')))) %>%
  dplyr::slice(1) %>% ungroup() %>%
  select(CHILDID, n_diarr, TYPE)
df_plot$TYPE <- factor(df_plot$TYPE, levels = c("not in microbiome cohort","subclinical","both","diarrheal"))
df_fig2A = df_plot

```

Supplemental Figure 2A
```{r}

n_kids = length(unique(total_abx_c_with_diarr$CHILDID))
to_plot = total_abx_c_with_diarr %>% mutate(age = date-DOB) %>% filter(CHILDID %in% burden_mirpur$SID) 
df_sf2A = to_plot

```

Figure 2B
```{r}

crypto_symptoms = merge(burden_crypto %>% filter(CHILDID %in% burden_mirpur$SID), DOB_df %>% select(-DOB), by = "CHILDID", all = T) %>% 
  select(-DOB, -DOC,-study,-CRYPTO_use) %>%
  mutate(type_use = ifelse(TYPE == 'subclinical', 1, ifelse(TYPE == 'diarrheal', 2, 0))) %>% select(-TYPE) %>%
  mutate(type = ifelse(is.na(type_use), 0, type_use)) %>%
  group_by(CHILDID) %>% summarize(val = mean(type)) %>%
  mutate(TYPE = ifelse(val == 0, 'never infected',
                       ifelse(val == 2, 'diarrheal', 
                              ifelse(val == 1, 'subclinical',
                                     'both')))) %>%
  select(-val)
temp_df = merge(total_diarrhea, crypto_symptoms, by = "CHILDID", all = T)

# get kids that had no diarrhea
no_diarr = temp_df[is.na(temp_df$EPINUM),] %>% mutate(n = 0) %>% select(n, TYPE, CHILDID)

# summarize for kids who had diarrhea
yes_diarr = temp_df %>% filter(!is.na(EPINUM)) %>% group_by(CHILDID) %>% mutate( n = n()) %>% select(n, TYPE, CHILDID) %>% dplyr::slice(1) %>% ungroup()

# join 
df_plot = rbind(no_diarr, yes_diarr)
rm(no_diarr, yes_diarr)
df_plot$TYPE[df_plot$TYPE == "never infected"] = "no infection\ndetected"
df_plot$TYPE <- factor(df_plot$TYPE, levels = c("no infection\ndetected","subclinical","diarrheal","both" ))

df_fig2B = df_plot

```

Figure 2C
```{r}

crypto_symptoms = merge(burden_crypto %>% filter(CHILDID %in% burden_mirpur$SID), DOB_df %>% select(-DOB), by = "CHILDID", all = T) %>% 
  select(-DOB, -DOC,-study,-CRYPTO_use) %>%
  mutate(type_use = ifelse(TYPE == 'subclinical', 1, ifelse(TYPE == 'diarrheal', 2, 0))) %>% select(-TYPE) %>%
  mutate(type = ifelse(is.na(type_use), 0, type_use)) %>%
  group_by(CHILDID) %>% summarize(val = mean(type)) %>%
  mutate(TYPE = ifelse(val == 0, 'no infection\ndetected',
                       ifelse(val == 2, 'diarrheal', 
                              ifelse(val == 1, 'subclinical',
                                     'both')))) %>%
  select(-val)
temp_df = merge(total_abx_c_with_diarr %>% select(CHILDID, value_use, study, date), 
                crypto_symptoms, by = "CHILDID", all = T)

# get kids that had no abx
no_diarr = temp_df[is.na(temp_df$value_use),] %>% mutate(n = 0) %>% select(n, TYPE, CHILDID)

# summarize for kids who had abx
yes_diarr = temp_df %>% filter(!is.na(value_use)) %>% group_by(CHILDID) %>% mutate( n = n()) %>% select(n, TYPE, CHILDID) %>% dplyr::slice(1) %>% ungroup()

# join 
df_plot = rbind(no_diarr, yes_diarr)
rm(no_diarr, yes_diarr)
df_plot$TYPE <- factor(df_plot$TYPE, levels = c("no infection\ndetected","subclinical","diarrheal","both" ))

df_fig2C = df_plot

```

## Figure 3

Frequent reads overall - Figure 3A + 3D
```{r}

# subset the phyloseq object to only include clinical types 1,2,10, and 12 since analyses will focus on these
ps_for_explore = subset_samples(ps_rare, Clinical.type %in% c(1,2,10,12))
top_quantity = 10
rare_min = 10000

# Get the most abundant ASVs across the entire dataset
most_abundant_names = paste(1:top_quantity,tax_table(ps_for_explore)[names(sort(taxa_sums(ps_for_explore), TRUE)[1:top_quantity]),"Genus"],"")
most_abundant_genus = tax_table(ps_for_explore)[names(sort(taxa_sums(ps_for_explore), TRUE)[1:top_quantity]),"Genus"]
most_abundant = names(sort(taxa_sums(ps_for_explore), TRUE)[1:top_quantity])
ps_for_abundplot = prune_taxa(most_abundant,ps_for_explore)
readsumsdf = data.frame(nreads = sort(taxa_sums(ps_for_abundplot), TRUE), sorted = 1:ntaxa(ps_for_abundplot), 
                        type = "OTUs")
readsumsdf$names = factor(most_abundant_names,levels=most_abundant_names)
readsumsdf$genus = factor(most_abundant_genus)
rownames(readsumsdf) = NULL
df_fig3A = readsumsdf
ps_for_explore_fig3A = ps_for_explore

## to show lack of dominance of individual microbes
samples_df = as(sample_data(ps_for_explore),"data.frame") %>% select(symptomatic, pre_detection)
temp_df_tax = tax_table(ps_for_explore) %>% melt()
n_otus = otu_table(ps_for_explore) %>% melt() %>% 
  group_by(Var1) %>% summarize(n_reads = sum(value))
temp_df_otu = otu_table(ps_for_explore) %>% melt() %>% 
  group_by(Var1) %>% # group by sample
  arrange(-value) %>% dplyr::slice(1) %>% ungroup() %>% mutate(percent_reads = value/10000) %>%
  select(-value) %>%
  dplyr::rename(sample = Var1, ASV = Var2)
plot_df = merge(temp_df_otu, temp_df_tax %>% filter(Var2 == 'Genus'), all.x = T, by.x = 'ASV',by.y = "Var1") %>% select(-Var2)
colnames(plot_df)[colnames(plot_df) == 'value'] = 'Genus'
replacement = temp_df_tax %>%  filter(Var1 == plot_df[is.na(plot_df$Genus),1], Var2 == 'Order') %>% select(value)
plot_df$Genus = as.character(plot_df$Genus)
plot_df[is.na(plot_df$Genus),4] = paste0('[o]',replacement$value[[1]] %>% as.character())
plot_df$Genus = as.factor(plot_df$Genus)
rownames(plot_df) = plot_df$Var1
alt_df = cbind(plot_df, samples_df)

plot_df_stats = alt_df %>% mutate(grps = paste0(symptomatic,pre_detection))
res.aov <- aov(percent_reads ~ symptomatic +pre_detection, data = alt_df)
summary(res.aov)

## to demonstrate issue with sample-to-sample diversity, plot the number of samples containing each ASV
taxa_counts = sort(colSums(otu_table(ps_for_explore) > 1),TRUE)[1:top_quantity] # number of samples each ASV is in, taking top 25
highest_sample_count_names = paste(1:top_quantity,tax_table(ps_for_explore)[names(sort(colSums(otu_table(ps_for_explore) > 1),TRUE)[1:top_quantity]),"Genus"],"")
highest_sample_count_genus = tax_table(ps_for_explore)[names(sort(colSums(otu_table(ps_for_explore) > 1),TRUE)[1:top_quantity]),"Genus"]
highest_sample_count = names(sort(colSums(otu_table(ps_for_explore) > 1),TRUE)[1:top_quantity])
ps_for_countplot = prune_taxa(highest_sample_count,ps_for_explore)
countsumsdf = data.frame(nsamples = sort(colSums(otu_table(ps_for_explore) > 1),TRUE)[1:top_quantity], sorted = 1:ntaxa(ps_for_countplot), 
                        type = "OTUs")
countsumsdf$names = factor(highest_sample_count_names,levels=highest_sample_count_names)
countsumsdf$genus = factor(highest_sample_count_genus)
df_fig3D = countsumsdf

## how does this change if we merge by genus?
ps_rare_bygenus = tax_glom(ps_for_explore,taxrank="Genus")
taxa_counts = sort(colSums(otu_table(ps_rare_bygenus) > 1),TRUE)[1:top_quantity] # number of samples each ASV is in, taking top 25
highest_sample_count_names = paste(1:top_quantity,tax_table(ps_rare_bygenus)[names(sort(colSums(otu_table(ps_rare_bygenus) > 1),TRUE)[1:top_quantity]),"Genus"],"")
highest_sample_count_genus = tax_table(ps_rare_bygenus)[names(sort(colSums(otu_table(ps_rare_bygenus) > 1),TRUE)[1:top_quantity]),"Genus"]
highest_sample_count = names(sort(colSums(otu_table(ps_rare_bygenus) > 1),TRUE)[1:top_quantity])
ps_for_countplot = prune_taxa(highest_sample_count,ps_rare_bygenus)
countsumsdf = data.frame(nsamples = sort(colSums(otu_table(ps_rare_bygenus) > 1),TRUE)[1:top_quantity], sorted = 1:ntaxa(ps_for_countplot), type = "OTUs")
countsumsdf$names = factor(highest_sample_count_names,levels=highest_sample_count_names)
countsumsdf$genus = factor(highest_sample_count_genus)
alt_countsumsdf = countsumsdf
ps_rare_bygenus_alt_fig = ps_rare_bygenus

```

Richness and Evenness - Figure 2B and 2C

```{r}

# First, demonstrate the difference in alpha diversity in clinical vs. sublinical children
ps_for_alpha = ps_rare

# only include codes 1, 10, (pre and post diarrheal) 2, and 12 (pre and post non-diarrheal)
ps_for_alpha = prune_samples(sample_data(ps_for_alpha)$Clinical.type %in% c(1,10,2,12),ps_for_alpha)
#infection_labeller = list("FALSE"="Uninfected","TRUE"="Infected")
# change infected label from True/False to infected/uninfected
tempsampledata = sample_data(ps_for_alpha)
tempsampledata$Clinical.type = as.factor(tempsampledata$Clinical.type)
sample_data(ps_for_alpha) = tempsampledata

# prep for stats
richness = estimate_richness(ps_for_alpha,measures="Observed")
row.names(richness) = gsub("X","",row.names(richness)) # for some reason, estimate richness adds an "X" at the beginneing of row names.
asdf = as(sample_data(ps_for_alpha),"data.frame")
asdf = merge(asdf,richness,by="row.names")
clinical_anova = aov(Observed~Clinical.type,data=asdf)
clinical_tukey_posthoc_richness = TukeyHSD(clinical_anova)

evenness = estimate_richness(ps_for_alpha,measures="InvSimpson")
asdf = as(sample_data(ps_for_alpha),"data.frame")
asdf = merge(asdf,evenness,by="row.names")
clinical_anova = aov(InvSimpson~Clinical.type,data=asdf)
clinical_tukey_posthoc_evenness = TukeyHSD(clinical_anova)

# save the anova results
saveRDS(clinical_tukey_posthoc_evenness, "/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/InvSimpson_anova_posthoc_Mirpur.rds")
saveRDS(clinical_tukey_posthoc_richness, "/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/Observed_anova_posthoc_Mirpur.rds")

```

## Figure 4A

```{r}

# Samples are too distinct from each other NMDS to converge, so let's use PCoA
ps_rare_for_beta = ps_rare
ps_rare_for_beta = prune_samples(sample_data(ps_rare_for_beta)$Clinical.type %in% c(1,2,10,12),ps_rare_for_beta)

# get samples with fewer than 5 non-zero OTUs
nonzero_count = apply(otu_table(ps_rare_for_beta), 1, function(c)sum(c!= 0))
ps_rare_for_beta = prune_samples(nonzero_count > 5, ps_rare_for_beta)
ps_rare_for_beta = transform_sample_counts(ps_rare_for_beta, function(x) x / sum(x) )

levels(sample_data(ps_rare_for_beta)$Clinical.type) = c(1,1,10,12)
# ordination
ps_rare.ord = phyloseq::ordinate(ps_rare_for_beta,  method="PCoA", distance = "euclidean")#,distance = "bray", trymax=100,k=2)
d = ps_rare.ord$vectors
grps = as(sample_data(ps_rare_for_beta), "data.frame") %>% select(symptomatic, pre_detection) %>% 
  mutate(group_use = paste(symptomatic,pre_detection, sep = '_'))
adon.results<-adonis(d ~ grps$symptomatic + grps$pre_detection, method="euclidean",perm=999)
print(adon.results)

```

## Figure 3, focusing on Abx

```{r}

# prep data for if a kid had no abx
kids_IDs = as(sample_data(ps_rare),"data.frame") %>% 
  select(Child.ID, Date.of.collection,symptomatic,pre_detection) %>% 
  filter(pre_detection == F) %>% select(-pre_detection)
kids_IDs$Child.ID = as.integer(as.character(kids_IDs$Child.ID))
kids_IDs_all = as(sample_data(ps_rare),"data.frame") %>% 
  select(Child.ID, Date.of.collection,symptomatic,pre_detection)
kids_IDs_all$Child.ID = as.integer(as.character(kids_IDs_all$Child.ID))

# count abx events
tempsampledata = as(sample_data(ps_rare),"data.frame") %>% 
                  select(Child.ID, sample.id, Number.of.crypto.positive.events,Date.of.birth, 
                         Date.of.collection, symptomatic, pre_detection)
temp_df1 = merge(tempsampledata, (total_abx_c_with_diarr %>% select(CHILDID, value, date)), 
                all.x = T, by.x = "Child.ID", by.y = "CHILDID")
temp_df1 = temp_df1 %>% 
  dplyr::rename(Date.of.drug = date, drug = value)

temp_df = temp_df1 %>% mutate(drug_used_X_days_before_crypto = as.numeric(Date.of.collection - Date.of.drug)) %>%
  filter(drug_used_X_days_before_crypto >= 0) %>% # only antibiotics used before crypto event
  select(-Date.of.drug) %>%
  group_by(Child.ID, sample.id, Number.of.crypto.positive.events,Date.of.birth, 
                         Date.of.collection, symptomatic, pre_detection,drug) %>%
  mutate(n_uses = n(), min_drug_used_X_days_before_crypto = min(drug_used_X_days_before_crypto)) %>% ungroup() %>%
  select(-drug_used_X_days_before_crypto) %>% distinct()

temp_df_month = temp_df1 %>% mutate(drug_used_X_days_before_crypto = as.numeric(Date.of.collection - Date.of.drug)) %>%
  filter(drug_used_X_days_before_crypto >= 0 & drug_used_X_days_before_crypto < 32) %>% # only antibiotics w/i 1 mo & used before crypto event
  select(-Date.of.drug) %>%
  group_by(Child.ID, sample.id, Number.of.crypto.positive.events,Date.of.birth, 
                         Date.of.collection, symptomatic, pre_detection,drug) %>%
  mutate(n_uses = n(), min_drug_used_X_days_before_crypto = min(drug_used_X_days_before_crypto)) %>% ungroup() %>%
  select(-drug_used_X_days_before_crypto) %>% distinct()
rm(temp_df1)
num_abx_epi = temp_df %>% group_by(Child.ID, sample.id, Number.of.crypto.positive.events,Date.of.birth, 
                         Date.of.collection, symptomatic, pre_detection) %>% mutate(n_abx_doses = sum(n_uses)) %>%
  ungroup() %>% select(-drug,-n_uses, -min_drug_used_X_days_before_crypto) %>% distinct()

num_abx_epi_month = temp_df_month %>% 
  group_by(Child.ID, sample.id, Number.of.crypto.positive.events,Date.of.birth, 
                         Date.of.collection, symptomatic, pre_detection) %>% mutate(n_abx_doses = sum(n_uses)) %>%
  ungroup() %>% select(-drug,-n_uses, -min_drug_used_X_days_before_crypto) %>% distinct()

```

Do all kids have at least one abx between predetction and time of detection???

```{r}

temp_df2 = merge(num_abx_epi, kids_IDs_all, by = c("Child.ID","Date.of.collection","symptomatic","pre_detection"), all = T)
temp_df2$n_abx_doses[is.na(temp_df2$n_abx_doses)] = 0
temp_df2$Number.of.crypto.positive.events = as.numeric(as.character(temp_df2$Number.of.crypto.positive.events))
temp_df2 = temp_df2 %>% group_by(Child.ID, Number.of.crypto.positive.events, Date.of.birth) %>% 
  mutate(abx_between_samples = ifelse(min(n_abx_doses) != max(n_abx_doses), 1, 0)) %>% 
  ungroup() %>% filter(pre_detection == T)
#stats
tbl = table(temp_df2 %>% select(symptomatic,abx_between_samples))
chisq.test(tbl)
melted_tbl = melt(tbl)
melted_tbl$symptomatic = as.factor(melted_tbl$symptomatic)
levels(melted_tbl$symptomatic) <- list(subclinical="FALSE", diarrheal="TRUE")
melted_tbl$abx_between_samples = as.factor(melted_tbl$abx_between_samples)
levels(melted_tbl$abx_between_samples) <- list(no="0", yes="1")
df_abx_between = melted_tbl

```

Do sympotmatic kids have more abx before time_of_detection sample?

```{r}

temp_df2 = merge(num_abx_epi, kids_IDs_all, by = c("Child.ID","Date.of.collection","symptomatic","pre_detection"), all = T) 
keep_for_rf_all_abx = temp_df2
temp_df2 = temp_df2 %>% filter(pre_detection == F)
temp_df2$n_abx_doses[is.na(temp_df2$n_abx_doses)] = 0
keep_for_rf_all_abx$n_abx_doses[is.na(keep_for_rf_all_abx$n_abx_doses)] = 0
t.test(x = temp_df2[temp_df2$symptomatic == T,c('n_abx_doses')], y = temp_df2[temp_df2$symptomatic == F,c('n_abx_doses')])
temp_df2$symptomatic = as.factor(temp_df2$symptomatic)
levels(temp_df2$symptomatic) <- list(subclinical="FALSE", diarrheal="TRUE")
keep_for_rf_all_abx$symptomatic = as.factor(keep_for_rf_all_abx$symptomatic)
levels(keep_for_rf_all_abx$symptomatic) <- list(subclinical="FALSE", diarrheal="TRUE")
df_abx_total = temp_df2

```

Do sympotmatic kids have more abx before time_of_detection sample MONTH PRIOR ONLY?

```{r}

temp_df2 = merge(num_abx_epi_month, kids_IDs_all, by = c("Child.ID","Date.of.collection","symptomatic","pre_detection"), all = T) 
keep_for_rf_all_abx = temp_df2
temp_df2 = temp_df2 %>% filter(pre_detection == F)
temp_df2$n_abx_doses[is.na(temp_df2$n_abx_doses)] = 0
keep_for_rf_all_abx$n_abx_doses[is.na(keep_for_rf_all_abx$n_abx_doses)] = 0
t.test(x = temp_df2[temp_df2$symptomatic == T,c('n_abx_doses')], y = temp_df2[temp_df2$symptomatic == F,c('n_abx_doses')])
temp_df2$symptomatic = as.factor(temp_df2$symptomatic)
levels(temp_df2$symptomatic) <- list(subclinical="FALSE", diarrheal="TRUE")
keep_for_rf_all_abx$symptomatic = as.factor(keep_for_rf_all_abx$symptomatic)
levels(keep_for_rf_all_abx$symptomatic) <- list(subclinical="FALSE", diarrheal="TRUE")
df_abx_month = temp_df2

```

What happens if we break this down to look at a single drug at a time?

```{r}

num_abx_epi_per_drug = temp_df %>% group_by(Child.ID, sample.id, Number.of.crypto.positive.events,Date.of.birth, 
                         Date.of.collection, symptomatic, pre_detection) %>% mutate(n_distinct_abx = n()) %>%
  ungroup() %>% select(-min_drug_used_X_days_before_crypto) %>% distinct()
num_abx_epi_per_drug = merge(num_abx_epi_per_drug, kids_IDs_all, by = c("Child.ID","Date.of.collection","symptomatic","pre_detection"), all = T)
# replace missing values
num_abx_epi_per_drug$n_uses[is.na(num_abx_epi_per_drug$n_uses)] = 0
num_abx_epi_per_drug$n_distinct_abx[is.na(num_abx_epi_per_drug$n_distinct_abx)] = 0
# easier to replace missing values by coverting to wide 
num_abx_epi_per_drug = num_abx_epi_per_drug %>% select(-sample.id,-Number.of.crypto.positive.events, -Date.of.birth)
num_abx_epi_per_drug$Child.ID = as.integer(as.character(num_abx_epi_per_drug$Child.ID))
stats_df1 = tidyr::spread(num_abx_epi_per_drug %>% filter(), drug, n_uses); stats_df1[,ncol(stats_df1)] = NULL
stats_df1[is.na(stats_df1)] = 0
keep_for_rf_each_abx = stats_df1
stats_df2 = num_abx_epi_per_drug %>% select(-drug,-n_uses)
#stats
for (x in unique(stats_df1$drug)) {
  if (x != "Syp, Mebendazole") {
  temp_df3 = stats_df1 %>% filter(drug == x, pre_detection == F)
  res = t.test(x = temp_df3[temp_df3$symptomatic == T,c('n_uses')], y = temp_df3[temp_df3$symptomatic == F,c('n_uses')])
  if (res$p.value < 0.05) { print(x)}
  }}
melted_stats_df1 = melt(stats_df1, id.vars = c('Child.ID','Date.of.collection','symptomatic','pre_detection','n_distinct_abx'))
plt_df = melted_stats_df1 %>% filter(pre_detection == F)
plt_df$symptomatic = as.factor(plt_df$symptomatic)
levels(plt_df$symptomatic) <- list(subclinical="FALSE", diarrheal="TRUE")
levels(plt_df$variable) <- list(Cefixime="Cefixime sup (Cef-3)", Nystatin="Nystatin Suspension",
                                Amoxycillin="Syp. Amoxycillin",Azythromycin="Syp. Azythromycin",
                                Ciprofloxacin="Syp. Ciprofloxacin",Cotrimoxasole="Syp. Cotrimoxasole",
                                Erythromycin="Syp. Erythromycin",Mebendazole = "Syp, Mebendazole",
                                Flucloxacillin="Syp. Flucloxacillin",Metronidzole="Syp. Metronidzole",
                                `Nalidixic acid`="Syp. Nalidixic acid",Albendazole="Tab. Albendazole")
df_individual_abx = plt_df

```

# Save Abx history for other scripts

```{r}

save(list = c("keep_for_rf_all_abx","keep_for_rf_each_abx"), file = paste0(paste0("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P2output_" , format(Sys.time(), "%Y_%m_%d")),".RData"))

```

Also check number of distinct antibiotics

```{r}

t.test(x = stats_df2[stats_df2$symptomatic == T,c('n_distinct_abx')], y = stats_df2[stats_df2$symptomatic == F,c('n_distinct_abx')])
plt_df = stats_df2 %>% filter(pre_detection == F)
plt_df$symptomatic = as.factor(plt_df$symptomatic)
levels(plt_df$symptomatic) <- list(subclinical="FALSE", diarrheal="TRUE")
df_distinct_abx = plt_df

#log model symptoms ~ X1 * n_uses_abx1 + X2 * n_uses_abx2 + ...
#long to wide format spread(data, key, value)
wide_df1 = tidyr::spread(num_abx_epi_per_drug %>% filter(pre_detection == F) %>% 
                           select(Child.ID,Date.of.collection,symptomatic,drug,n_uses),
                         drug, n_uses)
wide_df2 = num_abx_epi_per_drug %>% filter(pre_detection == F) %>%
  select(Child.ID,Date.of.collection,symptomatic,n_distinct_abx) %>% distinct() 
wide_df1[is.na(wide_df1)] = 0
wide_df1$Child.ID = as.integer(as.character(wide_df1$Child.ID)); wide_df2$Child.ID = as.integer(as.character(wide_df2$Child.ID))
wide_df = merge(merge(wide_df1, wide_df2, by = c("Child.ID","Date.of.collection","symptomatic"), all = T), kids_IDs, 
                by = c("Child.ID","Date.of.collection","symptomatic"), all = T) 
wide_df[is.na(wide_df)] = 0
wide_df[,colnames(wide_df) == "<NA>"] = NULL
wide_df$symptomatic = as.factor(as.numeric(wide_df$symptomatic))
data_for_model = wide_df %>% select(-Child.ID,-Date.of.collection)
colnames(data_for_model)[colnames(data_for_model) == "Cefixime sup (Cef-3)"] = "Cefixime"
colnames(data_for_model)[colnames(data_for_model) == "Nystatin Suspension"] = "Nystatin"
colnames(data_for_model)[colnames(data_for_model) == "Syp. Amoxycillin"] = "Amoxycillin"
colnames(data_for_model)[colnames(data_for_model) == "Syp. Azythromycin"] = "Azythromycin"
colnames(data_for_model)[colnames(data_for_model) == "Syp. Ciprofloxacin"] = "Ciprofloxacin"
colnames(data_for_model)[colnames(data_for_model) == "Syp. Cotrimoxasole"] = "Cotrimoxasole"
colnames(data_for_model)[colnames(data_for_model) == "Syp. Erythromycin"] = "Erythromycin"
colnames(data_for_model)[colnames(data_for_model) == "Syp. Flucloxacillin"] = "Flucloxacillin"
colnames(data_for_model)[colnames(data_for_model) == "Syp. Metronidzole"] = "Metronidzole"
colnames(data_for_model)[colnames(data_for_model) == "Syp. Nalidixic acid"] = "Nalidixic_acid"
colnames(data_for_model)[colnames(data_for_model) == "Tab. Albendazole"] = "Albendazole"
abx_model = glm(data_for_model$symptomatic~.,family=binomial(link='logit'),data=data_for_model %>% select(-symptomatic))
summary(abx_model)

# rf = randomForest(data_for_model %>% select(-symptomatic), data_for_model$symptomatic,
#                   importance = TRUE, ntree = 10000, replace=TRUE, nodesize=1,
#                   classwt=c(10,0.00015))
# extracted_importance = importance(rf,type=2) # type 2 = node impurity, more useful for auc
# extracted_importance = extracted_importance[rownames(extracted_importance)[order(extracted_importance)],]
# top10 = tail(extracted_importance,10)
# names(top10)
# predictions = as.data.frame(predict(rf,type="prob"))
# predictions$predict <- names(predictions)[1:2][apply(predictions[,1:2], 1, which.max)]
# predictions$observed <- data_for_model$symptomatic
# roc <- roc(ifelse(predictions$observed==0, 0, 1), as.numeric(predictions$`0`))
# plot(roc, col = "gray60")
# AUC = auc(roc)
# AUC

```

# Save data for other scripts

```{r}

save(list = c("df_fig2A","df_sf2A", "df_fig2B","df_fig2C","df_fig3A","ps_for_explore_fig3A","alt_df","df_fig3D","alt_countsumsdf","ps_rare_bygenus_alt_fig","ps_for_alpha","ps_rare_for_beta","ps_rare.ord","df_abx_between","df_abx_total","df_abx_month","df_individual_abx","df_distinct_abx","abx_model"), file = paste0(paste0("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P2output_plot_" , format(Sys.time(), "%Y_%m_%d")),".RData"))

```

# Make figures

```{r}

#rmarkdown::render("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/code/R_markdown/P2_cohort_figures.Rmd")

```

### 2020-09-03 code/R_markdown/P2_cohort_figures.Rmd
added code and figures, did not add data or results - need to check consent re data availability

---
title: "Microbiome analysis - cohort figures"
author: "Maureen A. Carey"
date: "09/01/2020"
output: html_document
---

# Load packages and data

```{r setup, include=FALSE, echo=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/")

# load packages
library(ggplot2); library(rmarkdown); library(vegan); library(dotwhisker)
library(dplyr); library(phyloseq)

# set constants
SEED_INT = 10; set.seed(SEED_INT); options(scipen=10000); theme_set(theme_bw())
colors = c("#1F78B4","#FF7F00","#666666", 'light gray')
names(colors) = c('diarrheal','subclinical','no crypto','both')
# blue = "#1F78B4" orange = "#FF7F00" gray = "#666666"

# load data
load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P2output_plot_2020_09_02.RData")

```

## Figure 2 - Emphasizing amount of diarrhea and antibiotic usage

Figure 2A
``` {r}

df_plot = df_fig2A
colors_temp = colors; names(colors_temp)[names(colors_temp) == 'no crypto'] = "not in microbiome cohort"
colors_temp_use = colors_temp; colors_temp_use[names(colors_temp_use) != 'not in microbiome cohort'] = c("red","red","red")

p = ggplot(data = df_plot) +
  geom_histogram( aes(x = n_diarr, fill = TYPE), bins = 45)+
  scale_fill_manual(values = colors_temp_use) +
  theme(axis.text.x = element_text(angle=45,hjust=1),
        axis.text.y = element_text(size=12),
        axis.title.x = element_text(size=20),
        axis.title.y = element_text(size=20),
        plot.margin=unit(c(.5,.5,.5,.5),"cm"),
        legend.position = c(.6,.7),
        legend.title = element_blank()) + xlab('episodes of diarrhea') +
  guides(fill = F) +
  ylab('number of children')
p
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/Fig2A_diarrhea_freq_mirpur.png",
       p, width = 4, height = 4, dpi = 600)
rm(p, df_plot, colors_temp, colors_temp_use)

```

Supplemental Figure 2A
```{r}

n_kids = length(unique(df_sf2A$CHILDID))
df_sf2A %>% 
  ggplot(aes(x = age, y = (..count..)/n_kids)) +
  stat_bin(binwidth = 30) +
  geom_bar() + facet_wrap(~value_use)+
  theme(plot.margin=unit(c(.5,.5,.5,.5),"cm")) +
  geom_vline(xintercept = 365, linetype = "dotted") +
  ylab('percent of children') + xlab('age in days') + scale_y_continuous(labels = scales::percent)
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/Sfig2A_antibiotic_usage_mirpur.png",
       width = 6.8, height = 5,units = "in",dpi = 300)

```

Figure 2B
```{r}

# stats and plot
my_comparisons <- list(#c("no infection\ndetected", "subclinical"), 
                       c("no infection\ndetected", "both"),
                       c("subclinical","both"),
                       c("subclinical", "diarrheal"))
colors_temp = colors; colors_temp['no infection\ndetected'] = "#666666"
p = ggplot(data = df_fig2B, aes(x = TYPE, y = n, color = TYPE)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_point(alpha = 0.1, position = position_jitter(width = 0.3, height = 0.01, seed = SEED_INT)) +
  scale_color_manual(values = colors_temp) +
  theme(axis.text.x = element_text(angle = -45, hjust = 0, vjust = 1),
        plot.margin=unit(c(.25,.5,.5,.25),"cm")) + 
  ggpubr::stat_compare_means(comparisons = my_comparisons, size = 2, method = 't.test') +
  ylab('episodes of diarrhea (age 0-2yrs)') + xlab(NULL) + guides(color = F)
# SOMETHING IS WEIRD -> NUMBER OF DIARRHEAL POINTS IS TOO LOW - NEVERMIND \/
# this is not using the microbiome samples' definition of subclin/diarrhea, this is using the FULL dataset
p
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/Fig2B_num_diarrheal_events_mirpur.png",
       p , width =2, height = 3,units = "in",dpi = 300)

```

Figure 2C
```{r}

# stats and plot
my_comparisons <- list(#c("no infection\ndetected", "subclinical"), 
                       #c("no infection\ndetected", "both"), 
                       #c("subclinical", "both"), 
                       #c("diarrheal", "both"), 
                       c("no infection\ndetected", "diarrheal"))
                       #c("subclinical", "diarrheal"))
colors_temp = colors; colors_temp['no infection\ndetected'] = "#666666"
p = ggplot(data = df_fig2C, aes(x = TYPE, y = n, color = TYPE)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_point(alpha = 0.1, position = position_jitter(width = 0.3, height = 0.01, seed = SEED_INT)) +
  scale_color_manual(values = colors_temp) +
  theme(axis.text.x = element_text(angle = -45, hjust = 0, vjust = 1),
        plot.margin=unit(c(.25,.5,.5,.25),"cm")) + 
  ggpubr::stat_compare_means(comparisons = my_comparisons, size = 2, method = 't.test') +
  ylab('doses of antibiotics (age 0-2yrs)') + xlab(NULL) + guides(color = F)
# this is not using the microbiome samples' definition of subclin/diarrhea, this is using the FULL dataset
p
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/Fig2C_num_abx_events_mirpur.png",
       p , width =2, height = 3,units = "in",dpi = 300)

```

## Figure 3

Frequent reads overall - Figure 3A + 3D
```{r}

rare_min = 10000
p1 = ggplot(df_fig3A, 
           aes(x = names, y = nreads/(rare_min*nrow(sample_data(ps_for_explore_fig3A))), fill=genus)) + geom_bar(stat="identity")
p1 = p1 + theme(axis.text.x = element_text(size=14,angle=45,hjust=1),
              axis.text.y = element_text(size=14),
              axis.title.x = element_text(size=20),
              axis.title.y = element_text(size=20),
              plot.margin=unit(c(.5,.5,.5,.5),"cm")) + 
  labs(y="Fraction of all\nreads from ASV", x="ASV assigned genus") +
  scale_x_discrete(labels = df_fig3A$genus) + 
  scale_fill_grey() + guides(fill = F) + scale_y_continuous(labels = scales::percent)
p1
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/Fig3A_abundant_asvs_mirpur.png",
       plot=p1, width = 6.8,height = 4.2,units = "in",dpi = 300)

## to show lack of dominance of individual microbes

p_alt = ggplot(alt_df, aes(x = reorder(sample, percent_reads), y = percent_reads, fill = Genus)) +
  geom_bar(stat = "identity")
p_alt = p_alt + theme(axis.text.x = element_blank(),
              axis.ticks.x = element_blank(),
              axis.text.y = element_text(size=14),
              axis.title.x = element_text(size=18),
              axis.title.y = element_text(size=18),
              plot.margin=unit(c(.7,.5,.5,.5),"cm")) + 
  scale_fill_grey(na.value = "grey") +
  scale_y_continuous(labels = scales::percent, limits = c(0,1)) +
  labs(y="Percent of sample reads\nfrom most common ASV", x="samples") #+ guides(fill = F)

symptomatic.lab <- c("diarrheal", "subclinical"); names(symptomatic.lab) <- c("TRUE", "FALSE")
pre_detection.lab <- c("pre-detection", "time-of-detection"); names(pre_detection.lab) <- c("TRUE", "FALSE")
p = ggplot(alt_df, aes(x = reorder(sample, percent_reads), y = percent_reads)) +
  geom_bar(stat = "identity") + 
  facet_grid(symptomatic~pre_detection,labeller = labeller(
    symptomatic = symptomatic.lab, pre_detection = pre_detection.lab)) 
p = p + theme(axis.text.x = element_blank(),
              axis.ticks.x = element_blank(),
              axis.text.y = element_text(size=14),
              axis.title.x = element_text(size=18),
              axis.title.y = element_text(size=18),
              plot.margin=unit(c(.7,.5,.5,.5),"cm")) +
  scale_y_continuous(labels = scales::percent, limits = c(0,1)) + scale_fill_grey(na.value = "grey") +
  labs(y="Percent of sample reads\nfrom most common ASV", x="samples") #+ guides(fill = F)
p
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/SFig3_dominant_asvs.png", 
       plot=p, width=6.8, height = 4.2,units = "in",dpi = 300)

p = ggplot(df_fig3D, aes(x = names, y = nsamples/(nrow(sample_data(ps_for_explore_fig3A))), fill=genus)) + geom_bar(stat="identity")
p = p + theme(axis.text.x = element_text(size=14,angle=45,hjust=1),
              axis.text.y = element_text(size=14),
              axis.title.x = element_text(size=18),
              axis.title.y = element_text(size=18),
              plot.margin=unit(c(.7,.5,.5,.5),"cm")) + 
  labs(y="Percent of stool samples\ncontaining ASV", x="genus of ASV") + guides(fill = F) +
  scale_x_discrete(labels = df_fig3D$genus) + scale_fill_grey(na.value = "grey") + xlab(NULL) +
  geom_hline(yintercept = .5, linetype="dashed", 
                color = "gray", size=1) +
  scale_y_continuous(labels = scales::percent, limits = c(0,1))
p
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/Fig3D_samplecount_asvs_mirpur.png",
       plot=p, width=6.8, height = 4.2,units = "in",dpi = 300)

p = ggplot(alt_countsumsdf, aes(x = names, y = nsamples/(nrow(sample_data(ps_rare_bygenus))), fill=genus)) + geom_bar(stat="identity")
p = p + theme(axis.text.x = element_text(angle=45,hjust=1),
              axis.text.y = element_text(size=12),
              axis.title.x = element_text(size=20),
              axis.title.y = element_text(size=20),
              plot.margin=unit(c(.5,.5,.5,.5),"cm")) + 
  labs(y="Fraction of samples containing ASV", x="ASV assigned genus") +
  scale_x_discrete(labels = alt_countsumsdf$genus) + guides(fill = F) + #readsumsdf?
  scale_y_continuous(labels = scales::percent, limits = c(0,1))

```

Richness and Evenness - Figure 3B and 3C

```{r}

theme_set(theme_bw())
p_observed = plot_richness(ps_for_alpha,x='Clinical.type', measures = "Observed", color = 'Clinical.type') + 
  geom_boxplot(alpha=0.5, outlier.shape = NA) + 
  geom_point(size = 1,alpha = 0.25,position = position_jitter(w = 0.1, h = 0)) + 
  labs(y="Richness (observed ASVs)",x=NULL) + 
  theme(strip.text.x=element_blank(),
        axis.title.x = element_text(size=20), 
        axis.title.y = element_text(size=20),
        axis.text.x = element_text(size=12,angle=-45,hjust=0,vjust = 1),
        plot.margin=unit(c(.5,.5,.5,.5),"cm")) +
  scale_x_discrete(labels = c('prior to\nsubclinical', 'prior to\ndiarrheal', 'subclinical','diarrheal')) +
  scale_color_manual(values = c("#666666","#666666","#FF7F00","#1F78B4")) +
  guides(color = F)
p_observed$layers <- p_observed$layers[-1]
p_observed
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/Fig3B_Observed_div_mirpur.png",
       plot=p_observed,width=4.5, height = 4.2,units = "in",dpi = 300)

p_simpson = plot_richness(ps_for_alpha,measures = "InvSimpson", x='Clinical.type', color = 'Clinical.type') + 
  geom_boxplot(alpha=0.5, outlier.shape = NA) + 
  geom_point(size = 1,alpha = 0.25,position = position_jitter(w = 0.1, h = 0)) + 
  labs(y="Evenness (Inv. Simpson)",x=NULL) + 
  theme(strip.text=element_blank(),
        axis.title.x = element_text(size=20), 
        axis.title.y = element_text(size=20),
        axis.text.x = element_text(size=12,angle=-45,hjust=0,vjust = 1),
        plot.margin=unit(c(.5,.5,.5,.5),"cm")) +
  scale_x_discrete(labels = c('prior to\nsubclinical', 'prior to\ndiarrheal', 'subclinical','diarrheal')) +
  scale_color_manual(values = c("#666666","#666666","#FF7F00","#1F78B4")) +
  guides(color = F)
p_simpson$layers <- p_simpson$layers[-1]
p_simpson
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/Fig3C_invsimpson_mirpur.png",
       plot=p_simpson, width=4.5, height = 4.2,units = "in",dpi = 300)

```

## Figure 4A

```{r}

p = plot_ordination(
  physeq = ps_rare_for_beta,
  ordination = ps_rare.ord,
  color = "Clinical.type",
  shape = "Number.of.crypto.positive.events"
)   + 
  scale_color_manual(values = c("#666666","#FF7F00","#1F78B4"),
                     labels = c("12" = "diarrheal","10" = "subclinical","1" = "pre-detection") )+ # 12 is diarrhea, 1,2 are pre
  theme(plot.margin=unit(c(.5,.5,.5,.5),"cm"))  + 
  labs(color = "symptomatic?", shape = "Number of infections")
p
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/Fig4A_pcoa_mirpur_euclidean.png", 
       plot=p, width = 5.4,height = 3,units = "in",dpi = 600)

```

Do all kids have at least one abx between predetction and time of detection???

```{r}

ggplot(df_abx_between, aes(x = abx_between_samples, y = value, fill = symptomatic)) + geom_bar(stat = "identity",position = "dodge")+
  annotate(geom = "text", x = 1.8, y = 70, label = "Pearson's Chi-\nsquared test\np-value = NS") +
  scale_fill_manual(values= c(colors['diarrheal'],colors['subclinical'])) +
  ylab('number of children') + xlab('antibiotics between pre-detection\nand time-of-detection samples?') +
  guides(fill= guide_legend(title='symptomatic?'))

```

Do sympotmatic kids have more abx before time_of_detection sample?

```{r}

my_comparisons = list(c("subclinical","diarrheal"))
ggplot(df_abx_total,aes(x= symptomatic, y = n_abx_doses, color = symptomatic)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_point(alpha = 0.5, position = position_jitter(width = 0.3, height = 0, seed = SEED_INT)) +
  scale_color_manual(values = c(colors['diarrheal'],colors['subclinical'])) +
  ylab('number of antibiotic doses') + xlab('symptomatic?') + guides(color = F) +
  ggpubr::stat_compare_means(comparisons = my_comparisons, size = 3, method = 't.test')

```

Do sympotmatic kids have more abx before time_of_detection sample MONTH PRIOR ONLY?

```{r}

my_comparisons = list(c("subclinical","diarrheal"))
ggplot(df_abx_month,aes(x= symptomatic, y = n_abx_doses, color = symptomatic)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_point(alpha = 0.5, position = position_jitter(width = 0.3, height = 0, seed = SEED_INT)) +
  scale_color_manual(values = c(colors['diarrheal'],colors['subclinical'])) +
  ylab('number of antibiotic doses') + xlab('symptomatic?') + guides(color = F) +
  ggpubr::stat_compare_means(comparisons = my_comparisons, size = 3, method = 't.test')
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/SFig2B_abx_MONTH_mirpur.png", 
              width = 3, height = 4, dpi = 600)

```

What happens if we break this down to look at a single drug at a time?

```{r}

ggplot(df_individual_abx, aes(x = value,y = variable, color = symptomatic)) + 
  geom_boxplot(position = "dodge", outlier.shape = NA) + 
  geom_point(alpha = 0.1, 
             position = position_jitterdodge(jitter.width = 0.3, jitter.height = 0, seed = SEED_INT, dodge.width = 0.75)) +
  scale_color_manual(values = c(colors['diarrheal'],colors['subclinical'])) +
  ylab('Antibiotics') + xlab('number of doses') +
  guides(color= guide_legend(title='symptomatic?')) +
  theme(legend.position = c(.8,.15))

```

Also check number of distinct antibiotics

```{r}

p = ggplot(df_distinct_abx, aes(x = n_distinct_abx,y = symptomatic, color = symptomatic)) + 
  geom_boxplot(position = "dodge", outlier.colour = NA) + geom_point(alpha = 0.2, position = position_jitter(width = 0.3, height = 0.3, seed = SEED_INT)) +
  scale_color_manual(values = c(colors['diarrheal'],colors['subclinical'])) + guides(color = F) +
  xlab('number of distinct antibiotics used') + ylab('symptomatic?')
p

dwplot(abx_model, vline = geom_vline(xintercept = 0, colour = "grey60", linetype = 2)) %>%
  relabel_predictors(c(n_distinct_abx = "number of distinct antibiotics",
                       Nalidixic_acid = "Nalidixic acid")) +
     theme_bw() + xlab("Coefficient Estimate") + ylab("") +
  theme(plot.title = element_text(face="bold"),
        plot.margin = unit(c(1.5,1,1,1), "cm")) + 
  scale_color_manual(values = c("black")) + 
  guides(color = F)

```

### 2020-09-03 code/R_markdown/P3_machine_learning.Rmd
added code and figures, did not add data or results - need to check consent re data availability

---
title: "Microbiome analysis - machine learning"
author: "Maureen A. Carey"
date: "09/01/2020"
output: html_document
---

# Load packages and data

```{r setup, include=FALSE, echo=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/")

# load packages
library(phyloseq); library(reshape2); library(tibble)
library(randomForest); library(rmarkdown); library(dplyr); library(tidyr); library(pROC)

# set constants
SEED_INT = 10; set.seed(SEED_INT); options(scipen=10000)
colors = c("#1F78B4","#FF7F00","#666666", 'light gray')
names(colors) = c('diarrheal','subclinical','no crypto','both')
# blue = "#1F78B4" orange = "#FF7F00" gray = "#666666"

# load data
load("/Users/maureencarey/local_documents/work/crypto_microbiome/data_cannot_be_shared/data.RData")
total_diarrhea = total_diarrhea  %>% filter(CHILDID %in% burden_mirpur$SID) # remaining kids never were sick
all_cause_diar = total_diarrhea %>% group_by(CHILDID,study) %>% summarize(episodes = n()) %>% ungroup()
rm(list= ls()[! (ls() %in% c('all_cause_diar','SEED_INT','colors'))])
load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P1output_2020_09_02.RData")
load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P2output_2020_09_02.RData")

# make useful functions
randomForest_res = function(x,y,N) { 
  # x = variables used to predict class, y = classs labels, N = number imp variables you want
  # THIS ONLY WORKS IF YOU"RE CLASSIFYING 10 v. 12 (subclinical v. diarrheal)
  rf_res_object = randomForest(x,y,importance = TRUE, ntree = 50*dim(x)[[2]], replace=TRUE, nodesize=1, classwt=c(0.0001,100))

  extracted_importance = importance(rf_res_object,type=2) # type 2 = node impurity, more useful for auc
  extracted_importance = extracted_importance[rownames(extracted_importance)[order(extracted_importance)],]
  topN = tail(extracted_importance,N)

  predictions = as.data.frame(predict(rf_res_object,type="prob"))
  predictions$predict = names(predictions)[1:2][apply(predictions[,1:2], 1, which.max)]
  predictions$observed = y

  ROC_res = roc(ifelse(predictions$observed==10, 10, 12), as.numeric(predictions$`10`))
  AUC_res = auc(ROC_res)
  return(list(rf_res_object,ROC_res,AUC_res,topN))  
}

```

# Can we predict whether a sample was sub-clinical or diarrheal from predetection microbiota?

```{r}

compare <- c(1,2) # predetection -> +/- symptoms
ps_for_class <- subset_samples(ps_rare_diff, Clinical.type %in% compare)

x <- cbind(otu_table(ps_for_class), as(sample_data(ps_for_class),"data.frame") %>% select(Age.in.days))
y <- as.factor(sample_data(ps_for_class)$Clinical.type)
rf <- randomForest(x,y,importance = TRUE, ntree = 50*dim(x)[[2]], replace=TRUE, nodesize=1, classwt=c(0.00015,130))
save(list = c("rf"), file = paste0(paste0("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/rf_PD_" , format(Sys.time(), "%Y_%m_%d")),".RData"))   
x_for_PD = x
y_for_PD = y

extracted_importance = importance(rf,type=2) # type 2 = node impurity, more useful for auc
extracted_importance = extracted_importance[rownames(extracted_importance)[order(extracted_importance)],]
top10_PD = tail(extracted_importance,10)

predictions = as.data.frame(predict(rf,type="prob"))
predictions$predict <- names(predictions)[1:2][apply(predictions[,1:2], 1, which.max)]
predictions$observed <- y

roc_PD <- roc(ifelse(predictions$observed==1, 1, 2), as.numeric(predictions$`1`))
AUC_PD = auc(roc_PD)
AUC_PD

rm(x,y,rf,predictions,extracted_importance,ps_for_class)

```

# Can we predict whether a sample was sub-clinical or diarrheal from TOD microbiota?

```{r}

compare <- c(10,12) # TOD +/- symptoms
ps_for_class <- subset_samples(ps_rare_diff, Clinical.type %in% compare)
x <- cbind(otu_table(ps_for_class), as(sample_data(ps_for_class),"data.frame") %>% select(Age.in.days))
y <- as.factor(sample_data(ps_for_class)$Clinical.type)
x_for_TOD = x
y_for_TOD = y

rf = randomForest_res(x_for_TOD,y_for_TOD,10)
top10_TOD = rf[[4]]
roc_TOD = rf[[2]]
AUC_TOD = rf[[3]]
AUC_TOD

rf = rf[[1]]
# use this on rivanna for faster computing
save(list = c("rf","x","y"), file = paste0(paste0("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/rf_TOD_" , format(Sys.time(), "%Y_%m_%d")),".RData"))  
rm(x,y,rf,ps_for_class)

```

# Save data for other scripts

```{r}

save(list = c("top10_TOD","top10_PD"), file = paste0(paste0("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P3output_p1_" , format(Sys.time(), "%Y_%m_%d")),".RData"))

```

## DATA NOT SHOWN: Can we predict whether a sample was a first or second infection from TOD microbiota?

```{r}

# time-of-detection
compare <- c(10,12) # TOD +/- symptoms
ps_for_class <- subset_samples(ps_rare_diff, Clinical.type %in% compare)
x <- cbind(otu_table(ps_for_class), as(sample_data(ps_for_class),"data.frame") %>% select(Age.in.days))
y <- as.numeric(as.character(sample_data(ps_for_class)$Number.of.crypto.positive.events))
y[y!=1] = 2 # merge third infections with second

temp_df = as(sample_data(ps_for_class),"data.frame") %>% 
  select(Number.of.crypto.positive.events, symptomatic) %>%
  mutate(n_crypto = ifelse(Number.of.crypto.positive.events == 1, '1', '2+')) %>% select(-Number.of.crypto.positive.events) %>%
  table() %>% as.data.frame() %>%
  mutate(symptomatic_use = ifelse(symptomatic == T, 'diarrheal','subclinical')) %>%
  select(-symptomatic) %>% dplyr::rename(symptomatic = symptomatic_use)
chi_df = temp_df

rf = randomForest(x,as.factor(y),importance = TRUE, ntree = 50*dim(x)[[2]], replace=TRUE, nodesize=1, classwt=c(0.1,100))
rf
save(list = c("rf"), file = paste0(paste0("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/rf_TOD_1stv2nd_" , format(Sys.time(), "%Y_%m_%d")),".RData"))

extracted_importance = importance(rf,type=2) # type 2 = node impurity, more useful for auc
extracted_importance = extracted_importance[rownames(extracted_importance)[order(extracted_importance)],]
topN_first_infection = tail(extracted_importance,10)

predictions = as.data.frame(predict(rf,type="prob"))
predictions$predict = names(predictions)[1:2][apply(predictions[,1:2], 1, which.max)]
predictions$observed = y

ROC_first_infection = roc(ifelse(predictions$observed==1, 1, 2), as.numeric(predictions$`1`))
auc(ROC_first_infection)

rm(x,y,rf,ps_for_class)

```

## DATA NOT SHOWN: Can we predict whether a sample was a first or second infection from PD microbiota?

```{r}

# predetection
compare <- c(1,2) # PD +/- symptoms
ps_for_class <- subset_samples(ps_rare_diff, Clinical.type %in% compare)
x <- cbind(otu_table(ps_for_class), as(sample_data(ps_for_class),"data.frame") %>% select(Age.in.days))
y <- as.numeric(as.character(sample_data(ps_for_class)$Number.of.crypto.positive.events))
y[y!=1] = 2 # merge third infections with second
rf = randomForest(x,as.factor(y),importance = TRUE, ntree = 50*dim(x)[[2]], replace=TRUE, nodesize=1, classwt=c(0.1,100))
rf
save(list = c("rf"), file = paste0(paste0("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/rf_PD_1stv2nd_" , format(Sys.time(), "%Y_%m_%d")),".RData"))

extracted_importance = importance(rf,type=2) # type 2 = node impurity, more useful for auc
extracted_importance = extracted_importance[rownames(extracted_importance)[order(extracted_importance)],]
topN_first_infection_PD = tail(extracted_importance,10)

predictions = as.data.frame(predict(rf,type="prob"))
predictions$predict = names(predictions)[1:2][apply(predictions[,1:2], 1, which.max)]
predictions$observed = y

ROC_first_infection = roc(ifelse(predictions$observed==1, 1, 2), as.numeric(predictions$`1`))
auc(ROC_first_infection)

rm(x,y,rf,ps_for_class)

```

## DATA NOT SHOWN: Can we predict all cause diarrhea from TOD microbiota?

```{r}

compare <- c(10,12) # TOD +/- symptoms
ps_for_class2 <- subset_samples(ps_rare_diff, Clinical.type %in% compare)
all_cause_diar_temp = all_cause_diar %>% filter(CHILDID %in% sample_data(ps_for_class2)$Child.ID)
all_cause_diar_temp = merge(all_cause_diar_temp, as(sample_data(ps_for_class2)[,1:2], "data.frame"), by.x = "CHILDID", by.y = "Child.ID", all.y = T) %>% select(-study,-CHILDID)
sample_data(ps_for_class2)$episodes = all_cause_diar_temp$episodes[order(match(all_cause_diar_temp$sample_id_mod, sample_data(ps_for_class2)$sample_id_mod))]
x <- cbind(otu_table(ps_for_class2), as(sample_data(ps_for_class2),"data.frame") %>% select(Age.in.days))
y <- sample_data(ps_for_class2)$episodes
## WHY IS THERE AN NA
y[is.na(y)] = 0

rf = randomForest(x,y,importance = TRUE, ntree = 50*dim(x)[[2]], replace=TRUE, nodesize=1, classwt=c(0.0001,100))
rf

save(list = c("rf"), file = paste0(paste0("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/rf_all_diar_fromTOD_" , format(Sys.time(), "%Y_%m_%d")),".RData"))

rm(x,y,rf,ps_for_class2)

```

## DATA NOT SHOWN: Can we predict all cause diarrhea from PD microbiota?

```{r}

compare <- c(1,2) # PD +/- symptoms
ps_for_class2 <- subset_samples(ps_rare_diff, Clinical.type %in% compare)
all_cause_diar_temp = all_cause_diar %>% filter(CHILDID %in% sample_data(ps_for_class2)$Child.ID)
all_cause_diar_temp = merge(all_cause_diar_temp, as(sample_data(ps_for_class2)[,1:2], "data.frame"), by.x = "CHILDID", by.y = "Child.ID", all.y = T) %>% select(-study,-CHILDID)
sample_data(ps_for_class2)$episodes = all_cause_diar_temp$episodes[order(match(all_cause_diar_temp$sample_id_mod, sample_data(ps_for_class2)$sample_id_mod))]
x <- cbind(otu_table(ps_for_class2), as(sample_data(ps_for_class2),"data.frame") %>% select(Age.in.days))
y <- sample_data(ps_for_class2)$episodes
## WHY IS THERE AN NA
y[is.na(y)] = 0

rf = randomForest(x,y,importance = TRUE, ntree = 50*dim(x)[[2]], replace=TRUE, nodesize=1, classwt=c(0.0001,100))
rf

save(list = c("rf"), file = paste0(paste0("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/rf_all_diar_fromPD_" , format(Sys.time(), "%Y_%m_%d")),".RData"))

rm(x,y,rf,ps_for_class2)

```

## Can we predict whether a sample was sub-clinical or diarrheal from PD & TOD microbiota?
# not a great analysis because trees don't scale well to compare AUC

```{r}

compare <- c(10,12) # TOD +/- symptoms
ps_for_class <- subset_samples(ps_rare_diff, Clinical.type %in% compare)
ps_for_class <- subset_samples(ps_for_class, !is.na(previous_surv))
ps_for_surv <- subset_samples(ps_rare_diff, sample.id %in% sample_data(ps_for_class)$previous_surv)
taxa_names(ps_for_surv) <- paste0("pre_", taxa_names(ps_for_surv))
time_of_detect_microbes = otu_table(ps_for_class)
surv_microbes = otu_table(ps_for_surv)
# reorder the surveilance to match the TOD
proper_order = sample_data(ps_for_class)$previous_surv
surv_microbes = surv_microbes[proper_order]

x <- cbind(cbind(time_of_detect_microbes,surv_microbes), as(sample_data(ps_for_class),"data.frame") %>% select(Age.in.days))
y = as.factor(sample_data(subset_samples(ps_for_class, Clinical.type %in% compare))$Clinical.type)
rf = randomForest_res(x,y,15)
rf[[1]]

top_15_for_all_microbiome = rf[[4]]
roc_allM = rf[[2]]
AUC_allM = rf[[3]]
AUC_allM
rm(rf, x,y,proper_order,ps_for_surv)

```

## Can we predict whether a sample was sub-clinical or diarrheal from PD & TOD microbiota AND metadata?
# not a good analysis because trees don't scale well to compare AUC

```{r}

meta_for_rf = as(sample_data(ps_for_class)[,c( "rainfall","drain","food_deficit",
                                                #"toilet", # too many NAs
                                                "drinking_water_source","income_mo",
                                                "water_treat","siblings_under_5",
                                                #"CREM_rs2148483_A","PRKCA_rs58296998_T", #too many NAs
                                                "month.of.birth","year.of.birth",
                                                "month.of.collection",
                                                "year.of.collection", "Any..Giardia",
                                                "Any..E.histolytica",
                                                "WAZ","WHZ","HAZ",#"MUACZ",
                                               "BAZ",
                                                "Number.of.crypto.positive.events",
                                                "Age.in.days")],"data.frame")
meta_for_rf$water_treat = as.factor(meta_for_rf$water_treat)

x = cbind(time_of_detect_microbes,meta_for_rf,surv_microbes)
y = as.factor(sample_data(subset_samples(ps_for_class, Clinical.type %in% compare))$Clinical.type)
rf = randomForest_res(x,y,20)
rf[[1]]

top15_without_abx = rf[[4]]
roc_allM_M = rf[[2]]
AUC_allM_M = rf[[3]]
AUC_allM_M

rm(rf,x,y,ps_for_class)

```

# Prep Abx data for integration into model

```{r}

for_metadata = subset_samples(ps_rare_diff, Clinical.type %in% c(10,12))
for_metadata = subset_samples(for_metadata, !is.na(previous_surv))
temp_df = as(sample_data(for_metadata)[,c("sample.id")],"data.frame") 
rownames(temp_df) = temp_df$sample.id
temp_df2 = as(sample_data(for_metadata)[,c("Child.ID","Date.of.collection","sample.id")],"data.frame")
# kids with zero ABx don't have their Run# in these dataframes - cut those kids and then add after
keep_for_rf_all_abx2 = merge(keep_for_rf_all_abx %>% filter(!is.na(sample.id)), temp_df, 
                            by = 'sample.id', all.y = T) %>% select(sample.id,n_abx_doses)
keep_for_rf_all_abx2$n_abx_doses[is.na(keep_for_rf_all_abx2$n_abx_doses)] = 0
rownames(keep_for_rf_all_abx2) = keep_for_rf_all_abx2$sample.id; keep_for_rf_all_abx2$sample.id = NULL

keep_for_rf_each_abx2 = merge(keep_for_rf_each_abx,temp_df2, 
                            by = c("Child.ID","Date.of.collection"), all.y = T) %>%
  select(-Child.ID,-Date.of.collection,-symptomatic,-pre_detection)
keep_for_rf_each_abx2 = keep_for_rf_each_abx2[!is.na(keep_for_rf_each_abx2$sample.id),]
rownames(keep_for_rf_each_abx2) = keep_for_rf_each_abx2$sample.id; keep_for_rf_each_abx2$sample.id = NULL

abx_history = cbind(keep_for_rf_all_abx2,keep_for_rf_each_abx2)
rm(keep_for_rf_all_abx2,keep_for_rf_each_abx2)

#get metadata for only samples that have abx history
limited_metadata = as(sample_data(for_metadata)[,c("rainfall","drain","food_deficit",
                                                "drinking_water_source","income_mo",
                                                "water_treat","siblings_under_5",
                                                "month.of.birth","year.of.birth",
                                                "month.of.collection",
                                                "year.of.collection", "Any..Giardia",
                                                "Any..E.histolytica",
                                                "WAZ","WHZ","HAZ","BAZ",
                                                "Number.of.crypto.positive.events",
                                                "Age.in.days")],"data.frame")
limited_metadata$water_treat = as.factor(limited_metadata$water_treat)

```

# Can we predict whether a sample was sub-clinical or diarrheal from TOD microbiome AND ABx data?

```{r}

for_metadata = subset_samples(ps_rare_diff, Clinical.type %in% c(10,12))
for_metadata = subset_samples(for_metadata, !is.na(previous_surv))
temp_df = as(sample_data(for_metadata)[,c("sample.id")],"data.frame") 
rownames(temp_df) = temp_df$sample.id
temp_df2 = as(sample_data(for_metadata)[,c("Child.ID","Date.of.collection","sample.id")],"data.frame")
# kids with zero ABx don't have their Run# in these dataframes - cut those kids and then add after
keep_for_rf_all_abx2 = merge(keep_for_rf_all_abx %>% filter(!is.na(sample.id)), temp_df, 
                            by = 'sample.id', all.y = T) %>% select(sample.id,n_abx_doses)
keep_for_rf_all_abx2$n_abx_doses[is.na(keep_for_rf_all_abx2$n_abx_doses)] = 0
rownames(keep_for_rf_all_abx2) = keep_for_rf_all_abx2$sample.id; keep_for_rf_all_abx2$sample.id = NULL

keep_for_rf_each_abx2 = merge(keep_for_rf_each_abx,temp_df2, 
                            by = c("Child.ID","Date.of.collection"), all.y = T) %>%
  select(-Child.ID,-Date.of.collection,-symptomatic,-pre_detection)
keep_for_rf_each_abx2 = keep_for_rf_each_abx2[!is.na(keep_for_rf_each_abx2$sample.id),]
rownames(keep_for_rf_each_abx2) = keep_for_rf_each_abx2$sample.id; keep_for_rf_each_abx2$sample.id = NULL

abx_history = cbind(keep_for_rf_all_abx2,keep_for_rf_each_abx2)
rm(keep_for_rf_all_abx2,keep_for_rf_each_abx2)

x_withabx = cbind(cbind(abx_history, otu_table(for_metadata)), as(sample_data(for_metadata),"data.frame") %>% select(Age.in.days))
y = as.factor(sample_data(for_metadata)$Clinical.type)

rf = randomForest_res(x_withabx,y,25)
rf[[1]]

top20_TOD_with_abx = rf[[4]]
roc_TOD_abx = rf[[2]]
AUC_TOD_abx = rf[[3]]
AUC_TOD_abx

save(list = c("rf"), file = paste0(paste0("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/rf_TOD_abx_" , format(Sys.time(), "%Y_%m_%d")),".RData"))

rm(rf,x_withabx,y)

```

## DATA NOT SHOWN: Can we predict whether a sample was sub-clinical or diarrheal from metadata AND ABx data?

```{r}

x = cbind(limited_metadata,abx_history)
y = as.factor(sample_data(for_metadata)$Clinical.type)

rf = randomForest_res(x,y,15)
rf[[1]]

top15_metadata_with_abx = rf[[4]]
roc_meta_abx = rf[[2]]
AUC_meta_abx = rf[[3]]
AUC_meta_abx

save(list = c("rf"), file = paste0(paste0("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/rf_meta_abx_" , format(Sys.time(), "%Y_%m_%d")),".RData"))

rf = randomForest_res(limited_metadata,y,15)
rf[[1]]
top15_metadata_without_abx = rf[[4]]
roc_meta_no_abx = rf[[2]]
AUC_meta_no_abx = rf[[3]]
AUC_meta_no_abx

save(list = c("rf"), file = paste0(paste0("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/rf_meta_" , format(Sys.time(), "%Y_%m_%d")),".RData"))

rm(rf,x,y,limited_metadata)

roc_list_mirpur = list(roc_meta_abx,roc_meta_no_abx)
names(roc_list_mirpur) = c("metadata","metadata &\nantibiotics")

```

# Prep for Figure 4D: Plot top 10 most important variables wth both microbiota classifiers

```{r}

## gini
plt_df1 = as.data.frame(top10_TOD) %>% rownames_to_column('variable') %>% dplyr::rename(value = top10_TOD) 
plt_df2 = as.data.frame(top10_PD) %>%rownames_to_column('variable') %>% dplyr::rename(value = top10_PD)
plt_df = rbind(plt_df1 %>% mutate(model = 'time-of-detection'),plt_df2 %>% mutate(model = 'pre-detection'))
plt_df$model = as.factor(plt_df$model)
save_df1 = plt_df1
save_df2 = plt_df2
save(list = c("save_df1","save_df2"), file = paste0(paste0("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P3output_p2_" , format(Sys.time(), "%Y_%m_%d")),".RData"))

tax = tax_table(ps_rare_diff)
tax = tax[rownames(tax) %in% plt_df$variable,]
tax = merge(tax,plt_df, by.x= 'row.names', by.y = "variable", all.x = F, all.y = T)
tax$Genus = as.character(tax$Genus)
tax2 = tidyr::spread(tax, key = model, value = value)  %>% # this line is new here
  transform(id=as.numeric(factor(Row.names))) %>%
  mutate(var = ifelse(is.na(Family),
                      as.character(Row.names),
                      ifelse(is.na(Genus), paste0(paste0(paste0(paste0(as.character('ASV'),id),' ([f]'),Family),')'),
                             paste0(paste0(paste0(paste0(as.character('ASV'),id),' ('),Genus),')'))))

plot_tax = tax2 %>% select(pre.detection, time.of.detection, var) %>% melt(id.vars = "var", na.rm = T)
plot_tax$var = as.factor(plot_tax$var)
plot_tax$var = reorder(plot_tax$var, -plot_tax$value)
plot_tax$var <- relevel(plot_tax$var, "ASV19 (Megasphaera)")
plot_tax_4D = plot_tax

```

## Save data for interaction plots - interested but DATA NOT SHOWN (Also, needs to be performed on a computer with more mem than my laptop)

```{r}

# save data for interaction plots
ASV = tax2[!(is.na(tax2$Genus))& tax2$Genus == "Megasphaera",1]
ASV_10 = tax2[,1]

tax = tax_table(ps_rare_diff) %>% as.data.frame()
tax$Genus = as.character(tax$Genus)
save_tax = tax %>% 
  rownames_to_column("Row.names") %>%
  transform(id=as.numeric(factor(Row.names))) %>%
  mutate(var = ifelse(is.na(Family),
                      as.character(Row.names),
                      ifelse(is.na(Genus), paste0(paste0(paste0(paste0(as.character('ASV'),id),' ([f]'),Family),')'),
                             paste0(paste0(paste0(paste0(as.character('ASV'),id),' ('),Genus),')')))) 
save(list = c("ASV", "save_tax"), file = paste0("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/megasphaera_ASV_"
              , format(Sys.time(), "%Y_%m_%d")
              , ".RData"))
save(list = c("ASV_10", "save_tax"), file = paste0("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/top_ASVs_"
              , format(Sys.time(), "%Y_%m_%d")
              , ".RData"))

```

## DATA NOT SHOWN: Plot important variables (raw values)

```{r}

#extract ASV that are in Tax2 from predetection otu_table
otu_PD = as.data.frame(x_for_PD)[,plt_df2$variable] %>% rownames_to_column('sample') %>% mutate(symptomatic = y_for_PD)
#extract ASV that are in Tax2 from TOD otu_table
otu_TOD = as.data.frame(x_for_TOD)[,plt_df1$variable] %>% rownames_to_column('sample') %>%mutate(symptomatic = y_for_TOD)
ids = as(sample_data(ps_rare_diff),"data.frame") %>% select(Child.ID) %>% rownames_to_column('sample')
otu_PD = merge(otu_PD, ids, all.x = T, all.y = F, by = "sample")
otu_TOD = merge(otu_TOD, ids, all.x = T, all.y = F, by = "sample")

plt_df = melt(otu_TOD) %>%
  group_by(variable) %>%
  mutate(value_use = value/max(value)) %>%
  ungroup() %>%
  mutate(symptomatic_use = ifelse(symptomatic == '10','subclinical','diarrheal')) %>%
  tidyr::separate(sample, c("sample", "remove_extra")) %>%
  select(-remove_extra)

tax = tax_table(ps_rare_diff)
tax = tax[rownames(tax) %in% plt_df$variable,]
tax = merge(tax,plt_df, by.x= 'row.names', by.y = "variable", all.x = F, all.y = T)
tax$Genus = as.character(tax$Genus)
tax2 = tax %>% 
  transform(id=as.numeric(factor(Row.names))) %>%
  mutate(var = ifelse(is.na(Family),
                      as.character(Row.names),
                      ifelse(is.na(Genus), paste0(paste0(paste0(paste0(as.character('ASV'),id),' ([f]'),Family),')'),
                             paste0(paste0(paste0(paste0(as.character('ASV'),id),' ('),Genus),')'))))
#ASV10 (Collinsella) for PD
#ASV5 (Parabacteroides) for TOD
TOD_twoASVs = tax2 %>% filter(var == 'ASV10 (Megasphaera)' | var == 'ASV5 (Parabacteroides)')

tax2_TOD = tax2
rm(tax, tax2, plt_df)

plt_df = melt(otu_PD) %>%
  group_by(variable) %>%
  mutate(value_use = value/max(value)) %>%
  ungroup() %>%
  mutate(symptomatic_use = ifelse(symptomatic == '1','subclinical','diarrheal')) %>%
  tidyr::separate(sample, c("sample", "remove_extra")) %>%
  select(-remove_extra)

tax = tax_table(ps_rare_diff)
tax = tax[rownames(tax) %in% plt_df$variable,]
tax = merge(tax,plt_df, by.x= 'row.names', by.y = "variable", all.x = F, all.y = T)
tax$Genus = as.character(tax$Genus)
tax2 = tax %>% 
  transform(id=as.numeric(factor(Row.names))) %>%
  mutate(var = ifelse(is.na(Family),
                      as.character(Row.names),
                      ifelse(is.na(Genus), paste0(paste0(paste0(paste0(as.character('ASV'),id),' ([f]'),Family),')'),
                             paste0(paste0(paste0(paste0(as.character('ASV'),id),' ('),Genus),')'))))
#ASV10 (Collinsella) for PD
#ASV5 (Parabacteroides) for TOD
PD_twoASVs = tax2 %>% filter(var == 'ASV10 (Megasphaera)' | var == 'ASV3 (Collinsella)')
tax2_PD = tax2

rm(tax, tax2, plt_df)

```

# Prep for Figure 4E - important variables

```{r}

twoASVs = rbind(TOD_twoASVs %>% filter(var == "ASV10 (Megasphaera)") %>% mutate(time = 'time-of-detection'), 
                PD_twoASVs %>% filter(var == "ASV10 (Megasphaera)") %>% mutate(time = 'pre-detection'))

# trying to plot TOD on y axis and PD on x axis
temp1 = twoASVs %>% select(value_use,symptomatic_use, Genus,time, Child.ID) %>% filter(time == 'time-of-detection') %>% select(-time) %>% dplyr::rename(`time-of-detection` = value_use)
temp2 = twoASVs %>% select(value_use,symptomatic_use, Genus,time, Child.ID) %>% filter(time == 'pre-detection') %>% select(-time) %>% dplyr::rename(`pre-detection` = value_use)
two_WIDE = merge(temp1, temp2, all = T)
colnames(two_WIDE)[colnames(two_WIDE) == 'symptomatic_use'] = 'symptomatic?'

twoASVs_TOD_wide = tidyr::spread(TOD_twoASVs %>% select(value_use,symptomatic_use,sample, Genus),
                             key = Genus, value = value_use) %>% 
  mutate(`symptomatic?` = as.factor(symptomatic_use))

twoASVs_PD_wide = tidyr::spread(PD_twoASVs %>% select(value_use,symptomatic_use,sample, Genus),
                             key = Genus, value = value_use) %>% 
  mutate(`symptomatic?` = as.factor(symptomatic_use))

count_df_PD_mirpur = twoASVs_PD_wide
count_df_TOD_mirpur = twoASVs_TOD_wide
save(list = c("count_df_PD_mirpur","count_df_TOD_mirpur"), file = paste0(paste0("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P3output_p3_" , format(Sys.time(), "%Y_%m_%d")),".RData"))

```

## DATA NOT SHOWN: Imp var for first v later infection

```{r}

plt_df1 = as.data.frame(topN_first_infection) %>% rownames_to_column('variable') %>% dplyr::rename(value = topN_first_infection) 
plt_df2 = as.data.frame(topN_first_infection_PD) %>%rownames_to_column('variable') %>% dplyr::rename(value = topN_first_infection_PD)
plt_df = rbind(plt_df1 %>% mutate(model = 'time-of-detection'),plt_df2 %>% mutate(model = 'pre-detection'))
plt_df$model = as.factor(plt_df$model)

tax = tax_table(ps_rare_diff)
tax = tax[rownames(tax) %in% plt_df$variable,]
tax = merge(tax,plt_df, by.x= 'row.names', by.y = "variable", all.x = F, all.y = T)
tax$Genus = as.character(tax$Genus)
tax2 = tax %>% 
  transform(id=as.numeric(factor(Row.names))) %>%
  mutate(var = ifelse(is.na(Family),
                      as.character(Row.names),
                      ifelse(is.na(Genus), paste0(paste0(paste0(paste0(as.character('ASV'),id),' ([f]'),Family),')'),
                             paste0(paste0(paste0(paste0(as.character('ASV'),id),' ('),Genus),')'))))

tax2_infect = tax2
rm(tax, tax2, plt_df)

```

# Prep Supplemental Figure 5 - ROC and Important variables, Adding in Abx to model

```{r}

plt_df1 = as.data.frame(top10_TOD) %>%rownames_to_column('variable') %>% dplyr::rename(value = top10_TOD) 
plt_df2 = as.data.frame(top20_TOD_with_abx) %>%rownames_to_column('variable') %>% dplyr::rename(value = top20_TOD_with_abx) %>% top_n(10,value)
plt_df = rbind(plt_df1 %>% mutate(model = 'without abx'),plt_df2 %>% mutate(model = 'with abx'))
plt_df$model = as.factor(plt_df$model)

tax = tax_table(ps_rare_diff)
tax = tax[rownames(tax) %in% plt_df$variable,]
tax = merge(tax,plt_df, by.x= 'row.names', by.y = "variable", all.x = F, all.y = T)
tax$Genus = as.character(tax$Genus)
tax2 = tax %>% 
  transform(id=as.numeric(factor(Row.names))) %>%
  mutate(var = ifelse(is.na(Family),
                      as.character(Row.names),
                      ifelse(is.na(Genus), paste0(paste0(paste0(paste0(as.character('ASV'),id),' ([f]'),Family),')'),
                             paste0(paste0(paste0(paste0(as.character('ASV'),id),' ('),Genus),')'))))
tax2_abx = tax2
rm(tax,tax2,plt_df,plt_df1,plt_df2)

```

## DATA NOT SHOWN: Can we predict whether a sample was sub-clinical or diarrheal from microbiota based on the change in microbiome?

```{r}

compare <- c(10,12) # TOD +/- symptoms
ps_for_class = change_in_microbiota_ps
ps_for_class <- subset_samples(ps_for_class, Clinical.type %in% compare)
ps_for_class_rel = ps_for_class#transform_sample_counts(ps_for_class, function(x) x/sum(x))

x <- otu_table(ps_for_class_rel)
y <- as.factor(sample_data(subset_samples(ps_for_class_rel, Clinical.type %in% compare))$Clinical.type)

rf = randomForest_res(x,y,10)
save(list = c("rf"), file = paste0(paste0("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/rf_change_" , format(Sys.time(), "%Y_%m_%d")),".RData"))     
roc_change = rf

rm(x,y,rf,ps_for_class_rel,ps_for_class,p)

```

# Save data for other scripts

```{r}

#"roc_list_mirpur","otu_TOD","","",
# 
# save(list = c("predetection_tax2","tod_tax2"), file = paste0(paste0("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P3output_plot_" , format(Sys.time(), "%Y_%m_%d")),".RData"))

save(list = c("chi_df","roc_TOD","roc_PD","plot_tax_4D","otu_PD","tax2_TOD","tax2_PD","twoASVs","two_WIDE","tax2_infect","roc_TOD","roc_TOD_abx","tax2_abx","roc_change","roc_allM_M","roc_allM","AUC_TOD","AUC_PD","AUC_allM","AUC_allM_M","AUC_TOD_abx"), file = paste0(paste0("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P3output_plot_" , format(Sys.time(), "%Y_%m_%d")),".RData"))

```

# Make figures
```{r}

#rmarkdown::render("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/code/R_markdown/P3_machine_learning_figures.Rmd")

```

### 2020-09-03 code/R_markdown/P3_machine_learning_figures.Rmd
added code and figures, did not add data or results - need to check consent re data availability

---
title: "Microbiome analysis - machine learning"
author: "Maureen A. Carey"
date: "09/01/2020"
output: html_document
---

# Load packages and data

```{r setup, include=FALSE, echo=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/")

# load packages
library(ggplot2); library(pROC); library(dplyr); library(tibble); library(reshape2)

# set constants
SEED_INT = 10; set.seed(SEED_INT); options(scipen=10000); theme_set(theme_bw())
colors = c("#1F78B4","#FF7F00","#666666", 'light gray')
names(colors) = c('diarrheal','subclinical','no crypto','both')
# blue = "#1F78B4" orange = "#FF7F00" gray = "#666666"

# load data
load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P3output_plot_2020_09_02.RData")

```

# Figure 4C: ROC Curve

```{r}

roc_list = list(roc_TOD,roc_PD)
names(roc_list) = c("TOD microbiota","pre-detection microbiota")
p = pROC::ggroc(roc_list, aes = "colour") + 
  theme_bw() +
  geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color="grey", linetype="dashed") +
  scale_color_manual(values = c("blue","red"), guide = F) + 
  theme(plot.margin=unit(c(.5,.5,.5,.5),"cm"),
        legend.title = element_blank()) +
  annotate("text", label = paste0("TOD AUC = ",round(AUC_TOD, digits = 2)), x = .25, y = .27, color = "blue") +
  annotate("text", label = paste0("PD AUC = ",round(AUC_PD, digits = 2)), x = .25, y = .13, color = "red") 
p
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/Fig4C_ROC_microbiome_mirpur.png",
       p,width = 3.2,height = 3,units = "in",dpi = 300)

```

# Figure 4D: Plot top 10 most important variables wth both microbiota classifiers

```{r}

plot_tax_4D %>%
  ggplot() +geom_bar(aes(x=var, y = value, fill = variable),stat = "identity", width=0.4, position = position_dodge(width=0.5)) +
  scale_fill_manual(values = c("red","blue")) +
  xlab(NULL) + ylab('Mean decrease in node impurity') + #facet_wrap(~tod_or_pre, scales = "free_x") +
  coord_flip() + 
  theme(legend.position = c(.76, 0.13), 
        axis.text.x = element_text(size = 7),
        legend.title=element_text(size=10), 
        legend.text=element_text(size=8))  + scale_x_discrete(limits = rev(levels(plot_tax_4D$var)))
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/Fig4D_imp_var_microbiome_mirpur.png",
       width = 4.5,height = 4,units = "in",dpi = 300)

```

## DATA NOT SHOWN: Plot important variables (raw values)

```{r}

# #plot PD
# reshape2::melt(otu_PD %>% rownames_to_column('sample')) %>%
#   group_by(variable) %>%
#   mutate(value_use = value/max(value)) %>%
#   ungroup() %>%
#   mutate(symptomatic_use = ifelse(symptomatic == '10','subclinical','diarrheal')) %>%
#   ggplot(aes(x = reorder(variable,value), y = value_use, group = sample, color = symptomatic_use)) +
#   geom_point(alpha = .5) + geom_line(alpha = .5) +
#   scale_color_manual(values = c(colors['diarrheal'],colors['subclinical'])) +
#   ylab('relative abundance') + xlab(NULL) + labs(color = "") + scale_y_sqrt()

#plot TOD
ggplot(tax2_TOD, aes(x = reorder(var,value), y = value_use, color = symptomatic_use)) +
  geom_boxplot(position = "dodge", outlier.shape = NA) +
  geom_point(alpha = .5, size = 1,
             position = position_jitterdodge(jitter.width = 0.3, jitter.height = 0, 
                                             seed = SEED_INT, dodge.width = 0.75)) +
  scale_color_manual(values = c(colors['diarrheal'],colors['subclinical'])) +
  ylab('relative abundance') + xlab(NULL) + labs(color = "")+ scale_y_sqrt() +
  theme(axis.text.x = element_text(angle = -45, hjust = 0, vjust = 0))
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/TOD_imp_var_boxplot_mirpur.png",
       width = 5,height = 3,units = "in",dpi = 300)

#plot PD
ggplot(tax2_PD, aes(x = reorder(var,value), y = value_use, color = symptomatic_use)) +
  geom_boxplot(position = "dodge", outlier.shape = NA) +
  geom_point(alpha = .5, size = 1,
             position = position_jitterdodge(jitter.width = 0.3, jitter.height = 0, 
                                             seed = SEED_INT, dodge.width = 0.75)) +
  scale_color_manual(values = c(colors['diarrheal'],colors['subclinical'])) +
  ylab('relative abundance') + xlab(NULL) + labs(color = "")+ scale_y_sqrt() +
  theme(axis.text.x = element_text(angle = -45, hjust = 0, vjust = 0))
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/PD_imp_var_boxplot_mirpur.png",
       width = 5,height = 3,units = "in",dpi = 300)

```

# Figure 4E - important variables

```{r}

ggplot(twoASVs, aes(x = var, y = value_use, color = symptomatic_use)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_point(alpha = 0.5, position = position_jitterdodge(jitter.width = 0.2, jitter.height = 0, dodge.width = 0.75, seed = SEED_INT)) + 
  facet_wrap(~time) +
  scale_color_manual(values= colors[1:2]) +
  scale_y_sqrt() + ylab('relative abundance') + xlab(NULL) + 
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), 
        legend.position = "bottom",
        legend.title = element_text(size = 9)) + guides(color=guide_legend(title="symptomatic?")) +
  geom_hline(yintercept = 0.2, linetype = 'dotted', alpha = .75)
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/Fig4E_mega.png",
       width = 3.2,height = 4,units = "in",dpi = 300)

ggplot(two_WIDE, aes(x = `time-of-detection`, y = `pre-detection`, color = `symptomatic?`)) +
  geom_jitter(alpha = 0.5, position = position_jitter(width = 0.01, height = 0.01, seed = SEED_INT)) +
  scale_color_manual(values = colors[1:2]) + scale_x_sqrt() + scale_y_sqrt()+
  theme(panel.grid.minor = element_blank(),
        legend.position="bottom")

ggplot(two_WIDE, aes(x = `time-of-detection`, y = `pre-detection`, color = `symptomatic?`)) +
  geom_point(alpha = 0.5) +
  scale_color_manual(values = colors[1:2]) + scale_x_sqrt() + scale_y_sqrt()+
  theme(panel.grid.minor = element_blank(),
        legend.position="bottom")

```

## DATA NOT SHOWN: Imp var for first v later infection

```{r}

data_wide = tidyr::spread(chi_df, symptomatic, Freq)
rownames(data_wide) = data_wide$n_crypto; data_wide$n_crypto = NULL
res = chisq.test(data_wide) 
chi_df %>% ggplot(aes(x = n_crypto, y = Freq, fill = symptomatic)) + geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = colors) + ylab('episodes') + xlab('first or subsequent infection?') +
  annotate(geom = "text", x = 2, y = 45, label = paste0("Pearson's Chi-\nsquared test\np-value = ",round(res$p.value,3))) + ylim(0,52) + theme(legend.position = "bottom")

tax2_infect %>%
  ggplot() +geom_bar(aes(x=reorder(var,value), y = value, fill = model),stat = "identity", width=0.4, position = position_dodge(width=0.5)) +
  scale_fill_manual(values = c("red","blue")) +
  xlab(NULL) + ylab('Mean decrease in node impurity') + #facet_wrap(~tod_or_pre, scales = "free_x") +
  coord_flip()
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/imp_var_Infect#_mirpur.png",
              width = 7,height = 3,units = "in",dpi = 300)

```

## DATA NOT SHOWN: metadata doesn't help classifier

```{r}

roc_list = list(roc_allM_M,roc_allM)
names(roc_list) = c("all microbiota & metadata","all microbiota")
p = pROC::ggroc(roc_list, aes = "colour") +
  theme_bw() +
  geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color="grey", linetype="dashed") +
  scale_color_manual(values = c("grey","red")) +
  theme(plot.margin=unit(c(.5,.5,.5,.5),"cm"),
        legend.title = element_blank()) +
  annotate("text", label = paste0("AUC = ",round(AUC_allM, digits = 2)), x = .2, y = .13, color = "red") +
  annotate("text", label = paste0("AUC = ",round(AUC_allM_M, digits = 2)), x = .2, y = .2, color = "grey")
p

```

# Supplemental Figure 5 - ROC and Important variables, Adding in Abx to model

```{r}

load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/rf_TOD_abx_2020_09_02.RData")
AUC_TOD_abx = rf[[3]]
rm(rf)
roc_list = list(roc_TOD,roc_TOD_abx)
names(roc_list) = c("TOD microbiota","TOD microbiota &\nantibiotic history")
p = pROC::ggroc(roc_list, aes = "colour") + 
  theme_bw() +
  geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color="grey", linetype="dashed") +
  scale_color_manual(values = c("blue","grey")) + 
  theme(plot.margin=unit(c(.5,.5,.5,.5),"cm"),
        legend.title = element_blank()) +
  annotate("text", label = paste0("AUC = ",round(AUC_TOD, digits = 2)), x = .2, y = .27, color = "blue") +
  annotate("text", label = paste0("AUC = ",round(AUC_TOD_abx, digits = 2)), x = .2, y = .2, color = "grey") 
p
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/SFig5A_ROC_TOD_abx_mirpur.png",
       p,width = 5,height = 3,units = "in",dpi = 300)

tax2_abx %>%
  ggplot() +geom_bar(aes(x=reorder(var,value), y = value, fill = model),stat = "identity", width=0.4, position = position_dodge(width=0.5)) +
  scale_fill_manual(values = c("black","grey")) +
  xlab(NULL) + ylab('Mean decrease in node impurity') + #facet_wrap(~tod_or_pre, scales = "free_x") +
  coord_flip()
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/SFig5B_imp_var_abx_mirpur.png",
       width = 7,height = 3,units = "in",dpi = 300)

```

## DATA NOT SHOWN: Can we predict whether a sample was sub-clinical or diarrheal from microbiota based on the change in microbiome?

```{r}

p = pROC::ggroc(roc_change[[2]]) + 
  theme_bw() +
  geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color="grey", linetype="dashed") + 
  theme(plot.margin=unit(c(.5,.5,.5,.5),"cm")) + guides(color = F) + ggtitle(roc_change[[3]])
p

```

### 2020-09-03 code/R_markdown/P4_univariate_stats.Rmd
added code and figures, did not add data or results - need to check consent re data availability

---
title: "Microbiome analysis - univariate statistics"
author: "Maureen A. Carey"
date: "09/01/2020"
output: html_document
---

# Load packages and data

```{r setup, include=FALSE, echo=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/")

# load packages
library(phyloseq); library(reshape2); library(tibble); library(rmarkdown)
library(dplyr); library(tidyr); library(DESeq2); library(ashr); library(ggrepel)

# set constants
SEED_INT = 10; set.seed(SEED_INT); options(scipen=10000)
colors = c("#1F78B4","#FF7F00","#666666", 'light gray')
names(colors) = c('diarrheal','subclinical','no crypto','both')
# blue = "#1F78B4" orange = "#FF7F00" gray = "#666666"

# load data
load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P1output_2020_09_02.RData")
load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P3output_p1_2020_09_02.RData")

```

# Univariate statistics

```{r}

# top10_TOD
# top10_PD

# subset to PD v TOD
ps_rare_diff_temp_PD = subset_samples(for_stats, pre_detection == T)
ps_rare_diff_temp_TOD = subset_samples(for_stats, pre_detection == F)

# remove very rare ASVs
nonzero_count = apply(otu_table(ps_rare_diff_temp_PD), 1, function(c)sum(c!= 0))
ps_rare_diff_temp_PD = prune_samples(nonzero_count > 5, ps_rare_diff_temp_PD)
nonzero_count = apply(otu_table(ps_rare_diff_temp_TOD), 1, function(c)sum(c!= 0))
ps_rare_diff_temp_TOD = prune_samples(nonzero_count > 5, ps_rare_diff_temp_TOD)

# transpose and replace zeros
predetection_count = t(otu_table(ps_rare_diff_temp_PD))
predetection_count[predetection_count == 0] = 1
TOD_count = t(otu_table(ps_rare_diff_temp_TOD))
TOD_count[TOD_count == 0] = 1

# make sample data more managable
predetection_sample_data = as(sample_data(ps_rare_diff_temp_PD),"data.frame") %>% 
  select(Child.ID, symptomatic, pre_detection)
TOD_sample_data = as(sample_data(ps_rare_diff_temp_TOD),"data.frame") %>% 
  select(Child.ID, symptomatic, pre_detection)

```

# DESeq2

```{r}

# make DESeq obj
dds_PD <- DESeqDataSetFromMatrix(countData = as.data.frame(predetection_count),
                              colData = predetection_sample_data,
                              design= ~symptomatic)
dds_TOD <- DESeqDataSetFromMatrix(countData = as.data.frame(TOD_count),
                                 colData = TOD_sample_data,
                                 design= ~symptomatic)
dds_PD = DESeq(dds_PD)
dds_TOD = DESeq(dds_TOD)

# shrinkage
# because we are interested in treated vs untreated, we set 'coef=2'
resNorm_PD <- lfcShrink(dds_PD, coef=2, type="ashr") # resultsNames(dds_proc)[2]
resNorm_TOD <- lfcShrink(dds_TOD, coef=2, type="ashr") # resultsNames(dds_proc)[2]

#resOrdered <- resNorm_PD[order(resNorm_PD$pvalue),]
summary(resNorm_PD)
summary(resNorm_TOD)

```

# Prep predetection volcano plot

```{r}

# Create the volcano plot - PREP FOR PREDETECTION
resNorm_PD_df = as.data.frame(resNorm_PD) %>% 
  rownames_to_column('ASV') %>%
  mutate(col = ifelse(padj<0.05,1,0),
         lab = ifelse(log2FoldChange>1.5 | abs(log2FoldChange) > 2,1,0),
         sig = -log10(padj))
sum(is.infinite(resNorm_PD_df$sig))
sum(is.na(resNorm_PD_df$sig))
resNorm_PD_df = resNorm_PD_df %>% mutate(padj = ifelse(is.na(padj),pvalue,padj),
                                         sig = ifelse(is.na(sig),0,-log10(pvalue)),
                                         col = ifelse(is.na(col),0,col),
                                         lab = ifelse(is.na(lab),0,lab))
#resNorm_PD[is.infinite(resNorm_PD$sig),"sig"] <- 350
sum(is.na(resNorm_PD_df$pvalue))
sum(is.infinite(resNorm_PD_df$pvalue))
#genes.to.plot <- !is.na(resNorm_PD$pvalue)

# Label ASVs
tax = tax_table(ps_rare_diff_temp_PD)
tax = tax[rownames(tax) %in% resNorm_PD_df$ASV,]
tax = merge(tax,resNorm_PD_df, by.x= 'row.names', by.y = "ASV", all.x = F, all.y = T)
tax$Genus = as.character(tax$Genus)
tax2 = tax %>% 
  transform(id=as.numeric(factor(Row.names))) %>%
  mutate(var = ifelse(is.na(Family),
                      as.character(Row.names),
                      ifelse(is.na(Genus),paste0('[f] ',Family),Genus)))

tax2 = merge(merge(tax2, as.data.frame(top10_PD)%>% rownames_to_column('Row.names'), by = "Row.names", all = T),
             as.data.frame(top10_TOD)%>% rownames_to_column('Row.names'), by = "Row.names", all = T)
tax2 = tax2 %>% 
  mutate(col2 = ifelse(col == 0, 'non sig',ifelse(!is.na(top10_TOD) & !is.na(top10_PD),'sig but dif color','sig')),
         size2 = ifelse(!is.na(top10_PD),1,0))

predetection_tax2 = tax2

```

# Prep time-of-detection volcano plot

```{r}

# Create the volcano plot - PREP FOR TIME OF DETECTION
resNorm_PD_df = as.data.frame(resNorm_TOD) %>% 
  rownames_to_column('ASV') %>%
  mutate(col = ifelse(padj<0.05,1,0),
         lab = ifelse(log2FoldChange>2.5 | abs(log2FoldChange) > 2.5,1,0),
         sig = -log10(padj))
sum(is.infinite(resNorm_PD_df$sig))
sum(is.na(resNorm_PD_df$sig))
resNorm_PD_df = resNorm_PD_df %>% mutate(padj = ifelse(is.na(padj),pvalue,padj),
                                         sig = ifelse(is.na(sig),0,-log10(pvalue)),
                                         col = ifelse(is.na(col),0,col),
                                         lab = ifelse(is.na(lab),0,lab))
#resNorm_PD[is.infinite(resNorm_PD$sig),"sig"] <- 350
sum(is.na(resNorm_PD_df$pvalue))
sum(is.infinite(resNorm_PD_df$pvalue))
#genes.to.plot <- !is.na(resNorm_PD$pvalue)

# Label ASVs
tax = tax_table(ps_rare_diff_temp_TOD)
tax = tax[rownames(tax) %in% resNorm_PD_df$ASV,]
tax = merge(tax,resNorm_PD_df, by.x= 'row.names', by.y = "ASV", all.x = F, all.y = T)
tax$Genus = as.character(tax$Genus)
tax2 = tax %>% 
  transform(id=as.numeric(factor(Row.names))) %>%
  mutate(var = ifelse(is.na(Family),
                      as.character(Row.names),
                      ifelse(is.na(Genus),paste0('[f] ',Family),Genus)))
tax2 = merge(merge(tax2, as.data.frame(top10_TOD)%>% rownames_to_column('Row.names'), by = "Row.names", all = T),
             as.data.frame(top10_PD)%>% rownames_to_column('Row.names'), by = "Row.names", all = T)
#tax2 = tax2 %>% mutate(col2 = ifelse(!is.na(top10_TOD > 0), ifelse(!is.na(top10_PD > 0), 'imp (both)', 'imp, TOD only'), ifelse(col == 0, 'insig', 'sig')))
tax2 = tax2 %>% 
  mutate(col2 = ifelse(col == 0, 'non sig',ifelse(!is.na(top10_TOD) & !is.na(top10_PD),'sig but dif color','sig')),
         size2 = ifelse(!is.na(top10_TOD),1,0))

tod_tax2 = tax2

```

# Save data for other scripts

```{r}

save(list = c("predetection_tax2","tod_tax2"), file = paste0(paste0("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P4output_plot_" , format(Sys.time(), "%Y_%m_%d")),".RData"))

```

# Make figures
```{r}

#rmarkdown::render("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/code/R_markdown/P4_univariate_stats_figures.Rmd")

```

### 2020-09-03 code/R_markdown/P4_univariate_stats_figures.Rmd
added code and figures, did not add data or results - need to check consent re data availability

---
title: "Microbiome analysis - univariate statistics"
author: "Maureen A. Carey"
date: "09/01/2020"
output: html_document
---

# Load packages and data

```{r setup, include=FALSE, echo=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/")

# load packages
library(ggplot2); library(ggrepel); library(dplyr)

# set constants
SEED_INT = 10; set.seed(SEED_INT); options(scipen=10000); theme_set(theme_bw())
colors = c("#1F78B4","#FF7F00","#666666", 'light gray')
names(colors) = c('diarrheal','subclinical','no crypto','both')
# blue = "#1F78B4" orange = "#FF7F00" gray = "#666666"

# load data
load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P4output_plot_2020_09_02.RData")

```

# Plot predetection volcano plot

```{r}

ggplot(predetection_tax2 %>% mutate(timepoint = 'pre-detection')) + 
  geom_point(aes(x = log2FoldChange, y = sig, color = as.factor(col2), size = as.factor(size2) ), alpha = .2)+
  xlab("Effect size: log2(fold change)") + ylab("-log10(adjusted p-value)")+
  scale_color_manual(values = c('black',"red","purple"),  guide = F) +
  scale_size_manual(values = c(1,3),  guide = F) + facet_wrap(~timepoint)
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/Fig4B_univar_PD.png",
       width = 4,height = 4,units = "in",dpi = 300)

ggplot(predetection_tax2) + 
  geom_point(aes(x = log2FoldChange, y = sig, color = as.factor(col2), size = as.factor(size2) ), alpha = .2)+
  xlab("Effect size: log2(fold change)") + ylab("-log10(adjusted p-value)")+
  scale_color_manual(values = c('black',"red","purple"),  guide = F) +
  scale_size_manual(values = c(1,3),  guide = F)  + 
  geom_label_repel(data = predetection_tax2 %>% filter(!(is.na(top10_PD))), #filter(lab == 1), 
                   aes(x = log2FoldChange, y = sig, color = as.factor(col2), label = var)) + xlim(-5,4)
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/FOR_REF_Fig4B_univar_PD.png",
       width = 5,height = 5,units = "in",dpi = 300)

```

# Plot time-of-detection volcano plot

```{r}

ggplot(tod_tax2 %>% mutate(timepoint = 'time-of-detection')) + 
  geom_point(aes(x = log2FoldChange, y = sig, color = as.factor(col2), size = as.factor(size2) ), alpha = .2)+
  xlab("Effect size: log2(fold change)") + ylab("-log10(adjusted p-value)")+
  scale_color_manual(values = c('black',"blue","purple"),  guide = F) +
  scale_size_manual(values = c(1,3),  guide = F)+ facet_wrap(~timepoint)
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/Fig4B_univar_TOD.png",
       width = 4,height = 4,units = "in",dpi = 300)

ggplot(tod_tax2) + 
  geom_point(aes(x = log2FoldChange, y = sig, color = as.factor(col2), size = as.factor(size2) ), alpha = .2)+
  xlab("Effect size: log2(fold change)") + ylab("-log10(adjusted p-value)")+
  scale_color_manual(values = c('black',"blue","purple"),  guide = F) +
  scale_size_manual(values = c(1,3),  guide = F)  + 
  geom_label_repel(data = tod_tax2 %>% filter(!(is.na(top10_TOD))), #filter(lab == 1), 
                   aes(x = log2FoldChange, y = sig, color = as.factor(col2), label = var)) + xlim(-5,4)
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/FOR_REF_Fig4B_univar_TOD.png",
       width = 5,height = 5,units = "in",dpi = 300)

```

### 2020-09-03 code/R_markdown/P5_mirzapur_analysis.Rmd
added code and figures, did not add data or results - need to check consent re data availability

---
title: "Microbiome analysis - Mirzapur follow-up"
author: "Maureen A. Carey"
date: "09/01/2020"
output: html_document
---

# Load packages and data

```{r setup, include=FALSE, echo=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/")

# load packages
library(phyloseq); library(reshape2)
library(rmarkdown); library(dplyr); library(tidyr); library(tibble)

# load data
load("/Users/maureencarey/local_documents/work/crypto_microbiome/data_cannot_be_shared/data.RData")
rm(list= ls()[! (ls() %in% c('left_study_ids'))])
load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P3output_p2_2020_09_02.RData")
load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P3output_p3_2020_09_02.RData")
ps_mirzapur = readRDS("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/ps_2020_08_12.rds")
load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P3output_p1_2020_09_02.RData") # load save_df1 and save_df2

# set constants
SEED_INT = 10; set.seed(SEED_INT); options(scipen=10000)
colors = c("#1F78B4","#FF7F00","#666666", 'light gray')
names(colors) = c('diarrheal','subclinical','no crypto','both')
# blue = "#1F78B4" orange = "#FF7F00" gray = "#666666"

```

# Check Megasphaera in Mirzapur

```{r}

ps_mirzapur = subset_samples(ps_mirzapur, Site == 2) # MIRZAPUR ONLY

# Quality control - remove mitochondrial and chloroplast reads
og_taxa = data.frame(tax_table(ps_mirzapur))

# first remove non-bacteria
keep = rownames(og_taxa[og_taxa$Kingdom %in% c("Bacteria","Archaea"),])
ps_mirzapur = prune_taxa(keep,ps_mirzapur)
keep = rownames(og_taxa[og_taxa$Family != "Mitochondria",])
ps_mirzapur = prune_taxa(keep,ps_mirzapur)
keep = rownames(og_taxa[og_taxa$Class != "Chloroplast",])
ps_mirzapur = prune_taxa(keep,ps_mirzapur)

# remove if error in any reporting re Crypto infection
keep1 = ifelse(sample_data(ps_mirzapur)$Crypto..Any. == '1' & as.numeric(sample_data(ps_mirzapur)$Clinical.type) < 3, 0, 1)
ps_mirzapur = prune_samples(as.logical(keep1),ps_mirzapur)
keep2 = ifelse(sample_data(ps_mirzapur)$Crypto..Any. == '999' & as.numeric(sample_data(ps_mirzapur)$Clinical.type) > 3, 0, 1)
ps_mirzapur = prune_samples(as.logical(keep2),ps_mirzapur)
keep3 = ifelse(sample_data(ps_mirzapur)$Crypto..Any. == '888' & as.numeric(sample_data(ps_mirzapur)$Clinical.type) > 3, 0, 1)
ps_mirzapur = prune_samples(as.logical(keep3),ps_mirzapur)
rm(keep, keep1, keep2, keep3)

# remove children who were lost to follow up
ps_mirzapur = subset_samples(ps_mirzapur, !(Child.ID %in% left_study_ids))

# Rarify reads in all samples
rare_min = 10000
# remove samples with fewer counts than the minimum
ps_mirzapur_rare = prune_samples(sample_sums(ps_mirzapur) >=rare_min,ps_mirzapur) #8 removed
# remove samples with NA age
ps_mirzapur_rare = prune_samples(!is.na(sample_data(ps_mirzapur_rare)[,"Age.in.days"])[,1],ps_mirzapur_rare) # none removed
ps_mirzapur_rare = rarefy_even_depth(ps_mirzapur_rare,sample.size=rare_min,rngseed=SEED_INT) # OTUs no longer in the dataset will be removed

# number of Megaspaheara ASVs
temp = as.data.frame(tax_table(ps_mirzapur_rare))
dim(temp[temp$Genus == 'Megasphaera',])
# temp2 = as.data.frame(tax_table(ps_rare))
# dim(temp2[temp2$Genus == 'Megasphaera',])

```

# Extract ASV that important from classifiers

```{r}

otu_for_comp_TOD = as.data.frame(otu_table(ps_mirzapur_rare))[,save_df1$variable]
otu_for_comp_PD = as.data.frame(otu_table(ps_mirzapur_rare))[,save_df1$variable]

# get TOD values
plt_df = melt(otu_for_comp_TOD %>% rownames_to_column('sample')) %>%
  group_by(variable) %>%
  mutate(value_use = value/max(value)) %>%
  ungroup() 
tax = tax_table(ps_mirzapur_rare)
tax = tax[rownames(tax) %in% plt_df$variable,]
tax = merge(tax,plt_df, by.x= 'row.names', by.y = "variable", all.x = F, all.y = T)
tax$Genus = as.character(tax$Genus)
tax2 = tax %>% 
  transform(id=as.numeric(factor(Row.names))) %>%
  mutate(var = ifelse(is.na(Family),
                      as.character(Row.names),
                      ifelse(is.na(Genus), paste0(paste0(paste0(paste0(as.character('ASV'),id),' ([f]'),Family),')'),
                             paste0(paste0(paste0(paste0(as.character('ASV'),id),' ('),Genus),')'))))
twoASVs_TOD = tax2 %>% filter(var == 'ASV10 (Megasphaera)')

# get PD values
plt_df = melt(otu_for_comp_PD %>% rownames_to_column('sample')) %>%
  group_by(variable) %>%
  mutate(value_use = value/max(value)) %>%
  ungroup() 
tax = tax_table(ps_mirzapur_rare)
tax = tax[rownames(tax) %in% plt_df$variable,]
tax = merge(tax,plt_df, by.x= 'row.names', by.y = "variable", all.x = F, all.y = T)
tax$Genus = as.character(tax$Genus)
tax2 = tax %>% 
  transform(id=as.numeric(factor(Row.names))) %>%
  mutate(var = ifelse(is.na(Family),
                      as.character(Row.names),
                      ifelse(is.na(Genus), paste0(paste0(paste0(paste0(as.character('ASV'),id),' ([f]'),Family),')'),
                             paste0(paste0(paste0(paste0(as.character('ASV'),id),' ('),Genus),')'))))
PD_twoASVs = tax2 %>% filter(var == 'ASV10 (Megasphaera)')

twoASVs = rbind(twoASVs_TOD %>% mutate(time = 'TOD'), PD_twoASVs %>% mutate(time = 'PD'))
twoASVs_wide = tidyr::spread(twoASVs %>% select(value_use,time,sample, Genus),
                             key = Genus, value = value_use) %>% 
  mutate(time = as.factor(time))
levels(twoASVs_wide$time)[levels(twoASVs_wide$time)=="PD"] =  "pre-detection"
levels(twoASVs_wide$time)[levels(twoASVs_wide$time)=="TOD"] =  "time-of-detection"

```

# Prep Plot - Fig 4F

```{r}

count_df_PD_mirzapur = twoASVs_wide %>% filter(time == 'pre-detection')
count_df_TOD_mirzapur = twoASVs_wide %>% filter(time != 'pre-detection')
df = data.frame(
  Site = c('Mirzapur','Mirpur'),
  Megasphaera = c(round(sum(count_df_PD_mirzapur$Megasphaera>0)/length(count_df_PD_mirzapur$Megasphaera),4)*100, round(sum(count_df_PD_mirpur$Megasphaera>0)/length(count_df_PD_mirpur$Megasphaera),4)*100))
df = melt(df)
# ggplot(df) + 
#   geom_bar(aes(x = Site, y = value, fill = Site), stat = 'identity', position = "dodge") + 
#   scale_fill_manual(values = c('black','grey'), guide = F) +
#   ylab('Percent of samples with reads') + xlab(NULL)

df_mirz = melt(rbind(count_df_PD_mirzapur, count_df_TOD_mirzapur))  %>% mutate(Site = 'Mirzapur')
df_mir = melt(rbind(count_df_PD_mirpur %>% select(-symptomatic_use,-Collinsella,-`symptomatic?`) %>% 
                      mutate(time = 'pre-detection'), 
                    count_df_TOD_mirpur %>% select(-symptomatic_use,-Parabacteroides,-`symptomatic?`) %>% 
                      mutate(time = 'time-of-detection'))) %>% mutate(Site = 'Mirpur')
df = rbind(df_mirz,df_mir)

```

# Save data for other scripts

```{r}

save(list = c("df","ps_mirzapur_rare"), file = paste0(paste0("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P5output_plot_" , format(Sys.time(), "%Y_%m_%d")),".RData"))

```

# Statistics

```{r}

t.test(df$value[df$Site == 'Mirpur'], df$value[df$Site == 'Mirzapur'])
df = data.frame(Site = c('Mirzapur','Mirpur'),
  Megasphaera = c(sum(count_df_PD_mirzapur$Megasphaera>0.2),sum(count_df_PD_mirpur$Megasphaera>0.2)),
  Total = c(length(count_df_PD_mirzapur$Megasphaera),length(count_df_PD_mirpur$Megasphaera))) %>% 
  mutate(NO_mega = Total - Megasphaera) %>% select(-Total)
rownames(df) = df$Site; df$Site = NULL
chisq.test(df)

```

# Make figures

```{r}

#rmarkdown::render("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/code/R_markdown/P5_mirzapur_analysis_figures.Rmd")

```

### 2020-09-03 code/R_markdown/P5_mirzapur_analysis_figures.Rmd
added code and figures, did not add data or results - need to check consent re data availability

---
title: "Microbiome analysis - Mirzapur follow-up"
author: "Maureen A. Carey"
date: "09/01/2020"
output: html_document
---

# Load packages and data

```{r setup, include=FALSE, echo=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/")

# load packages
library(ggplot2); library(phyloseq)

# load data
load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P5output_plot_2020_09_02.RData")

# set constants
SEED_INT = 10; set.seed(SEED_INT); options(scipen=10000); theme_set(theme_bw())
colors = c("#1F78B4","#FF7F00","#666666", 'light gray')
names(colors) = c('diarrheal','subclinical','no crypto','both')
# blue = "#1F78B4" orange = "#FF7F00" gray = "#666666"

```

# Plot - Fig 4F

```{r}

ggplot(df, aes(x = Site, y = value, color = Site)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_point(alpha = 0.5, position = position_jitter(width = 0.3, height = 0, seed = SEED_INT)) + 
  facet_wrap(~time) +
  scale_y_sqrt() +
  scale_color_manual(values = c('black','darkgray'), guide = F) +
  geom_hline(yintercept = 0.2, linetype = 'dotted', alpha = .75) +
  ylab('relative abundance')
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/Fig4F_mega_box_Mirzapur.png",
       width = 3.2,height = 3.5,units = "in",dpi = 300)

```

## Table 1-like data for methods

```{r}

length(unique(sample_data(ps_mirzapur_rare)$Child.ID))
length((sample_data(ps_mirzapur_rare)$Child.ID))
table(sample_data(ps_mirzapur_rare)$Clinical.type)
min(sample_data(ps_mirzapur_rare)$Date.of.collection)
max(sample_data(ps_mirzapur_rare)$Date.of.collection)

```

### 2020-09-03 code/on_rivanna/R_scripts/dada2_run1.R
added code for running dada2 on rivanna

@@ -1,0 +1,136 @@
+#install.packages("BiocManager", repos='http://cran.us.r-project.org') # WORKS
+#BiocManager::install("ShortRead", dependencies = TRUE)
+
+library("dada2")
+library("ShortRead")
+library("phyloseq")
+library("ggplot2")
+sink('/project/tumi/carey/crypto_micro/dada2_R_outfile_r1.txt')
+packageVersion('dada2')
+
+# get raw sample reads (these are pretrimmed, removing adapters via bbtools and primers via cutadapt)
+path = "/scratch/mac9jc/crypto_micro/trimmed_seqs/run1"
+#path <- "/project/tumi/carey/crypto_microbiome/trimmed_seqs/run1"
+fns <- list.files(path)
+fns = grep("_PT2.fq.gz",fns,value=TRUE)
+
+# remove reads that couldn't be assigned to samples in the demultiplexing process
+remove = grep("Undetermined_",fns,value=TRUE)
+fns = fns[!fns %in% remove]
+
+### Load forward and reverse reads
+fastqs <- fns[grepl(".fq.gz$", fns)]
+fastqs <- sort(fastqs) # Sort ensures forward/reverse reads are in same order
+fnFs <- fastqs[grepl("_1_PT2", fastqs)] # Just the forward read files
+fnRs <- fastqs[grepl("_2_PT2", fastqs)] # Just the reverse read files
+# Get sample names from the first part of the forward read filenames
+sample.names <- sapply(strsplit(fnFs, "_"), `[`, 1)
+# Fully specify the path for the fnFs and fnRs
+fnFs <- file.path(path, fnFs)
+fnRs <- file.path(path, fnRs)
+
+# remove samples if they have fewer than 10,000 reads prior to filtration
+# this is VERY SLOW, so don't rerun everytime
+remove_list = list()
+for (i in 1:length(sample.names)) {
+  f_seq = getSequences(fnFs[i]); r_seq = getSequences(fnRs[i])
+  if (length(f_seq) < 10000 | length(r_seq) < 10000 ) {
+   print(sample.names[i])
+   remove_list = list(remove_list, i) }}
+remove_list = unlist(remove_list)
+# nto sure if this shortcut is working
+#remove_list = which(c("130491008","130771018","131361012","neg1","neg2","neg3") %in% sample.names)
+print(remove_list)
[…]

### 2020-09-03 code/on_rivanna/R_scripts/dada2_run2.R

@@ -1,0 +1,130 @@
+#install.packages("BiocManager", repos='http://cran.us.r-project.org') # WORKS
+#BiocManager::install("ShortRead", dependencies = TRUE)
+
+library("dada2")
+library("ShortRead")
+library("phyloseq")
+library("ggplot2")
+sink('/project/tumi/carey/crypto_micro/dada2_R_outfile_r2.txt')
+packageVersion('dada2')
+
+# get raw sample reads (these are pretrimmed, removing adapters via bbtools and primers via cutadapt)
+path2 = "/scratch/mac9jc/crypto_micro/trimmed_seqs/run2"
+#path2 <- "/project/tumi/carey/crypto_microbiome/trimmed_seqs/run2"
+fns2 <- list.files(path2)
+fns2 = grep("_PT2.fq.gz",fns2,value=TRUE)
+
+# remove reads that couldn't be assigned to samples in the demultiplexing process
+remove = grep("Undetermined_",fns2,value=TRUE)
+fns2 = fns2[!fns2 %in% remove]
+
+### Load forward and reverse reads for datapool 2
+fastqs2 <- fns2[grepl(".fq.gz$", fns2)]
+fastqs2 <- sort(fastqs2) # Sort ensures forward/reverse reads are in same order
+fnFs2 <- fastqs2[grepl("_1_PT2", fastqs2)] # Just the forward read files
+fnRs2 <- fastqs2[grepl("__2_PT2", fastqs2)] # Just the reverse read files
+# Get sample names from the first part of the forward read filenames
+sample.names2 <- sapply(strsplit(fnFs2, "_"), `[`, 1)
+# Fully specify the path for the fnFs and fnRs
+fnFs2 <- file.path(path2, fnFs2)
+fnRs2 <- file.path(path2, fnRs2)
+
+# remove samples if they have fewer than 10,000 reads prior to filtration
+# this is VERY SLOW, so don't rerun everytime
+remove_list = list()
+for (i in 1:length(sample.names2)) {
+  f_seq = getSequences(fnFs2[i]); r_seq = getSequences(fnRs2[i])
+  if (length(f_seq) < 10000 | length(r_seq) < 10000 ) {
+   remove_list = list(remove_list, i) }}
+remove_list = unlist(remove_list)
+# replace with run 2 results and check if this line works in the other run
[…]

### 2020-09-03 code/on_rivanna/R_scripts/get_input_reads.R

@@ -1,0 +1,51 @@
+library("dada2")
+library("ShortRead")
+sink('/project/tumi/carey/crypto_micro/R_outfile_get_reads.txt')
+
+# get raw sample reads (these are pretrimmed, removing adapters via bbtools and primers via cutadapt)
+path1 = "/project/tumi/medlock/crypto_microbiome/data/sequence_backup/DataPool1/fastqs"
+fns1 <- list.files(path1)
+path2 = "/project/tumi/medlock/crypto_microbiome/data/sequence_backup/DataPool2/fastqs"
+fns2 <- list.files(path2)
+fns1 = grep(".fastq.gz",fns1,value=TRUE)
+fns2 = grep(".fastq.gz",fns2,value=TRUE)
+
+# remove reads that couldn't be assigned to samples in the demultiplexing process
+remove = grep("Undetermined_",fns1,value=TRUE)
+fns1 = fns1[!fns1 %in% remove]
+remove = grep("Undetermined_",fns2,value=TRUE)
+fns2 = fns2[!fns2 %in% remove]
+
+### Load forward and reverse reads
+fastqs1 <- fns1[grepl(".fastq.gz$", fns1)]
+fastqs1 <- sort(fastqs1) # Sort ensures forward/reverse reads are in same order
+fnFs1 <- fastqs1[grepl("_R1_001", fastqs1)] # Just the forward read files
+fnRs1 <- fastqs1[grepl("_R2_001", fastqs1)] # Just the reverse read files
+# Get sample names from the first part of the forward read filenames
+sample.names1 <- sapply(strsplit(fnFs1, "_"), `[`, 1)
+# Fully specify the path for the fnFs and fnRs
+fnFs1 <- file.path(path1, fnFs1)
+fnRs1 <- file.path(path1, fnRs1)
+
+fastqs2 <- fns2[grepl(".fastq.gz$", fns2)]
+fastqs2 <- sort(fastqs2) # Sort ensures forward/reverse reads are in same order
+fnFs2 <- fastqs2[grepl("_R1_001", fastqs2)] # Just the forward read files
+fnRs2 <- fastqs2[grepl("_R2_001", fastqs2)] # Just the reverse read files
+# Get sample names from the first part of the forward read filenames
+sample.names2 <- sapply(strsplit(fnFs2, "_"), `[`, 1)
+# Fully specify the path for the fnFs and fnRs
[…]

### 2020-09-03 code/on_rivanna/R_scripts/merge_dada2.R

@@ -1,0 +1,106 @@
+#install.packages("BiocManager", repos='http://cran.us.r-project.org') # WORKS
+#BiocManager::install("ShortRead", dependencies = TRUE)
+
+library("dada2")
+library("ShortRead")
+library("phyloseq")
+sink('/project/tumi/carey/crypto_micro/dada2_R_outfile_merge.txt')
+packageVersion('dada2')
+
+# load both sequence tables, merge
+st1 <- readRDS("/project/tumi/carey/crypto_microbiome/results/run1_withtrimming_seqtab.rds")
+st2 <- readRDS("/project/tumi/carey/crypto_microbiome/results/run2_withtrimming_seqtab.rds")
+
+# append _run1 and _run2 to each sample name
+rownames(st1) <- paste0(rownames(st1),'_run1')
+rownames(st2) <- paste0(rownames(st2),'_run2')
+
+st <- mergeSequenceTables(st1, st2)
+# remove sequences far outside the desired length
+# st <- st[,nchar(colnames(st)) %in% seq(272,282)]
+
+# collapse nomismatch sequences, which are identical other than being of differing lengths
+st = collapseNoMismatch(st)
+
+print('start assigning taxonomy')
+# assign taxonomy
+taxa <- assignTaxonomy(st,"/project/tumi/carey/crypto_microbiome/rdp_train_set_16.fa")
+
+### transfer to phyloseq
+#prep
+samples.out <- rownames(st)
+# load and format metadata
+meta <- read.table("/project/tumi/carey/crypto_microbiome/meta_16_12_5.csv",header=TRUE, sep='\t')
+# append _run1 and _run2 to sample id to match sequence table rows
+a = paste0(meta$sample.id,'_run')
+meta$sample.id <- paste0(a,meta$Library)
+rownames(meta) <- meta$sample.id
+
+#match rownames for merging
+otu_table_use = otu_table(st, taxa_are_rows = FALSE)
+rownames(otu_table_use)[rownames(otu_table_use) == 'HM-782D-Pos1_run1'] = 'pos1_run1'
+rownames(otu_table_use)[rownames(otu_table_use) == 'HM-782D-Pos2_run1'] = 'pos2_run1'
+rownames(otu_table_use)[rownames(otu_table_use) == 'HM-782D-Pos3_run1'] = 'pos3_run1'
[…]

### 2020-09-08 code/R_markdown/P1_analysis_prep_figures.Rmd
added species random forest and updated figure 1

@@
 ```
+
+Species
+```{r}
+
+#all
+ifelse(sample_data(subset_samples(ps_rare_diff, pre_detection == F))$LIB13.C..hominis<40,1,0) %>% table()
+ifelse(sample_data(subset_samples(ps_rare_diff, pre_detection == F))$C..mel.y<40,1,0) %>% table()
+ifelse(sample_data(subset_samples(ps_rare_diff, pre_detection == F))$Pan_crypto<40,1,0) %>% table()
+
+# diarrheal
+ifelse(sample_data(subset_samples(ps_rare_diff, pre_detection == F & symptomatic == T))$LIB13.C..hominis<40,1,0) %>% table()
+ifelse(sample_data(subset_samples(ps_rare_diff, pre_detection == F & symptomatic == T))$C..mel.y<40,1,0) %>% table()
+ifelse(sample_data(subset_samples(ps_rare_diff, pre_detection == F & symptomatic == T))$Pan_crypto<40,1,0) %>% table()
+
+# subclin
+ifelse(sample_data(subset_samples(ps_rare_diff, pre_detection == F & symptomatic == F))$LIB13.C..hominis<40,1,0) %>% table()
+ifelse(sample_data(subset_samples(ps_rare_diff, pre_detection == F & symptomatic == F))$C..mel.y<40,1,0) %>% table()
+ifelse(sample_data(subset_samples(ps_rare_diff, pre_detection == F & symptomatic == F))$Pan_crypto<40,1,0) %>% table()
+
+```

### 2020-09-08 code/R_markdown/P3_machine_learning.Rmd
added species random forest and updated figure 1

@@
 
+## DATA NOT SHOWN: Does microbiome predict Crypto species at TOD? HERE
+
+```{r}
+
+compare <- c(10,12) # TOD +/- symptoms
+ps_for_class2 <- subset_samples(ps_rare_diff, Clinical.type %in% compare)
+x <- cbind(otu_table(ps_for_class2), as(sample_data(ps_for_class2),"data.frame") %>% select(Age.in.days))
+y_hominis <- as.factor(ifelse(sample_data(ps_for_class2)$LIB13.C..hominis <40,1,0))
+y_meleagridis <- as.factor(ifelse(sample_data(ps_for_class2)$C..mel.y <40,1,0))
+
+rf = randomForest(x,y_hominis,importance = TRUE, ntree = 50*dim(x)[[2]], replace=TRUE, nodesize=1, classwt=c(0.0001,100))
+rf
+predictions = as.data.frame(predict(rf,type="prob"))
+predictions$predict = names(predictions)[1:2][apply(predictions[,1:2], 1, which.max)]
+predictions$observed = y_hominis
+ROC_res = roc(ifelse(predictions$observed==0, 0, 1), as.numeric(predictions$`0`))
+auc(ROC_res)
+save(list = c("rf","x","y_hominis"), file = paste0(paste0("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/rf_hominis_" , format(Sys.time(), "%Y_%m_%d")),".RData"))
+
+rf = randomForest(x,y_meleagridis,importance = TRUE, ntree = 50*dim(x)[[2]], replace=TRUE, nodesize=1, classwt=c(0.0001,100))
+rf
+predictions = as.data.frame(predict(rf,type="prob"))
+predictions$predict = names(predictions)[1:2][apply(predictions[,1:2], 1, which.max)]
+predictions$observed = y_meleagridis
+ROC_res = roc(ifelse(predictions$observed==0, 0, 1), as.numeric(predictions$`0`))
+auc(ROC_res)
+save(list = c("rf","x","y_meleagridis"), file = paste0(paste0("~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/rf_meleagridis_" , format(Sys.time(), "%Y_%m_%d")),".RData"))
+
+rm(x,y,rf,ps_for_class2)
+
+```
+
 # Save data for other scripts

### 2020-09-08 code/R_markdown/P5_mirzapur_analysis_figures.Rmd
added species random forest and updated figure 1

@@
 # load packages
-library(ggplot2); library(phyloseq)
+library(ggplot2); library(phyloseq); library(dplyr)
 
@@
 ```
+
+Species
+```{r}
+
+load("/Users/maureencarey/local_documents/work/crypto_microbiome/data_cannot_be_shared/data.RData")
+crypto_cts = burden_all_path %>% filter(crypto == 1) %>%
+  mutate(date_of_collection = as.Date(gsub('.{8}$', '', as.character(DOC)), "%m/%d/%Y")) %>%
+  select(CHILDID, date_of_collection, CRYCTRT, type) %>% filter(CHILDID %in% burden_mirpur$SID)
+rm(list= ls()[! (ls() %in% c('ps_mirzapur_rare','crypto_cts'))])
+
+tempsampledata = sample_data(ps_mirzapur_rare)
+tempsampledata = merge(tempsampledata, crypto_cts %>% select(-type), by.x = c("Child.ID","Date.of.collection"), by.y = c("CHILDID","date_of_collection"), all.x = T)
+tempsampledata[tempsampledata$Clinical.type %in% c(1,2),"CRYCTRT"] = 45 # if predetection, Crypto DNA was not assessed or detected
+
+# get other Ct values (targeted qPCRs)
+all_ct = read.table("~/local_documents/work/crypto_microbiome/data_cannot_be_shared/Cq_values_all.txt",header=TRUE, sep='\t',comment.char = "",stringsAsFactors = F, na.strings = c("#N/A"))
+all_ct[is.na(all_ct)] = 45 # replace #NA and real NAs with 0
+all_ct[all_ct == 0] = 45 # replace #NA and real NAs with 0
+tempsampledata = tempsampledata %>% mutate(sample_id_mod = substr(sample.id,1,nchar(sample.id)-5))
+tempsampledata = merge(tempsampledata, all_ct, by.x = "sample_id_mod", by.y = "sample.id", all.x = T)
+rownames(tempsampledata) = tempsampledata$use_as_rows; tempsampledata$use_as_rows = NULL
+
+#12=diarrhea; 10=Clinic Visit; 1 =pre-clinic visit positive; 2=pre-diarrhea case positive
+
+#all
+ifelse(tempsampledata %>% filter(Clinical.type %in% c(10,12)) %>% select(LIB13.C..hominis)<40,1,0) %>% table()
+ifelse(tempsampledata %>% filter(Clinical.type %in% c(10,12)) %>% select(C..mel.y)<40,1,0) %>% table()
+ifelse(tempsampledata %>% filter(Clinical.type %in% c(10,12)) %>% select(Pan_crypto)<40,1,0) %>% table()
+
+# diarrheal
+ifelse(tempsampledata %>% filter(Clinical.type %in% c(12)) %>% select(LIB13.C..hominis)<40,1,0) %>% table()
+ifelse(tempsampledata %>% filter(Clinical.type %in% c(12)) %>% select(C..mel.y)<40,1,0) %>% table()
+ifelse(tempsampledata %>% filter(Clinical.type %in% c(12)) %>% select(Pan_crypto)<40,1,0) %>% table()
+
+# subclin
+ifelse(tempsampledata %>% filter(Clinical.type %in% c(10)) %>% select(LIB13.C..hominis)<40,1,0) %>% table()
+ifelse(tempsampledata %>% filter(Clinical.type %in% c(10)) %>% select(C..mel.y)<40,1,0) %>% table()
+ifelse(tempsampledata %>% filter(Clinical.type %in% c(10)) %>% select(Pan_crypto)<40,1,0) %>% table()
+
+```

### 2020-09-14 code/R_markdown/P1_analysis_prep_figures.Rmd
updates re focusing on top 15 imp var and adding crypto species to rf analysis

@@
 # load data
-load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P1output_plot_2020_09_02.RData")
+load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P1output_plot_2020_09_11.RData")
 
@@
 
-# Generate supplemental file 1B to get read counts
+# Generate supplemental file 1D to get read counts
 
@@
 
-# Plot positive controls
+# Plot positive controls (Supplemental Figure 1A-B)
 
@@
 
-# Figure 2
+# Supplemental Figure 4
 
@@
 # Table 1
+
 ```{r}
@@
 ```
+
+How many ASVs map to Prevotella? (For Fig 3 caption)
+```{r}
+
+ntaxa(subset_taxa(ps_rare_diff, Genus == "Prevotella"))
+ntaxa(subset_taxa(ps_rare_diff, Genus == "Bifidobacterium"))
+
+ntaxa(subset_taxa(ps_rare_diff, Genus == "Megasphaera"))
+ntaxa(subset_taxa(ps, Genus == "Megasphaera"))
+
+```

### 2020-09-14 code/R_markdown/P2_cohort.Rmd
updates re focusing on top 15 imp var and adding crypto species to rf analysis

@@
 # load data
-load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P1output_2020_09_02.RData")
+load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P1output_2020_09_11.RData")
 load("/Users/maureencarey/local_documents/work/crypto_microbiome/data_cannot_be_shared/data.RData")

### 2020-09-14 code/R_markdown/P2_cohort_figures.Rmd
updates re focusing on top 15 imp var and adding crypto species to rf analysis

@@
 library(ggplot2); library(rmarkdown); library(vegan); library(dotwhisker)
-library(dplyr); library(phyloseq)
+library(dplyr); library(phyloseq); library(tibble)
 
@@
 # load data
-load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P2output_plot_2020_09_02.RData")
+load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P2output_plot_2020_09_11.RData")
 
@@
 
+```{r}
+
+ps_rare_for_beta_mode_sample_data = ps_rare_for_beta
+qu = quantile(sample_data(ps_rare_for_beta_mode_sample_data)$Age.in.days, c(0.25, 0.5, 0.75))
+temp_sample_data = as(sample_data(ps_rare_for_beta_mode_sample_data),"data.frame") %>% 
+  rownames_to_column('sample') %>%
+  mutate(age_quartile = as.factor(ifelse(Age.in.days < qu[[1]][1], 1,
+                               ifelse(Age.in.days > qu[[3]][1], 4,
+                                      ifelse(Age.in.days > qu[[2]][1], 3,2)))),
+         hominis = as.factor(ifelse(LIB13.C..hominis <40,1,0)),
+         meleagridis = as.factor(ifelse(C..mel.y <40,1,0))) %>%
+    column_to_rownames('sample')
+sample_data(ps_rare_for_beta_mode_sample_data) = temp_sample_data
+p = plot_ordination(
+  physeq = ps_rare_for_beta_mode_sample_data,
+  ordination = ps_rare.ord,
+  color = "Clinical.type",
+  shape = "age_quartile"
+)   + 
+  scale_color_manual(values = c("#666666","#FF7F00","#1F78B4"),
+                     labels = c("12" = "diarrheal","10" = "subclinical","1" = "pre-detection") )+ # 12 is diarrhea, 1,2 are pre
+  #scale_shape_manual(values = c(5,16,4),
+   #                  labels = c("1" = "youngest quartile","2" = "middle","4" = "oldest quartile") )+
+  theme(plot.margin=unit(c(.5,.5,.5,.5),"cm"))  + 
+  labs(color = "symptomatic?", shape = "Age (quartitled)")
+p
+
+p = plot_ordination(
+  physeq = ps_rare_for_beta_mode_sample_data,
+  ordination = ps_rare.ord,
+  color = "Clinical.type",
+  shape = "meleagridis"
+)   + 
+  scale_color_manual(values = c("#666666","#FF7F00","#1F78B4"),
+                     labels = c("12" = "diarrheal","10" = "subclinical","1" = "pre-detection") )+ # 12 is diarrhea, 1,2 are pre
+  scale_shape_manual(values = c(5,16),
+                     labels = c("1" = "positive","0" = "negative") )+
+  theme(plot.margin=unit(c(.5,.5,.5,.5),"cm"))  + 
+  labs(color = "symptomatic?", shape = "C. meleagridis")
+p
+
+p = plot_ordination(
+  physeq = ps_rare_for_beta_mode_sample_data,
+  ordination = ps_rare.ord,
+  color = "Clinical.type",
+  shape = "hominis"
+)   + 
+  scale_color_manual(values = c("#666666","#FF7F00","#1F78B4"),
+                     labels = c("12" = "diarrheal","10" = "subclinical","1" = "pre-detection") )+ # 12 is diarrhea, 1,2 are pre
+  scale_shape_manual(values = c(5,16),
+                     labels = c("1" = "positive","0" = "negative") )+
+  theme(plot.margin=unit(c(.5,.5,.5,.5),"cm"))  + 
+  labs(color = "symptomatic?", shape = "C. hominis")
+p
+
+```
+
 Do all kids have at least one abx between predetction and time of detection???

### 2020-09-14 code/R_markdown/P3_machine_learning.Rmd
updates re focusing on top 15 imp var and adding crypto species to rf analysis

@@
 rm(list= ls()[! (ls() %in% c('all_cause_diar','SEED_INT','colors'))])
-load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P1output_2020_09_02.RData")
-load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P2output_2020_09_02.RData")
+load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P1output_2020_09_11.RData")
+load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P2output_2020_09_11.RData")
 
@@
 
-x <- cbind(otu_table(ps_for_class), as(sample_data(ps_for_class),"data.frame") %>% select(Age.in.days))
+x <- cbind(otu_table(ps_for_class), as(sample_data(ps_for_class),"data.frame") %>% select(Age.in.days, LIB13.C..hominis, C..mel.y)) %>% 
+  mutate(hominis = as.factor(ifelse(LIB13.C..hominis <40,1,0)), 
+         meleagridis = as.factor(ifelse(C..mel.y <40,1,0))) %>% select(-LIB13.C..hominis,-C..mel.y)
 y <- as.factor(sample_data(ps_for_class)$Clinical.type)
@@
 extracted_importance = extracted_importance[rownames(extracted_importance)[order(extracted_importance)],]
-top10_PD = tail(extracted_importance,10)
+top10_PD = tail(extracted_importance,15)
 
@@
 ps_for_class <- subset_samples(ps_rare_diff, Clinical.type %in% compare)
-x <- cbind(otu_table(ps_for_class), as(sample_data(ps_for_class),"data.frame") %>% select(Age.in.days))
+x <- cbind(otu_table(ps_for_class), as(sample_data(ps_for_class),"data.frame") %>% select(Age.in.days, LIB13.C..hominis, C..mel.y)) %>% 
+  mutate(hominis = as.factor(ifelse(LIB13.C..hominis <40,1,0)), 
+         meleagridis = as.factor(ifelse(C..mel.y <40,1,0))) %>% select(-LIB13.C..hominis,-C..mel.y)
 y <- as.factor(sample_data(ps_for_class)$Clinical.type)
@@
 
-rf = randomForest_res(x_for_TOD,y_for_TOD,10)
+rf = randomForest_res(x_for_TOD,y_for_TOD,15)
 top10_TOD = rf[[4]]
@@
 ps_for_class <- subset_samples(ps_rare_diff, Clinical.type %in% compare)
-x <- cbind(otu_table(ps_for_class), as(sample_data(ps_for_class),"data.frame") %>% select(Age.in.days))
+x <- cbind(otu_table(ps_for_class), as(sample_data(ps_for_class),"data.frame") %>% select(Age.in.days, LIB13.C..hominis, C..mel.y)) %>% 
+  mutate(hominis = as.factor(ifelse(LIB13.C..hominis <40,1,0)), 
+         meleagridis = as.factor(ifelse(C..mel.y <40,1,0))) %>% select(-LIB13.C..hominis,-C..mel.y)
 y <- as.numeric(as.character(sample_data(ps_for_class)$Number.of.crypto.positive.events))
@@
 sample_data(ps_for_class2)$episodes = all_cause_diar_temp$episodes[order(match(all_cause_diar_temp$sample_id_mod, sample_data(ps_for_class2)$sample_id_mod))]
-x <- cbind(otu_table(ps_for_class2), as(sample_data(ps_for_class2),"data.frame") %>% select(Age.in.days))
+x <- cbind(otu_table(ps_for_class2), as(sample_data(ps_for_class2),"data.frame") %>% select(Age.in.days, LIB13.C..hominis, C..mel.y)) %>% 
+  mutate(hominis = as.factor(ifelse(LIB13.C..hominis <40,1,0)), 
+         meleagridis = as.factor(ifelse(C..mel.y <40,1,0))) %>% select(-LIB13.C..hominis,-C..mel.y)
 y <- sample_data(ps_for_class2)$episodes
@@
 
-x <- cbind(cbind(time_of_detect_microbes,surv_microbes), as(sample_data(ps_for_class),"data.frame") %>% select(Age.in.days))
+x <- cbind(cbind(time_of_detect_microbes,surv_microbes), as(sample_data(ps_for_class),"data.frame") %>% select(Age.in.days, LIB13.C..hominis, C..mel.y)) %>% 
+  mutate(hominis = as.factor(ifelse(LIB13.C..hominis <40,1,0)), 
+         meleagridis = as.factor(ifelse(C..mel.y <40,1,0))) %>% select(-LIB13.C..hominis,-C..mel.y)
 y = as.factor(sample_data(subset_samples(ps_for_class, Clinical.type %in% compare))$Clinical.type)
@@
 
-x_withabx = cbind(cbind(abx_history, otu_table(for_metadata)), as(sample_data(for_metadata),"data.frame") %>% select(Age.in.days))
+x_withabx = cbind(cbind(abx_history, otu_table(for_metadata)), as(sample_data(for_metadata),"data.frame") %>% select(Age.in.days, LIB13.C..hominis, C..mel.y)) %>% 
+  mutate(hominis = as.factor(ifelse(LIB13.C..hominis <40,1,0)), 
+         meleagridis = as.factor(ifelse(C..mel.y <40,1,0))) %>% select(-LIB13.C..hominis,-C..mel.y)
 y = as.factor(sample_data(for_metadata)$Clinical.type)
 
-rf = randomForest_res(x_withabx,y,25)
+rf = randomForest_res(x_withabx,y,15)
 rf[[1]]
@@
 plot_tax$var = reorder(plot_tax$var, -plot_tax$value)
-plot_tax$var <- relevel(plot_tax$var, "ASV19 (Megasphaera)")
+plot_tax$var <- relevel(plot_tax$var, levels(plot_tax$var)[grep("Megasphaera", levels(plot_tax$var))])
 plot_tax_4D = plot_tax
@@
 #ASV5 (Parabacteroides) for TOD
-TOD_twoASVs = tax2 %>% filter(var == 'ASV10 (Megasphaera)' | var == 'ASV5 (Parabacteroides)')
+TOD_twoASVs = tax2 %>% filter(var ==  tax2$var[grep("Megasphaera", tax2$var)] | 
+                                var == tax2$var[grep("Parabacteroides", tax2$var)])
 
@@
 #ASV5 (Parabacteroides) for TOD
-PD_twoASVs = tax2 %>% filter(var == 'ASV10 (Megasphaera)' | var == 'ASV3 (Collinsella)')
+PD_twoASVs = tax2 %>% filter(var == tax2$var[grep("Megasphaera", tax2$var)] | 
+                                var == tax2$var[grep("Collinsella", tax2$var)])
 tax2_PD = tax2
@@
 
-twoASVs = rbind(TOD_twoASVs %>% filter(var == "ASV10 (Megasphaera)") %>% mutate(time = 'time-of-detection'), 
-                PD_twoASVs %>% filter(var == "ASV10 (Megasphaera)") %>% mutate(time = 'pre-detection'))
+twoASVs = rbind(TOD_twoASVs %>% filter(var == TOD_twoASVs$var[grep("Megasphaera", TOD_twoASVs$var)]) %>% 
+                  mutate(time = 'time-of-detection'), 
+                PD_twoASVs %>% filter(var == PD_twoASVs$var[grep("Megasphaera", PD_twoASVs$var)]) %>% 
+                  mutate(time = 'pre-detection'))
 
@@
 
-twoASVs_PD_wide = tidyr::spread(PD_twoASVs %>% select(value_use,symptomatic_use,sample, Genus),
-                             key = Genus, value = value_use) %>% 
+twoASVs_PD_wide = tidyr::spread(PD_twoASVs %>% select(value_use,symptomatic_use,var, sample, Genus),
+                             key = var, value = value_use) %>% 
   mutate(`symptomatic?` = as.factor(symptomatic_use))
@@
 plt_df1 = as.data.frame(top10_TOD) %>%rownames_to_column('variable') %>% dplyr::rename(value = top10_TOD) 
-plt_df2 = as.data.frame(top20_TOD_with_abx) %>%rownames_to_column('variable') %>% dplyr::rename(value = top20_TOD_with_abx) %>% top_n(10,value)
+plt_df2 = as.data.frame(top20_TOD_with_abx) %>%rownames_to_column('variable') %>% dplyr::rename(value = top20_TOD_with_abx)
 plt_df = rbind(plt_df1 %>% mutate(model = 'without abx'),plt_df2 %>% mutate(model = 'with abx'))
@@
 
-## DATA NOT SHOWN: Does microbiome predict Crypto species at TOD? HERE
+## DATA NOT SHOWN: Does microbiome predict Crypto species at TOD?
 

### 2020-09-14 code/R_markdown/P3_machine_learning_figures.Rmd
updates re focusing on top 15 imp var and adding crypto species to rf analysis

@@
 # load data
-load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P3output_plot_2020_09_02.RData")
+#load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P3output_plot_2020_09_02.RData")
+#load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P3output_plot_2020_09_10.RData")
+#load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/cmel_P3output_plot_2020_09_11.RData")
+#load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/chom_P3output_plot_2020_09_11.RData")
+load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P3output_plot_2020_09_11.RData")
+#load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/WOC_P3output_plot_2020_09_11.RData")
 
@@
         axis.text.x = element_text(size = 7),
-        legend.title=element_text(size=10), 
+        legend.title=element_blank(), 
         legend.text=element_text(size=8))  + scale_x_discrete(limits = rev(levels(plot_tax_4D$var)))
@@
   geom_point(alpha = 0.5, position = position_jitterdodge(jitter.width = 0.2, jitter.height = 0, dodge.width = 0.75, seed = SEED_INT)) + 
-  facet_wrap(~time) +
+  facet_wrap(~time, scales = "free_x") +
   scale_color_manual(values= colors[1:2]) +
@@
        width = 3.2,height = 4,units = "in",dpi = 300)
-
-ggplot(two_WIDE, aes(x = `time-of-detection`, y = `pre-detection`, color = `symptomatic?`)) +
-  geom_jitter(alpha = 0.5, position = position_jitter(width = 0.01, height = 0.01, seed = SEED_INT)) +
-  scale_color_manual(values = colors[1:2]) + scale_x_sqrt() + scale_y_sqrt()+
-  theme(panel.grid.minor = element_blank(),
-        legend.position="bottom")
-
-ggplot(two_WIDE, aes(x = `time-of-detection`, y = `pre-detection`, color = `symptomatic?`)) +
-  geom_point(alpha = 0.5) +
-  scale_color_manual(values = colors[1:2]) + scale_x_sqrt() + scale_y_sqrt()+
-  theme(panel.grid.minor = element_blank(),
-        legend.position="bottom")
 
@@
 
-load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/rf_TOD_abx_2020_09_02.RData")
+load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/rf_TOD_abx_2020_09_11.RData")
+#load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/WOC_rf_TOD_abx_2020_09_11.RData")
+
 AUC_TOD_abx = rf[[3]]
@@
   ggplot() +geom_bar(aes(x=reorder(var,value), y = value, fill = model),stat = "identity", width=0.4, position = position_dodge(width=0.5)) +
-  scale_fill_manual(values = c("black","grey")) +
+  scale_fill_manual(values = c("grey","blue")) +
   xlab(NULL) + ylab('Mean decrease in node impurity') + #facet_wrap(~tod_or_pre, scales = "free_x") +

### 2020-09-14 code/R_markdown/P4_univariate_stats.Rmd
updates re focusing on top 15 imp var and adding crypto species to rf analysis

@@
 # load data
-load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P1output_2020_09_02.RData")
-load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P3output_p1_2020_09_02.RData")
+load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P1output_2020_09_11.RData")
+load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P3output_p1_2020_09_11.RData")
 

### 2020-09-14 code/R_markdown/P4_univariate_stats_figures.Rmd
updates re focusing on top 15 imp var and adding crypto species to rf analysis

@@
 # load data
-load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P4output_plot_2020_09_02.RData")
+load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P4output_plot_2020_09_11.RData")
 
@@
 ```
+
+Megasphaera fold change
+```{r}
+
+tod_tax2 %>% filter(Genus == 'Megasphaera', size2 == 1) %>% select(log2FoldChange,padj)
+
+```

### 2020-09-14 code/R_markdown/P5_mirzapur_analysis.Rmd
updates re focusing on top 15 imp var and adding crypto species to rf analysis

@@
 rm(list= ls()[! (ls() %in% c('left_study_ids'))])
-load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P3output_p2_2020_09_02.RData")
-load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P3output_p3_2020_09_02.RData")
+load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P3output_p2_2020_09_11.RData")
+load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P3output_p3_2020_09_11.RData")
 ps_mirzapur = readRDS("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/ps_2020_08_12.rds")
-load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P3output_p1_2020_09_02.RData") # load save_df1 and save_df2
+load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P3output_p1_2020_09_11.RData") # load save_df1 and save_df2
 
@@
 
-# number of Megaspaheara ASVs
-temp = as.data.frame(tax_table(ps_mirzapur_rare))
-dim(temp[temp$Genus == 'Megasphaera',])
-# temp2 = as.data.frame(tax_table(ps_rare))
-# dim(temp2[temp2$Genus == 'Megasphaera',])
+ntaxa(subset_taxa(ps_mirzapur_rare, Genus == "Megasphaera"))
 
@@
                              paste0(paste0(paste0(paste0(as.character('ASV'),id),' ('),Genus),')'))))
-twoASVs_TOD = tax2 %>% filter(var == 'ASV10 (Megasphaera)')
+twoASVs_TOD = tax2 %>% filter(var == tax2$var[grep("Megasphaera", tax2$var)])
 
@@
                              paste0(paste0(paste0(paste0(as.character('ASV'),id),' ('),Genus),')'))))
-PD_twoASVs = tax2 %>% filter(var == 'ASV10 (Megasphaera)')
+PD_twoASVs = tax2 %>% filter(var == tax2$var[grep("Megasphaera", tax2$var)])
 
@@
 
+#UPDATE HERE: count_df_PD_mirpur column names have switched from genus to ASV
+count_df_PD_mirzapur
 df_mirz = melt(rbind(count_df_PD_mirzapur, count_df_TOD_mirzapur))  %>% mutate(Site = 'Mirzapur')
-df_mir = melt(rbind(count_df_PD_mirpur %>% select(-symptomatic_use,-Collinsella,-`symptomatic?`) %>% 
+df_mir = melt(rbind(count_df_PD_mirpur %>% 
+                      rename(Megasphaera= `ASV15 (Megasphaera)`) %>%
+                      select(sample, Megasphaera) %>% 
                       mutate(time = 'pre-detection'), 

### 2020-09-14 code/R_markdown/P5_mirzapur_analysis_figures.Rmd
updates re focusing on top 15 imp var and adding crypto species to rf analysis

@@
 # load data
-load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P5output_plot_2020_09_02.RData")
+load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P5output_plot_2020_09_11.RData")
 
@@
 
+ntaxa(subset_taxa(ps_mirzapur_rare, Genus == "Megasphaera"))
+
 ```

### 2020-09-30 code/R_markdown/P2_cohort.Rmd
revision updates

@@
   by = "CHILDID", all = T)
+write.csv(df_plot$CHILDID,"/Users/maureencarey/local_documents/work/crypto_microbiome/manuscript_drafts/info_for_dbGaP/abx_samples.csv", row.names=FALSE)
 df_plot$n_diarr[is.na(df_plot$n_diarr)] = 0

### 2021-02-22 code/R_markdown/response_to_reviewers_p1.Rmd
response to reviewers

---
title: "Response to Reviewers - Part 1"
author: "Maureen A. Carey"
date: "1/14/2021"
output: html_document
---

```{r setup, include=FALSE}
# load products from P1_analysis_prep and ASV

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/")

# load packages
library(phyloseq); library(reshape2); library(tibble)
library(rmarkdown); library(dplyr); library(tidyr)
library(ggplot2)

# set constants
SEED_INT = 10; set.seed(SEED_INT); options(scipen=10000)
colors = c("#1F78B4","#FF7F00","#666666", 'light gray')
names(colors) = c('diarrheal','subclinical','no crypto','both')
# blue = "#1F78B4" orange = "#FF7F00" gray = "#666666"

# load data
load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P1output_plot_2020_09_11.RData")
load("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/results/P3output_plot_2020_09_11.RData")

colors_temp = c(colors[c(1:2,4)],colors[4]); names(colors_temp)[3] = ' prior to\ndiarrheal'
names(colors_temp)[4] = ' prior to\nsubclinical'
colors_temp2 = as.data.frame(colors_temp)

```

# Reviewer 1

Carey et al. uses 16s rRNA sequencing to identify gut microbes associated with the development of symptomatic vs asymptomatic cryptosporidiosis in infants in Bangladesh. Using univariate analysis and RandomForest machine learning of pre-infection and time-of-infection samples, Megasphaera was found to be negatively correlated with symptomatic disease.
The paper is clearly written, excellent figures, and the study is about an interesting and important issue. The sample set used is unique & valuable, since there are "pre-disease" samples, which are rare in the microbiome world. Regarding the methods, there is excellent accounting for antibiotics in this manuscript with clear info on use and this includes an analysis of how antibiotic use was not found to impact the classifier performance in the model. Although the sequencing methods were rigorous and appropriate (use of DADA2, RDP ref database and inclusion of mock communities as controls), unfortunately a lot data (average of 17k reads per sample, when only 10k were used) was thrown out in the normalization process-I was left wondering if other normalization methods (or no normalization method) would have resulted in the same conclusions.

I want to also mention that I know there is a significant literature on megasphaera (specifically eldensii) as a probiotic used in the veterinary world. I am not a veterinary microbiome expert, but perhaps looking into this data could add some interesting/relevant background to this manuscript, especially regarding potential hypotheses for mechanisms of this relationship.
Finally, 16S sequencing was used. Taxonomic identifications only (without gene-level information) have, over time, become harder to interpret due to a high amount of genetic diversity (ie functional capacity) in closely related organisms (the Prevotella genus is a great example of this). Due to this, and because 16S defines only to the genus level and each genera can span multiple ASVs, 16S analyses are by nature significantly limited in the conclusions that can be drawn, even in assessing correlations. That said, the conclusions drawn in this manuscript are appropriately modest, which was great to see (too many 16S papers overstate their findings).

## Major Concern #1

* Why were Euclidean distances used for principal coordinate analysis, and were other methods used also? There are many kinds of analytic methods for comparing populations (ie Unifrac), and Euclidean is not among the most common. For example, Euclidean distances are insensitive to rare taxonomic groups.

*Authors' response: *

PCoA with alternative methods
run rivanna script

## Major Concern #2

"Authors mention one Megasphaera ASV representing at least 1% of reads of the entire study. The prevalence of this ASV within each group is not shown. From Fig 4E, it seems the high abundance of Megasphaera is driven by 17 out of the 60 subclinical infections samples while there are 4-5 of 31 diarrheal samples with presence of Megasphaera. A heatmap of top ASVs would be beneficial to show the relevance of these ASVs in the overall cohort."

*Authors' response: *

Megasphaera prevalence
```{r}

twoASVs_seq = twoASVs %>% select(Row.names,Genus) %>% unique() 
Mega = twoASVs_seq$Row.names # SPECIFIC ASV
just_MegaASV = subset_taxa(ps_rare, rownames(tax_table(ps_rare)) %in% c(Mega))
asv_percent = cbind(as(sample_data(ps_rare),"data.frame") %>% select(pre_detection,symptomatic),
                     sample_sums(just_MegaASV)/sample_sums(ps_rare))
asv_percent_withIDs = cbind(as(sample_data(ps_rare),"data.frame") %>% select(pre_detection,symptomatic,Child.ID),
                     sample_sums(just_MegaASV)/sample_sums(ps_rare))
colnames(asv_percent)[3] = 'percent'
colnames(asv_percent_withIDs)[4] = 'percent'
asv_percent_withIDs = asv_percent_withIDs %>% group_by(Child.ID, symptomatic) %>% summarize(total_percent = sum(percent))%>% mutate(prevalence = ifelse(total_percent > 0, 1, 0))

asv_percent = asv_percent %>% 
  mutate(sample_group = ifelse(pre_detection,
                               ifelse(symptomatic, ' prior to\ndiarrheal',' prior to\nsubclinical'),
                               ifelse(symptomatic, 'diarrheal','subclinical')),
         prevalence = ifelse(percent > 0, 1, 0))

my_comparisons = list(c('diarrheal','subclinical'),
                      c('diarrheal',' prior to\ndiarrheal'),
                      c(' prior to\nsubclinical','subclinical'),
                      c(' prior to\nsubclinical',' prior to\ndiarrheal'))
ggplot(asv_percent, aes(x = sample_group, y = percent, color = sample_group)) + 
  scale_color_manual(values = colors_temp) +
  geom_boxplot(outlier.colour = NA) +
  geom_point(alpha = 0.5, position = position_jitter(width = 0.3,height = 0)) + 
  scale_y_sqrt() +
  ggpubr::stat_compare_means(comparisons = my_comparisons, size = 2, method = 't.test') +
  guides(color = F) + xlab(NULL) + ylab('percent of total reads') +ggtitle('Important Megasphaera ASV')
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/figures_for_reviewers/mega_prevalence_p1.png",
       width = 3.2,height = 3,units = "in",dpi = 300)

# All other Mega ASVs
excluding_MegaASV = subset_taxa(ps_rare, Genus=="Megasphaera" & !(rownames(tax_table(ps_rare)) %in% c(Mega)))
genus_percent = cbind(as(sample_data(ps_rare),"data.frame") %>% select(pre_detection,symptomatic),
                    sample_sums(excluding_MegaASV)/sample_sums(ps_rare))
colnames(genus_percent)[3] = 'percent'

genus_percent = genus_percent %>% 
  mutate(sample_group = ifelse(pre_detection,
                               ifelse(symptomatic, ' prior to\ndiarrheal',' prior to\nsubclinical'),
                               ifelse(symptomatic, 'diarrheal','subclinical')),
         prevalence = ifelse(percent > 0, 1, 0))

ggplot(genus_percent, aes(x = sample_group, y = percent, color = sample_group)) + 
  scale_color_manual(values = colors_temp) +
  geom_boxplot(outlier.colour = NA) +
  geom_point(alpha = 0.5, position = position_jitter(width = 0.3,height = 0)) + 
  scale_y_sqrt() +
  ggpubr::stat_compare_means(comparisons = my_comparisons, size = 2, method = 't.test') +
  guides(color = F) + xlab(NULL) + ylab('percent of total reads') +ggtitle('All other Megasphaera ASVs')
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/figures_for_reviewers/mega_prevalence_p2.png",
       width = 3.2,height = 3,units = "in",dpi = 300)

# prevalence plot

subclin = asv_percent %>% filter(sample_group == 'subclinical') %>% select(prevalence) %>% table()
diar = asv_percent %>% filter(sample_group == 'diarrheal') %>% select(prevalence) %>% table()
p_subclin = asv_percent %>% filter(sample_group == ' prior to\nsubclinical') %>% select(prevalence) %>% table()
p_diar = asv_percent %>% filter(sample_group == ' prior to\ndiarrheal') %>% select(prevalence) %>% table() 
groups = c('subclinical','diarrheal',' prior to\nsubclinical',' prior to\ndiarrheal')
asv_prev = as.data.frame(cbind(groups,
      rbind(subclin, diar,p_subclin, p_diar)))
asv_prev$`0` = as.numeric(as.character(asv_prev$`0`))
asv_prev$`1` = as.numeric(as.character(asv_prev$`1`))
asv_prev = asv_prev %>% mutate(prevalence_frac = (`1`/(`0`+`1`)),
                                   prevalence = 100*prevalence_frac)
asv_prev_across_study = 100*sum(asv_prev$`1`)/(sum(asv_prev$`0`)+sum(asv_prev$`1`))

kids_with_ASV = with(asv_percent_withIDs, table(prevalence, symptomatic))
kids_with_ASV

subclin = genus_percent %>% filter(sample_group == 'subclinical') %>% select(prevalence) %>% table()
diar = genus_percent %>% filter(sample_group == 'diarrheal') %>% select(prevalence) %>% table()
p_subclin = genus_percent %>% filter(sample_group == ' prior to\nsubclinical') %>% select(prevalence) %>% table()
p_diar = genus_percent %>% filter(sample_group == ' prior to\ndiarrheal') %>% select(prevalence) %>% table() 
groups = c('subclinical','diarrheal',' prior to\nsubclinical',' prior to\ndiarrheal')
genus_prev = as.data.frame(cbind(groups,
      rbind(subclin, diar,p_subclin, p_diar)))
genus_prev$`0` = as.numeric(as.character(genus_prev$`0`))
genus_prev$`1` = as.numeric(as.character(genus_prev$`1`))
genus_prev = genus_prev %>% mutate(prevalence_frac = (`1`/(`0`+`1`)),
                                   prevalence = 100*prevalence_frac)
genus_prev_across_study = 100*sum(genus_prev$`1`)/(sum(genus_prev$`0`)+sum(genus_prev$`1`))

ggplot(asv_prev, aes(x = groups, y = prevalence, fill = groups)) + 
  scale_fill_manual(values = colors_temp) +
  geom_bar(stat = "identity") + 
  guides(fill = F) + xlab(NULL) + ylab('detected in X% of samples') +ggtitle('Important Megasphaera ASV')
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/figures_for_reviewers/mega_prevalence_p3.png",
       width = 3.2,height = 3,units = "in",dpi = 300)

ggplot(genus_prev, aes(x = groups, y = prevalence, fill = groups)) + 
  scale_fill_manual(values = colors_temp) +
  geom_bar(stat = "identity") + 
  guides(fill = F) + xlab(NULL) + ylab('detected in X% of samples') +ggtitle('All other Megasphaera ASVs')
ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/figures_for_reviewers/mega_prevalence_p4.png",
       width = 3.2,height = 3,units = "in",dpi = 300)

fyi_important_seq = twoASVs_seq$Row.names
fyi_unimportant_seq = colnames(otu_table(excluding_MegaASV))
```

Megasphaera ASV prevalence across study: `r asv_prev_across_study`

Megasphaera prevalence (all ASVs) across study: `r genus_prev_across_study`

Heatmap of top ASVs
```{r}

sample_data(ps_rare)$group = ifelse(sample_data(ps_rare)$pre_detection,ifelse(sample_data(ps_rare)$symptomatic,' prior to diarrheal',' prior to subclinical'),ifelse(sample_data(ps_rare)$symptomatic,'diarrheal','subclinical'))

# # all
# (ps_fam_rare <- tax_glom(ps_rare, "Family"))
# plot_heatmap(ps_fam_rare, sample.label = "group", taxa.label = "Family", 
#              sample.order="group")

# ## most common (top 100)
# ps_rare_top <- prune_taxa(names(sort(taxa_sums(ps_rare),TRUE)[1:100]), ps_rare)
# (ps_fam_rare <- tax_glom(ps_rare_top, "Family"))
# plot_heatmap(ps_fam_rare, method = "NMDS", distance = "bray", sample.label = "group", taxa.label = "Family", 
#              sample.order="group")

## most common (top 10)
ps_rare_top <- prune_taxa(names(sort(taxa_sums(ps_rare),TRUE)[1:10]), ps_rare)

groups = as.data.frame(sample_data(ps_rare_top)$group, row.names = NULL); colnames(groups) = 'group'
groups = merge(groups, colors_temp, all.y = T, by.y = 'row.names',by.x = 'group')
plot_heatmap(ps_rare_top, sample.label = "group", taxa.label = "Genus", sample.order="group")+
  theme(axis.ticks.x = element_line(color = as.character(groups$y)), axis.text.x = element_text(color = as.character(groups$y), size = 2)) + 
  guides( fill = F)

ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/figures_for_reviewers/heatmap.png",
       width = 8,height = 3,units = "in",dpi = 300)

```

## Major Concern #3

* Does analysis at a higher taxonomic level yield different results?

*Authors' response: This is a more involved response so see document: response_to_reviewers_p2_higher_tax.Rmd*

*In summary, looking at higher taxonomic level yields different results. For example, Megasphaera (as a genus rather than the specific ASV) is not important in the classifier because many of the ASVs in the genus are not different across groups (see prevalence plots above).*

*When we look at higher taxonomic levels, Bacteroides, Oisenella, Sutterella, Bifidobacterium, Streptococcus, Dialister, and Veillonella are important at both timepoints. None of these genera were important in the ASV-level predetection classifier and only Sutterella, Bifidobacterium, and Veillonella were important in the ASV-level time-of-detection classifier.*

## Major Concern #4

* There is no limitations section. I suggest small to moderate sample size, unknown generalizability to global populations, large amount of data thrown out for normalization, and the fact that pre -infection and post-infection samples sequenced in different batches as minimum discussion points. (While the last point is fully disclosed earlier in the paper, this should be further discussed including how this may have biased results).

## Major Concern #5

* Is there any in vitro data probing the mechanism of Megasphaera association with symptomatic vs. asymptomatic Cryptosporidium? This would strengthen the manuscript, but I understand it may be out of the scope of the study.

## Minor Concerns

Minor:
* The concluding sentence appears to be about a separate topic (treatment of disease). The manuscript is about ASVs found prior to infection (predictive of disease). The link between an ASV that may prevent colonization/initiation of disease is not related to the ASVs potential as a treatment once disease has begun - unless you are suggesting this ASV as a potential prophylactic among an exposed population. I think a different concluding sentence may serve the study better.  
* Include # of people studied, and # of people where Megasphaera was & was not found in samples, in the abstract.  
* Words/phrases like "classifier" and "node purity" need to be explained (or, even better, a different terminology used); there may be unfamiliarity with machine learning methods. 
* There is the axis label "symptomatic?" (with a question mark) in several suppl figs may be clearer with the label "Infection" (since the sublabels are "subclinical" and "diarrheal"). 
* Reference corrections: cited Ref 57 for Megasphaera colonization in the small intestines but that paper does not have Megasphaera in the text. Ref 58 for Megasphaera colonization in the large intestines but this paper found Megasphaera in vaginal swabs and there are no colonic samples.

# Reviewer 2

This study aimed at characterizing the role the intestinal microbiota plays in cryptosporidial infection and if there are differences between symptomatic and asymptomatic infections in children. This is an important area of research since infections of cryptosporidium are important causes of diarrheal illness in children, especially in developing countries and little is known about how the microbiota its a role in infection. The authors looked for associations with different groups and utilized different statistical analysis to try find patterns of microbiota with clinical disease. However, for the most part they were not able to identify distinct patterns. One of the challenges with this type of study was the diversity of the initial samples. This is understandably difficult to control in a setting where the population is exposed to a breadth of pathogens and antibiotic treatment in patient population appears widespread. Using initial samples with greater exclusions or more narrowly focused criteria might provide a less diverse initial samples, though arguably less relevant in the setting. One interesting result, is that cryptosporidium infections were associated with decreases in Megashaera group and if proven to be consistent in followed up studies could be expanded or be used as future biomarkers. Overall, the manuscript is well written and data appropriately displayed and analyzed.

### 2021-02-22 code/R_markdown/P1_analysis_prep_figures.Rmd
revision updates
figure mods for response to reviewers

@@
 
+write.csv(sample_data(ps_rare_diff)[,c(1,2)] %>% mutate(), "/Users/maureencarey/local_documents/work/crypto_microbiome/manuscript_drafts/info_for_dbGaP/mirpur_samples_passedQC.csv", row.names = F)
 ```
@@
         plot.margin=unit(c(.5,.5,.5,.5),"cm")) + 
-  labs(color = "symptomatic?", shape = "Number of infections")
+  labs(color = "infection", shape = "Number of infections")
 ggsave("/Users/maureencarey/local_documents/work/crypto_microbiome/cryptosporidium_microbiome/figures/SFig4_pcoa_change_mirpur.png",  width = 5.4,height = 3,units = "in",dpi = 600)

### 2021-02-22 code/R_markdown/P2_cohort_figures.Rmd
figure mods for response to reviewers

@@
   theme(plot.margin=unit(c(.5,.5,.5,.5),"cm"))  + 
-  labs(color = "symptomatic?", shape = "Number of infections")
+  labs(color = "infection", shape = "Number of infections")
 p
@@
   scale_color_manual(values = c(colors['diarrheal'],colors['subclinical'])) +
-  ylab('number of antibiotic doses') + xlab('symptomatic?') + guides(color = F) +
+  ylab('number of antibiotic doses') + xlab('infection') + guides(color = F) +
   ggpubr::stat_compare_means(comparisons = my_comparisons, size = 3, method = 't.test')

### 2021-02-22 code/R_markdown/P3_machine_learning_figures.Rmd
figure mods for response to reviewers

@@
         legend.position = "bottom",
-        legend.title = element_text(size = 9)) + guides(color=guide_legend(title="symptomatic?")) +
+        legend.title = element_text(size = 9)) + guides(color=guide_legend(title="infection")) +
   geom_hline(yintercept = 0.2, linetype = 'dotted', alpha = .75)

### 2021-02-22 code/on_rivanna/R_scripts/unifrac_phylo.R
response to reviewers - unifrac

@@ -1,0 +1,56 @@
+#install.packages("BiocManager", repos='http://cran.us.r-project.org') # WORKS
+#BiocManager::install("dada2", dependencies = TRUE)
+
+library("dada2")
+library("phyloseq")
+library("DECIPHER")
+library("phangorn")
+library("ggplot2")
+sink('/project/tumi/carey/crypto_micro/phylo_out.txt')
+packageVersion('dada2')
+
+load("/home/mac9jc/R_obj.Rdata")
+
+ps_rare = ps_rare_for_beta
+sequences = dada2::getSequences(otu_table(ps_rare))
+names(sequences) = sequences
+alignment <- AlignSeqs(DNAStringSet(sequences), anchor=NA)
+phang.align <- phyDat(as(alignment, "matrix"), type="DNA")
+print('aligned')
+dm <- dist.ml(phang.align)
+print('dist ml')
+treeNJ <- NJ(dm)
+print('NJ')
+fit = pml(treeNJ, data=phang.align)
+print('pml')
+fitGTR <- update(fit, k=4, inv=0.2)
+print('update')
+fitGTR <- optim.pml(fitGTR, model="GTR", optInv=TRUE, optGamma=TRUE,
+                    rearrangement = "stochastic", control = pml.control(trace = 0))
+print('optim.pml')
+print(fitGTR)
+detach("package:DECIPHER", unload=TRUE)
+detach("package:phangorn", unload=TRUE)
+phy_tree(ps_rare) = phy_tree(fitGTR$tree) 
+save(list = c("ps_rare"),file = paste0(paste0("/home/mac9jc/new_R_obj_" , format(Sys.time(), "%Y_%m_%d")),".RData"))
+
+# calc distance and do ordination
+dist_obj = phyloseq::distance(ps_rare, method = "wunifrac") #uunifrac for unweighted
+print('dist ibj')
+ord_obj = ordinate(ps_rare, method = "PCoA", distance = dist_obj) # default method is DCA
+print('ord obj')
+
+#plot_ordination(ps_rare, ord_obj, type = "samples")
+#
+p = plot_ordination(
+  physeq = ps_rare,
+  ordination = ord_obj,
+  color = "Clinical.type",
+  shape = "Number.of.crypto.positive.events"
+)   + 
+  scale_color_manual(values = c("#666666","#FF7F00","#1F78B4"),
[…]
